### 2014-03-29 HW_non-modal.tex
First commit

%\documentclass[showpacs,preprintnumbers,amsmath,amssymb,superscriptaddress,aip]{revtex4-1}
\documentclass[letter,scriptaddress,twocolumn, prl,showkeys]{revtex4}
\usepackage{graphicx}
%\usepackage{amssymb}
%\setlength{\parindent}{0in}

% General Latex  --------------------------------------------------
\def\beq{\begin{equation}}
\def\eeq{\end{equation}}
\def\beqar{\begin{eqnarray}}
\def\eeqar{\end{eqnarray}}
\def\nn{\nonumber}
\def\ol{\overline}
\def\para{\parallel}

% Operators  ------------------------------------------------------
\newcommand{\diff}[2]{\frac{d#1}{d#2}}
\newcommand{\diffs}[2]{\frac{d^2#1}{d#2^2}}
\newcommand{\pdiff}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\pdiffs}[2]{\frac{\partial^2#1}{\partial#2^2}}
\newcommand{\pdiffxy}[3]{\frac{\partial^2#1}{\partial#2 \partial#3}}
\newcommand{\pdt}{\partial_t}
\newcommand{\pdr}{\partial_r}
\newcommand{\pdth}{\partial_\theta}
\newcommand{\pdrr}{\partial^2_r}

\newcommand{\enum}[2]{{#1}\times10^{#2}} % 4.2x10^{3} = \enum{4.2}{3}

\newcommand{\vect}[1]{{\bf #1}}
%\newcommand{\vect}{\overrightarrow}
%\newcommand{\vect}{\vec}
\def\div{\nabla\cdot}
\def\grad{\nabla}
\def\curl{\nabla\times}
\newcommand{\gradpar}{\grad_\parallel}
\newcommand{\gradperp}{\grad_\perp}
\newcommand{\gradr}{\grad_r}
\newcommand{\defeq}{\ensuremath{\stackrel{\text{\tiny def}}{=}}}

\newcommand{\savg}[1]{\left<{#1}\right>}
\newcommand{\vavg}[1]{\left<{#1}\right>_V}
\newcommand{\thavg}[1]{\left<{#1}\right>_\theta}

% Variable names  -------------------------------------------------
\newcommand{\vpar} {v_\parallel}
\newcommand{\Apar} {A_\parallel}
\newcommand{\jpar} {j_\parallel}
\newcommand{\kpar} {k_\parallel}
\newcommand{\kperp} {k_\perp }
\newcommand{\vperp} {v_\perp }
\newcommand{\kthe}{k_\theta}

\newcommand{\Evec}{\ensuremath{\boldsymbol{{\rm E}}}}
\newcommand{\Bvec}{\ensuremath{\boldsymbol{{\rm B}}}}
\newcommand{\Jvec}{\ensuremath{\boldsymbol{{\rm J}}}}
\newcommand{\Fvec}{\ensuremath{\boldsymbol{{\rm F}}}}
\newcommand{\fvec}{\ensuremath{\boldsymbol{{\rm f}}}}
\newcommand{\vE}{\ensuremath{\boldsymbol{{\rm v}_{E}}}}
\newcommand{\bo}{\ensuremath{\boldsymbol{{\rm b}_0}}}
\newcommand{\bvec}{\ensuremath{\boldsymbol{{\rm b}}}}
\newcommand{\xvec}{\ensuremath{\boldsymbol{{\rm x}}}}
\newcommand{\yvec}{\ensuremath{\boldsymbol{{\rm y}}}}
\newcommand{\zvec}{\ensuremath{\boldsymbol{{\rm z}}}}
\newcommand{\vvec}{\ensuremath{\boldsymbol{{\rm v}}}}
\newcommand{\jvec}{\ensuremath{\boldsymbol{{\rm j}}}}

\newcommand{\bxgp}{\bvec\times\gradperp}

\newcommand{\vve}{\ensuremath{\boldsymbol{{\rm v}}_{e}}}
\newcommand{\vvi}{\ensuremath{\boldsymbol{{\rm v}}_{i}}}
\newcommand{\vpe}{v_{\parallel e}}
\newcommand{\vpi}{v_{\parallel i}}
\newcommand{\vvE}{\ensuremath{\boldsymbol{{\rm v}}_{E}}}
\newcommand{\vvD}{\ensuremath{\boldsymbol{{\rm v}}_{D}}}

\newcommand{\nuei}{\nu_{ei}}
\newcommand{\nuii}{\nu_{ii}}
\newcommand{\nue}{\nu_{e}}
\newcommand{\nuen}{\nu_{en}}
\newcommand{\nuin}{\nu_{in}}
\newcommand{\kpe}{\kappa_{\parallel e}}

\newcommand{\rs}{\rho_{s}}
\newcommand{\ri}{\rho_{i}}
\newcommand{\wci}{\Omega_{i}}
\newcommand{\wcix}{\Omega_{ix}}
\newcommand{\wce}{\Omega_{e}}
\newcommand{\tomega}{\tilde\omega}
\newcommand{\Isat}{I_{\rm sat}}
\newcommand{\fmie}{\frac{m_i}{m_e}}
\newcommand{\fmei}{\frac{m_e}{m_i}}

% Often used dimensions
\newcommand{\cm}{\rm cm}
\newcommand{\mm}{\rm mm}
\newcommand{\cmn}{{\rm cm}^{-3}}
\newcommand{\mn}{{\rm m}^{-3}}
\newcommand{\eV}{\rm eV}
\newcommand{\G}{\rm G}
\newcommand{\T}{\rm T}

\begin{document}

\title{}

\author{B. Friedman}
\email{friedman@physics.ucla.edu}

\author{T.A. Carter}

\affiliation{Department of Physics and Astronomy, University of California, Los Angeles, California 90095-1547, USA}

\begin{abstract}

\end{abstract}

\maketitle

\section{Introduction}

Normal mode analysis -- the calculation of eigenvalues and eigenvectors of a linearized dynamical system -- has been used to solve many problems over the years.
Despite its wide-ranging success, it has failed in important instances, particularly in predicting the onset of subcritical turbulence in hydrodynamic flows. 
The reason for this failure was explained in the early 1990's when Trefethen and others attributed the pitfalls of normal mode analysis to the non-normality of linear operators of
dynamical systems~\cite{trefethen1993,schmid2007}. A non-normal operator has 
eigenvectors that are not orthogonal to one another. One consequence of eigenvector nonorthogonality is that even when all eigenvectors decay exponentially under linear evolution, 
superpositions of eigenvectors can grow, albeit transiently.
In other words, certain fluctuations of the laminar state can access free energy from background gradients even though normal mode fluctuations cannot.
When combined with nonlinear effects, this allows for sustained subcritical turbulence.
Such behavior is obscured by traditional normal mode analysis, which only effectively describes the long time asymptotic behavior of fluctuations under  
action of the linear operator. Transient growth, which can dominate turbulent evolution, can be discovered only through non-modal calculations.

Non-modal analysis has been embraced by the hydrodynamics community over the past two decades in the attempt to explain and predict subcritical turbulence. But the plasma community
generally relies on normal mode analysis to inform turbulent predictions and observations, with a few exceptions~\cite{camargo1998,camporeale2010,schekochihin2012}. 
Furthermore, non-modal treatments have generally been explanatory rather than predictive and have centered around the transition to turbulence in subcritical systems rather 
than on properties of fully-developed turbulence.
This paper takes up the task of developing an approach to understand turbulent properties using only non-modal linear calculations with the goal of 
ultimately making quantitative predictions.  Our approach is to calculate an average growth rate of turbulent fluctuations due to linear processes, specifically due to transient growth.  We model the
turbulent steady state as a series of processes:  (1) the turbulence starts as a spatially random state, (2) linear transient growth deterministically amplifies the turbulent energy (or
decrease it in wavenumber ranges where linear damping dominates), (3) nonlinear transfer sets in at a specified timescale, terminating the
transient growth process and re-randomizing the turbulent state (at which point the cycle repeats).    Optimally, the timescale for the final step would be the
nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
arguments to use a characteristic \emph{linear} time. Since there is no obvious single linear time scale, we test several and compare the results to determine which works best.
The procedure ultimately produces a growth rate that can be used
to predict turbulent properties such as saturation levels and transport rates through mixing length arguments.

\section{Non-Normality of the Hasegawa-Wakatani Model}

Before employing the non-modal technique, we explore the effect of non-normality in the well-known and well-analyzed Hasegawa-Wakatani (HW) model~\cite{hasegawa1983}. 
We consider both the original 2D and the extended 3D~\cite{biskamp1995} versions of the model in a straight, unsheared magnetic field. 
The HW model consists of two equations for the density $n$ and the electrostatic potential $\phi$. The 3D version of the model is

\beqar
\label{n_eq}
\pdiff{n}{t} = - {\mathbf v_E} \cdot \grad n - \kappa \pdiff{\phi}{y} + \xi \gradpar^2 (n - \phi) - D \gradperp^4 n, \\
\label{phi_eq}
\pdiff{\gradperp^2 \phi}{t} = - {\mathbf v_E} \cdot \grad (\gradperp^2 \phi) + \xi \gradpar^2 (n - \phi) - D \gradperp^6 \phi
\eeqar
The normalizations are $\phi \to e \phi/T_e, n \to n/n_0, \grad \to \rho_s \grad , x,y,z \to x,y,z/\rho_s, t \to \omega_{ci} t $. Additionally,
$\mathbf{v_E} = \mathbf{b} \times \grad \phi$, $\xi = 4 \pi \omega_{ce}/\nu_{ei}$, 
and $\kappa = \rho_s/L_n$, where $L_n$ is the density scale length. The terms $D \gradperp^4 n$ and $D \gradperp^6 \phi$ are artificial hyper-diffusion and hyper-viscosity used to dissipate
energy on small scales. We set $D \to 10^{-5}$ except when otherwise indicated.
We associate the direction along $\mathbf{B}$ as the parallel or $z$ direction. The direction perpendicular $\mathbf{B}$ that
points opposite the density gradient is defined as the $x$ direction, and the $y$ direction is taken perpendicular to both of these. 
In the analysis and numerical simulations, which we perform with the BOUT++ code~\cite{dudson2009}, we use periodic boundary conditions in
the $y$ and $z$ directions with zero-value boundaries in the $x$ direction. For the 2D version of the model, we replace $-\xi \gradpar^2$ by 
$\alpha = k_z^2 \xi = 4 \pi k_z^2 \rho_s^2 \omega_{ce} /\nu_{ei}$, which is commonly known as the adiabaticity parameter~\cite{camargo1995,camargo1998}.

The degree of non-normality of the linear operator in the HW model is largely a function of the adiabaticity parameter~\cite{camargo1998}, or in the 3D case, of $\xi \gradpar^2$. 
Note that a matrix is normal if it commutes with its adjoint: $\mathbf{D} \mathbf{D}^\dagger = \mathbf{D}^\dagger \mathbf{D}$ for $\mathbf{D}$ normal. 
The HW model has a normal linear operator in the limit of $\alpha, \xi \gradpar^2 \to \infty$. 
To see this, we first transform the linear operator into a linear matrix by using a Fourier decomposition. Because we use a zero-value boundary condition in the $x$ direction, the
Fourier decomposition takes the form:

\beq
\label{fourier_form}
f_k = \int_0^{L_x,L_y,L_z} f(\mathbf{r}) \ {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z} d\mathbf{r}
\eeq
where $k_x = \pi l/L_x, k_y = 2 \pi m/L_y, {\rm and } \ k_z = 2 \pi n/L_z$ with $l, m, {\rm and } \ n$ integers.
The result for the 3D model is an equation for each mode $k = (k_x,k_y,k_z)$:

\beqar
\label{fourier_eqn}
\frac{\partial}{\partial t} \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right) = \mathbf{A}_k \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right) + \mathbf{N}_k \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right), \\ \nonumber \\
\mathbf{A}_k = \left( \begin{array}{cc} -\xi k_z^2 - D k_\perp^4 & -i \kappa k_y + \xi k_z^2 \\  \xi \frac{k_z^2}{k_\perp^2} & - \xi \frac{k_z^2}{k_\perp^2} - D k_\perp^4\end{array} \right)
\eeqar
with $\mathbf{A}_k$ being the linear coupling matrix and $\mathbf{N}_k$ the nonlinear matrix, which we don't write explicitly. 
Now in assessing the degree of normality of the HW model, we are not interested in the normality of $\mathbf{A}_k$ itself. 
Rather, we are interested in the energetically-weighted version of this matrix~\cite{camargo1998,schmid2007,camporeale2010}. Since the energy in a given Fourier mode is

\beq
\label{en_def}
E_k =  \frac{1}{2} \left( |n_k|^2 + k_\perp^2 |\phi_k|^2 \right).
\eeq
we rewrite Eq.~\ref{fourier_eqn} as

\beqar
\label{fourier_en_eqn}
\frac{\partial}{\partial t} \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right) = \mathbf{B}_k \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right) + \mathbf{N}_k \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right), \\ \nonumber \\
\mathbf{B}_k = \left( \begin{array}{cc} -\xi k_z^2 - D k_\perp^4 & -i \kappa \frac{k_y}{k_\perp} + \xi \frac{k_z^2}{k_\perp} \\  \xi \frac{k_z^2}{k_\perp} & - \xi \frac{k_z^2}{k_\perp^2} - D k_\perp^4\end{array} \right)
\eeqar
Since the square $L_2$-norm of the new state vector $\frac{1}{\sqrt{2}} \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right)$ gives the energy contained in the mode $k$,
$\mathbf{B}_k$ is the energetically-weighted matrix of interest. In the limit of large $\xi k_z^2$ as long as $k_\perp$ is also not too large:

\beq
\label{B_norm_limit}
\displaystyle\lim_{\xi k_z^2 \to \infty} \mathbf{B}_k = \left( \begin{array}{cc} -\xi k_z^2 & \xi \frac{k_z^2}{k_\perp} \\  \xi \frac{k_z^2}{k_\perp} & - \xi \frac{k_z^2}{k_\perp^2} \end{array} \right)
\eeq
where it is easy to see that $\mathbf{B}_k$ is self-adjoint and therefore normal. In the opposite limit:

\beq
\label{B_norm_limit0}
\displaystyle\lim_{\xi k_z^2 \to 0} \mathbf{B}_k = \left( \begin{array}{cc} - D k_\perp^4 & -i \kappa \frac{k_y}{k_\perp} \\ 0  & - D k_\perp^4\end{array} \right)
\eeq
in which case $\mathbf{B}_k$ is non-normal. In general, as the adiabaticity parameter is varied, the normality of $\mathbf{B}_k$ changes, but unfortunatley there is no absolute metric to quantify
the normality of a matrix or operator~\cite{trefethen2005}. In practice, however, what matters is not how non-normal $\mathbf{B}_k$ is, but how much transient energy growth is possible in the linear
system. The linear HW system is

\beq
\label{lin_HW}
\diff{u_k}{t} = \mathbf{B}_k u_k
\eeq
where $u_k = \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right)$. The solution to this $u_k(t) = e^{\mathbf{B}_k t} u(0)$ where the exponential of the matrix is defined in terms of
its Taylor expansion. The solution is dependent on the initial condition $u(0)$. The energy growth as a function of time is

\beq
\label{Gt}
G_k(t) = \frac{||u(t)||^2}{||u(0)||^2} = \frac{||e^{\mathbf{B}_k t} u(0)||^2}{||u(0)||^2}
\eeq
where the operator $|| \cdot ||$ is the $L_2$-norm. Different initial conditions provide optimal energy growth at different times, but the upper envelope of all such optimal growth curves is given by
$G_{k,\rm{max}}(t) = ||e^{\mathbf{B}_k t}||^2$, which does not depend on the initial condition.
Moreover, $||e^{\mathbf{B}_k t}||^2 \ge e^{2 \gamma_{s,k} t}$, where $\gamma_{s,k}$ is the real part of
the fastest growing (or least damped) eigenmode of $\mathbf{B}_k$ (and of $\mathbf{A}_k$, since $\mathbf{B}_k$ is just a similarity transform of $\mathbf{A}_k$~\cite{camargo1998}). 
The equality holds when $\mathbf{B}_k$ is normal, meaning the normal case provides a lower bound to the maximal energy growth. 
Thus, non-normality in practice allows a system to undergo more energy growth than is dictated by the most unstable linear eigenmode. We show this in Fig.~\ref{max_en_growth} where we plot
$||e^{\mathbf{B}_k t}||^2$ for different values of $\alpha$ along with $e^{2 \gamma_{s,k} t}$, showing that non-modal growth is possible for small adiabaticity parameter $\alpha$. Furthermore,
the short time transient growth becomes larger as $\alpha$ becomes smaller.

\begin{figure}
\centerline{\includegraphics[width=0.45\textwidth]{max_en_growth}}
\caption{$G_{k,\alpha,\rm{max}}(t) = ||e^{\mathbf{B}_{k,\alpha} t}||^2$ for $k_x=0, k_y=1$ for different values of the 2D adiabaticity parameter 
$\alpha$ compared to the energy growth $e^{2 \gamma_{s,\alpha,k} t}$ of the least stable eigenmode of  $\mathbf{B}_{k,\alpha}$. 
The time axis is normalized to $\gamma_{s,\alpha}$, which is different for each of the curves. This normalization makes the $e^{2 \gamma_{s,\alpha,k} t}$ line the same for all $\alpha$. 
$||e^{\mathbf{B}_{k,\alpha} t}||^2$ displays large non-exponential growth at early times that increases with decreasing $\alpha$.}
\label{max_en_growth}
\end{figure}

One explanation for this behavior is that non-normal matrices or operators have non-orthogonal linear eigenvectors. For the HW model, each Fourier matrix $\mathbf{B}_k$ has two eigenvectors associated
with it. They are complex and take the form:

\beq
\label{lin_eigen_form}
\psi_{k,q} = \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right)_q {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z} 
\eeq
where the subscript $q$ takes the values of $1$ and $2$, representing the two eigenvectors for each $k$. The two eigenvectors are different in that they have a different complex ratio $n_k/\phi_k$.
The two eigenvalues have different growth rates (the real part), but the same frequency, though with opposite sign.
As $\alpha \to 0$, the two eigenvectors become less orthogonal -- more parallel or anti-parallel to one another. When the eigenvectors are nearly anti-parallel and initialized with finite amplitude,
their superposition, which constitutes the total system at $k$, may linearly evolve so that the norm of the superposition grows faster than either individual eigenvector. A paradigmatic illustration
of this may be seen in a review by Schmid~\cite{schmid2007}.

\section{The Effect of Non-Normality on Turbulence}

Linear non-normality is important for the effect it has on turbulence. It is most often cited as the sufficient condition for turbulence in subcritical systems, 
but it can also impact systems that have unstable linear eigenmodes like the HW system. Since certain superpositions of the linear eigenmodes can inject energy faster than the most unstable linear
eigenmode, there is the possibility that the energy injection into the turbulence is faster and at different wavenumbers than that suggested by linear eigenmode analysis. 
To understand this concept and show that it occurs in the nonlinear HW model, we compare the fastest growing linear eigenmode growth rate $\gamma_{s,k}$ to the turbulent growth rate $\gamma_{t,k}$. 
First, however, we define the turbulent growth rate and show how to calcuate it for the HW model.
In order to find the turbulent growth rate for the HW model, we use a Fourier-decomposed energy dynamics analysis~\cite{camargo1995,friedman2012b,friedman2013}. 
To do this, we take Eqs.~\ref{n_eq} and~\ref{phi_eq} and substitute the following:

\beqar
\label{subs_fourier}
n = \sum_{k} n_k {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z} \\
\phi = \sum_{k} \phi_k {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z},
\eeqar
where again $k$ is shorthand for $(k_x,k_y,k_z)$.
Then, we multiply the resulting first equation by $n^*_{k'} {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$ and the second equation by 
$-\phi^*_{k'} {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$, take the volume integral and add the equations together. The result is

\beqar
\label{Explicit_En_eqn}
\frac{1}{2} \diff{}{t} \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right) = -i k_y \kappa \phi_k n^*_k - \xi k_z^2 |n_k - \phi_k|^2 \nonumber \\
- D k^4_\perp \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right) + \sum_{k'} T(k,k')
\eeqar
where, again, the energy $E_k$ is $\frac{1}{2} \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right)$, 
and the term $\sum_{k'} T(k,k')$ represents the nonlinear contribution (three-wave transfers between different $k$'s), 
which we do not write explicitly. This energy evolution equation may be written symbolically as

\beq
\label{dEdt_def}
\diff{E_k}{t} = \diff{E_{l,k}}{t} + \diff{E_{nl,k}}{t}
\eeq
where $\diff{E_{l,k}}{t}$ represents the terms that come from the linear terms in Eqs.~\ref{n_eq} and~\ref{phi_eq}. These are actually quadratic in the fluctuating
quantities $n_k$ and $\phi_k$ in Eq.~\ref{Explicit_En_eqn}. 
$\diff{E_{l,k}}{t}$ represents the injection of energy into the fluctuations from the free energy in the equilibrium gradients plus the dissipation from collisions and hyperdiffusion.
$\diff{E_{nl,k}}{t}$ represents $\sum_{k'} T(k,k')$ in  Eq.~\ref{Explicit_En_eqn} and comes from the nonlinear terms in 
Eqs.~\ref{n_eq} and~\ref{phi_eq}. The terms in $T(k.k')$ are triadic in the fluctuating quantities and
account for the energy exchange between fluctuations with different $k$. Furthermore, it is conservative: $\sum_{k} \diff{E_{nl,k}}{t} = \sum_{k,k'} T(k,k') = 0$.
Now, in quasi-steady state turbulence, the rate of net energy injection (or dissipation) into the fluctuations at each $k$ by the linear terms must be balanced by
the rate of energy removal (or deposition) from the nonlinear terms. This may be represented formally using growth rates:

\beqar
\label{steady_state}
\gamma_k \equiv  \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_k/dt}{2 E_k} dt \nonumber \\
= \displaystyle\lim_{T \to \infty} \frac{{\rm{Log}}\left[ E_k(T)/E_k(0) \right]}{2 T} = 0
\eeqar
where the equality to zero clearly holds as long as the turbulence is in a steady or quasi-steady state.
From, Eqs.~\ref{dEdt_def} and~\ref{steady_state}, it follows that $ \gamma_k = \gamma_{l,k} + \gamma_{nl,k} = 0$.

We now associate the linear rate of energy injection into the fluctuations $\gamma_{l,k}$ as the turbulent growth rate $\gamma_{t,k}$ discussed above. 
It should be clear now that this growth rate is the turbulence equivalent to the linear eigenmode growth rate $\gamma_{s,k}$~\cite{friedman2012b,terry2006b}. 
Moreover, $\gamma_{l,k}$ is calculated from the spatial structures of the plasma state variables, so it is always well-defined, even in a turbulent plasma. In a linear simulation run for a
sufficiently long time where transients die away, the calculation of $\gamma_{l,k}$ gives exactly $\gamma_{s,k}$.

In a turbulent state, which can be considered a linear superposition of linear eigenvectors, $\gamma_{t,k} \le \gamma_{s,k}$ if the system is normal. Recall that in normal systems, the maximal
energy growth is given by the most unstable linear eigenmode. In other words, maximal growth is achieved when the most stable eigenmode has zero amplitude. When this subdominant eigenmode has finite
amplitude, the growth rate must be lowered. On the other hand, non-normal systems may grow faster than is dictated by the most unstable eigenmode when the subdominant eigenmode has finite amplitude. 
It is therefore possible for $\gamma_{t,k} \ge \gamma_{s,k}$ in non-normal systems.

\begin{figure}
\centerline{\includegraphics[width=0.55\textwidth]{alpha1e-5_gamma_s_t}}
\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$ and b) the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 10^{-5}$. Note the different scales.}
\label{alpha1e-5_gamma_s_t}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.55\textwidth]{alpha1_gamma_s_t}}
\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$ and b) the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 1$.}
\label{alpha1_gamma_s_t}
\end{figure}

For the 2D HW model, we show
$\gamma_{s,k}$ and $\gamma_{t,k}$ for two different values of $\alpha$. In Fig.~\ref{alpha1e-5_gamma_s_t}, we show these growth rate spectra for $\alpha = 10^{-5}$, for which the system is
highly non-normal, and in Fig.~\ref{alpha1_gamma_s_t} for $\alpha = 1$, for which the system is very close to being normal. Note the different scales used in all of the figures. In the highly
non-normal case, the turbulent growth rate is somewhat higher than the eigenmode growth rate for most values of $k_x,k_y$, and about 6 times higher where the spectrum peaks. For the more normal
case, in contrast, $\gamma_{t,k}$ is generally lower than $\gamma_{s,k}$, with the peak at less than a factor of 2 less, though the peaks are at slightly different values of $k_y$.

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_alpha}}
\caption{}
\label{gamma_max_vs_alpha}
\end{figure}

We show the peak values for $\gamma_{s,k}$ and $\gamma_{t,k}$ in Fig.~\ref{gamma_max_vs_alpha} as a function of $\alpha$. The peak values always occur in $k$-space
at the lowest $k_x$ available to our system, corresponding to half of a period of a sine wave. The value of $k_y$ that gives the highest growth rate migrates upward as $\alpha$ increases,
as is clear from Figs.~\ref{alpha1e-5_gamma_s_t} and~\ref{alpha1_gamma_s_t}. Furthermore, as expected, when $\alpha$ is small $\lesssim 10^{-3}$, $\gamma_{t,k} > \gamma_{s,k}$, with
$\gamma_{t,k} \gg \gamma_{s,k}$ as $\alpha \to 0$. Additionally, when $\alpha$ becomes larger and the system becomes more normal ($\gtrsim 10^{-1}$), $\gamma_{t,k} < \gamma_{s,k}$.
For both the normal and highly non-normal limits of the 2D HW model, $\gamma_{s,k}$ does a poor job of predicting $\gamma_{t,k}$ because $\gamma_{s,k}$ neglects the stable branch of the
dispersion relation -- the damped eigenmode. The importance of the stable eigenmode branch has started to gain acceptance in the plasma community, although it has generally been used to explain
turbulent saturation, as is the case for nearly normal systems~\cite{terry2006}. The increased turbulent excitation in the highly non-normal case, however, is not as well-known but it also
important.

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_kz}}
\caption{}
\label{gamma_max_vs_kz}
\end{figure}

The nonlinear simulation reveals a fascinating property of the turbulence -- 
it is dominated by a nonlinear instability process despite being linearly unstable to drift waves~\cite{friedman2012b,friedman2013}.
The nonlinear instability, which was discovered by Drake et al.~\cite{drake1995}, works as follows: 
magnetic-field-aligned ($k_\para=0$) convective filaments transport density across the equilibrium density gradient, setting up $k_\para=0$ density filaments. 
These filaments are unstable to secondary drift waves, which grow on the periphery of the filaments. 
These drift waves, which have finite $k_\para$, nonlinearly couple to one another and generate new convective filaments.

Although the instability is called a nonlinear instability, the first part of the mechanism -- the transport of background density by the convective filaments -- is a linear one.
In fact, the other parts of the mechanism are driven by energetically conservative nonlinear interactions, 
meaning that the convective transport is the only step responsible for energy injection into the fluctuations.

This convective transport of density filaments is akin to the paradigmatic ``lift-up'' mechanism in hydrodynamic shear flows whereby streamwise vortices drive streamwise 
streaks~\cite{trefethen1993,krommes1999}.
Both are transient growth processes.

\section{Non-Modal Linear Technique of Turbulent Growth Rate}

We contend that the key to understanding and predicting turbulent properties through non-modal analysis is 
to successfully model the effect that the nonlinearities have on the transient linear processes. 
To this effect, we note that the advective nonlinearity in Eq.~\ref{fourier_en_eqn} has the form of the state vector divided by a time $\tau_{nl} \sim (v_E k_\perp)^{-1}$. This nonlinear
time scale is generally associated with the eddy turnover or decorrelation time. We therefore present a heuristic model of the nonlinearities 
as a randomizing force that acts on this characteristic nonlinear time scale. Again, this model is one in which 1) the turbulence begins as a random state, 
2) evolves linearly for a time $\tau_{nl}$, and 3) randomizes by nonlinear energy transfer, at which point the steps repeat.
In practice, we implement this model by starting with an ensemble of random initial conditions, which we evolve linearly for a time $\tau_{nl}$, 
and then take the time and ensemble averaged growth rate of these curves.
This procedure does not require actual linear simulations from an ensemble of random initial conditions. To see this, recall that the time evolution of the energy from an initial condition is

\beq
\label{E_t_from_u0}
E_k(t) = ||e^{\mathbf{B}_k t} u(0)||^2 = e^{\mathbf{B}_k t} u(0) u^{\dagger}(0) e^{\mathbf{B}_k^{\dagger}t}.
\eeq
If the $\mathbf{u}(0)$ in the ensemble are random with uncorrelated components and normalized to unity, it follows that~\cite{camargo1998}

\beq
\label{E_t_ensemble_avg}
\left< E_k(t)/E_k(0) \right>_{{\rm{ens}}} = \frac{1}{2} {\rm{tr}} \{ e^{\mathbf{B}_k t} e^{\mathbf{B}^{\dagger}t} \}.
\eeq
One point to note is that in global equation sets like ours -- where we discritize and use finite differences in the radial direction rather than a Fourier decomposition -- randomizing $\mathbf{u}(0)$
amounts to setting the initial $k_r$ spectrum to a step function that goes to zero at the Nyquist wavenumber. This means that $\left< E(t)/E(0) \right>_{{\rm{ens}}}$ may depend on $N_r$, so we must be careful in
choosing $N_r$ for this analysis. We choose $N_r$ such that the grid spacing equals $\rho_s$, the general Nyquist wavenumber of drift wave simulations~\cite{scott1992}.

Mathematically, our procedure is to calculate $\gamma_{\rm{n-m}}$ by the following formula:

\beq
\label{gamma_l_calc}
\gamma_{\rm{nm},k} = \frac{1}{\tau_{nl}} \int_0^{\tau_{nl}} \frac{\pdiff{E_k(t)}{t}}{2 E_k(t)} dt = \frac{1}{2 \tau_{nl}} {\rm{Log}} \left[ \frac{E_k(\tau_{nl})}{E_k(0)}\right]
\eeq
where $E(t)$ is the ensemble averaged energy calculated from Eq.~\ref{E_t_ensemble_avg}, and $E(0) = 1$ by our normalization.
In order to move toward predictive capability, we must estimate $\tau_{nl}$ with only knowledge of linear (modal or non-modal) information. 
We thus invoke the conjecture of \emph{critical balance}, which posits that the nonlinear time scale equals the linear time scale at all spatial scales~\cite{schekochihin2012}. This follows from 
the previously derived steady-state result: $\gamma_l = - \gamma_{nl}$.

Now there are two linear time scales that we may choose to test. The first linear time is the linear eigenmode frequency, labeled
$\omega_s^{-1}$. However, $\omega_s = 0$ for $k_z=0$ linear eigenmodes, so we are forced to use $\omega_s$ at the lowest finite $k_z$ to get a meaningful time scale.

Second, we use the steady-state condition $\gamma_l = - \gamma_{nl}$ and the approximation $\tau_{nl} \sim 1/|\gamma_{nl}|$ to get $\tau_{nl} = 1/|\gamma_{\rm{n-m}}|$. 
Inserting this into Eq.~\ref{gamma_l_calc} gives

\beq
\label{fourth_gamma_cond}
 {\rm{Log}} \left[ E_k(1/\gamma_{\rm{nm}})\right] = \pm 2 \ \rightarrow E_k(1/\gamma_{\rm{nm}}) = e^{\pm 2}.
\eeq
In other words, we find the time at which $E_k(t)$ grows to the value of $e^2$ or decays to the value of $e^{-2}$, and then use this time to get $\gamma_{nm}$.

\begin{figure}
\centerline{\includegraphics[width=0.55\textwidth]{alpha1e-5_gamma_spec_compare}}
\caption{}
\label{alpha1e-5_gamma_spec_compare}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.55\textwidth]{alpha1_gamma_spec_compare}}
\caption{}
\label{alpha1_gamma_spec_compare}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.55\textwidth]{gamma_max_with_nm}}
\caption{}
\label{gamma_max_with_nm}
\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_kz_with_nm}}
\caption{}
\label{gamma_max_vs_kz_with_nm}
\end{figure}

This work was supported by the National Science Foundation (Grant PHY-1202007)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\bibliographystyle{phaip}
\bibliography{refs}

\end{document}

### 2014-04-04 HW_non-modal.tex
April 4

@@
 
-Normal mode analysis -- the calculation of eigenvalues and eigenvectors of a linearized dynamical system -- has been used to solve many problems over the years.
-Despite its wide-ranging success, it has failed in important instances, particularly in predicting the onset of subcritical turbulence in hydrodynamic flows. 
-The reason for this failure was explained in the early 1990's when Trefethen and others attributed the pitfalls of normal mode analysis to the non-normality of linear operators of
+Turbulence in magnetically confined plasmas is almost always attributed to linear instabilities. The calculation of linear instabilities through normal mode analysis
+-- the calculation of eigenvalues and eigenvectors of a linearized dynamical system -- has successfully predicted the onset criterion for many MHD and micro-instabilities.
+Nevertheless, it is well-known in the hydrodynamics community that normal mode analysis fails at predicting turbulent onset for a large class of instabilities that are called subcritical.
+Subcritical turbulence is certainly possible in plasmas, and several specific cases have been studied in the past~\cite{camargo1998,camporeale2010,schekochihin2012}. 
+For the most part, however, normal mode analysis dominates plasma turbulence prediction.
+
+Since the hydrodynamics community has long studied subcritical turbulence, it has much to offer in terms of understanding the phenomenon. Most importantly,
+in the early 1990's, Trefethen and others attributed the failure of normal mode analysis in predicting subcritical turbulent onset to the non-normality of linear operators of
 dynamical systems~\cite{trefethen1993,schmid2007}. A non-normal operator has 
@@
 In other words, certain fluctuations of the laminar state can access free energy from background gradients even though normal mode fluctuations cannot.
-When combined with nonlinear effects, this allows for sustained subcritical turbulence.
+When combined with nonlinear effects, this allows for sustained subcritical turbulence by a bootstrapping process.
 Such behavior is obscured by traditional normal mode analysis, which only effectively describes the long time asymptotic behavior of fluctuations under  
@@
 
-Non-modal analysis has been embraced by the hydrodynamics community over the past two decades in the attempt to explain and predict subcritical turbulence. But the plasma community
-generally relies on normal mode analysis to inform turbulent predictions and observations, with a few exceptions~\cite{camargo1998,camporeale2010,schekochihin2012}. 
+Non-modal analysis generally employs tools such as pseudospectra and maximum possible transient growth curves, which have been used to successfully explain why turbulence is subcritical in various
+systems. However, to our knowledge, they have not been used to predict onset criteria in subcritical systems (e.g. the transition Reynold's number in pipe flows).
+
 Furthermore, non-modal treatments have generally been explanatory rather than predictive and have centered around the transition to turbulence in subcritical systems rather 
@@
 $\alpha = k_z^2 \xi = 4 \pi k_z^2 \rho_s^2 \omega_{ce} /\nu_{ei}$, which is commonly known as the adiabaticity parameter~\cite{camargo1995,camargo1998}.
+In nonlinear simulations, the turbulence can laminarize due to the build-up of $k_y=0$ density and potential structures~\cite{biskamp1995}. To prevent this, we subtract out the $k_y=0$
+$n$ and $\phi$ components in the simulations, which is the equivalent of adding source terms to Eqs.~\ref{n_eq} and~\ref{phi_eq}~\cite{friedman2012b}.
 
@@
 \frac{\partial}{\partial t} \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right) = \mathbf{A}_k \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right) + \mathbf{N}_k \left( \begin{array}{cc} n_k \\ \phi_k \end{array} \right), \\ \nonumber \\
+\label{A_k}
 \mathbf{A}_k = \left( \begin{array}{cc} -\xi k_z^2 - D k_\perp^4 & -i \kappa k_y + \xi k_z^2 \\  \xi \frac{k_z^2}{k_\perp^2} & - \xi \frac{k_z^2}{k_\perp^2} - D k_\perp^4\end{array} \right)
@@
 \frac{\partial}{\partial t} \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right) = \mathbf{B}_k \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right) + \mathbf{N}_k \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right), \\ \nonumber \\
+\label{B_k}
 \mathbf{B}_k = \left( \begin{array}{cc} -\xi k_z^2 - D k_\perp^4 & -i \kappa \frac{k_y}{k_\perp} + \xi \frac{k_z^2}{k_\perp} \\  \xi \frac{k_z^2}{k_\perp} & - \xi \frac{k_z^2}{k_\perp^2} - D k_\perp^4\end{array} \right)
@@
 \centerline{\includegraphics[width=0.45\textwidth]{max_en_growth}}
-\caption{$G_{k,\alpha,\rm{max}}(t) = ||e^{\mathbf{B}_{k,\alpha} t}||^2$ for $k_x=0, k_y=1$ for different values of the 2D adiabaticity parameter 
+\caption{$G_{k,\alpha,\rm{max}}(t) = ||e^{\mathbf{B}_{k,\alpha} t}||^2$ for $k_x=0, k_y=1, \kappa=1$ for different values of the 2D adiabaticity parameter 
 $\alpha$ compared to the energy growth $e^{2 \gamma_{s,\alpha,k} t}$ of the least stable eigenmode of  $\mathbf{B}_{k,\alpha}$. 
@@
 but it can also impact systems that have unstable linear eigenmodes like the HW system. Since certain superpositions of the linear eigenmodes can inject energy faster than the most unstable linear
-eigenmode, there is the possibility that the energy injection into the turbulence is faster and at different wavenumbers than that suggested by linear eigenmode analysis. 
+eigenmode, there is the possibility that the energy injection into the turbulence is faster and at different wavenumbers than is suggested by linear eigenmode analysis. 
 To understand this concept and show that it occurs in the nonlinear HW model, we compare the fastest growing linear eigenmode growth rate $\gamma_{s,k}$ to the turbulent growth rate $\gamma_{t,k}$. 
-First, however, we define the turbulent growth rate and show how to calcuate it for the HW model.
-In order to find the turbulent growth rate for the HW model, we use a Fourier-decomposed energy dynamics analysis~\cite{camargo1995,friedman2012b,friedman2013}. 
+To define the turbulent growth rate in general and specifically for the HW model, we employ a Fourier-decomposed energy dynamics analysis~\cite{camargo1995,friedman2012b,friedman2013}. 
 To do this, we take Eqs.~\ref{n_eq} and~\ref{phi_eq} and substitute the following:
@@
 \label{subs_fourier}
-n = \sum_{k} n_k {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z} \\
-\phi = \sum_{k} \phi_k {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z},
+n = \sum_{k} n_k \ {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z} \\
+\phi = \sum_{k} \phi_k \ {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z},
 \eeqar
-where again $k$ is shorthand for $(k_x,k_y,k_z)$.
-Then, we multiply the resulting first equation by $n^*_{k'} {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$ and the second equation by 
-$-\phi^*_{k'} {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$, take the volume integral and add the equations together. The result is
+where, again, $k$ stands for $(k_x,k_y,k_z)$.
+Then, we multiply the resulting first equation by $n^*_{k'} \ {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$ and the second equation by 
+$-\phi^*_{k'} \ {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$, take the volume integral and add the two equations together. The result is
 
@@
 \frac{1}{2} \diff{}{t} \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right) = -i k_y \kappa \phi_k n^*_k - \xi k_z^2 |n_k - \phi_k|^2 \nonumber \\
-- D k^4_\perp \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right) + \sum_{k'} T(k,k')
+- D k^4_\perp \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right) + \sum_{k'} T(k,k'), \quad \quad
 \eeqar
@@
 \eeq
-where $\diff{E_{l,k}}{t}$ represents the terms that come from the linear terms in Eqs.~\ref{n_eq} and~\ref{phi_eq}. These are actually quadratic in the fluctuating
+where $\diff{E_{l,k}}{t}$ represents the terms that come from the linear terms in Eqs.~\ref{n_eq} and~\ref{phi_eq}, which are quadratic in the fluctuating
 quantities $n_k$ and $\phi_k$ in Eq.~\ref{Explicit_En_eqn}. 
@@
 $\diff{E_{nl,k}}{t}$ represents $\sum_{k'} T(k,k')$ in  Eq.~\ref{Explicit_En_eqn} and comes from the nonlinear terms in 
-Eqs.~\ref{n_eq} and~\ref{phi_eq}. The terms in $T(k.k')$ are triadic in the fluctuating quantities and
-account for the energy exchange between fluctuations with different $k$. Furthermore, it is conservative: $\sum_{k} \diff{E_{nl,k}}{t} = \sum_{k,k'} T(k,k') = 0$.
+Eqs.~\ref{n_eq} and~\ref{phi_eq}. The terms in $T(k,k')$ are triadic in the fluctuating quantities and
+account for the energy exchange between fluctuations with different $k$'s. Furthermore, it is conservative: $\sum_{k} \diff{E_{nl,k}}{t} = \sum_{k,k'} T(k,k') = 0$.
+
 Now, in quasi-steady state turbulence, the rate of net energy injection (or dissipation) into the fluctuations at each $k$ by the linear terms must be balanced by
@@
 
-We now associate the linear rate of energy injection into the fluctuations $\gamma_{l,k}$ as the turbulent growth rate $\gamma_{t,k}$ discussed above. 
-It should be clear now that this growth rate is the turbulence equivalent to the linear eigenmode growth rate $\gamma_{s,k}$~\cite{friedman2012b,terry2006b}. 
-Moreover, $\gamma_{l,k}$ is calculated from the spatial structures of the plasma state variables, so it is always well-defined, even in a turbulent plasma. In a linear simulation run for a
-sufficiently long time where transients die away, the calculation of $\gamma_{l,k}$ gives exactly $\gamma_{s,k}$.
+We now associate the turbulent growth rate $\gamma_{t,k}$ discussed above with the linear rate of energy injection into the fluctuations $\gamma_{l,k}$ because $\gamma_{l.k}$
+determines how fast and where in wavenumber space energy is injected into the turbulence.
+$\gamma_{l,k}$, in other words, is the turbulence equivalent to the linear eigenmode growth rate $\gamma_{s,k}$~\cite{friedman2012b,terry2006b}. 
+That is, in a linear simulation run for a sufficiently long time where transients die away, the calculation of $\gamma_{l,k}$ gives exactly $\gamma_{s,k}$.
+Moreover, $\gamma_{l,k}$ is calculated from the spatial structures of the plasma state variables, so it is always well-defined, even in a turbulent plasma.
 
-In a turbulent state, which can be considered a linear superposition of linear eigenvectors, $\gamma_{t,k} \le \gamma_{s,k}$ if the system is normal. Recall that in normal systems, the maximal
-energy growth is given by the most unstable linear eigenmode. In other words, maximal growth is achieved when the most stable eigenmode has zero amplitude. When this subdominant eigenmode has finite
-amplitude, the growth rate must be lowered. On the other hand, non-normal systems may grow faster than is dictated by the most unstable eigenmode when the subdominant eigenmode has finite amplitude. 
-It is therefore possible for $\gamma_{t,k} \ge \gamma_{s,k}$ in non-normal systems.
+A turbulent system, which can be decomposed into a linear eigenvector basis, has $\gamma_{t,k} \le \gamma_{s,k}$ if the system is normal. Recall that in normal systems, the maximal
+energy growth is given by the most unstable linear eigenmode. In other words, maximal growth is achieved when the \emph{most stable} eigenmode has zero amplitude.
+On the other hand, non-normal systems may grow faster when the more stable eigenmode has finite amplitude than when it has none (see Fig.~\ref{max_en_growth}). 
+It is therefore possible for $\gamma_{t,k} > \gamma_{s,k}$ in non-normal systems.
 
 \begin{figure}
-\centerline{\includegraphics[width=0.55\textwidth]{alpha1e-5_gamma_s_t}}
-\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$ and b) the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 10^{-5}$. Note the different scales.}
+\centerline{\includegraphics[width=0.52\textwidth]{alpha1e-5_gamma_s_t}}
+\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$ and b) the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 10^{-5}, \kappa=1$. Note the different scales.}
 \label{alpha1e-5_gamma_s_t}
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.55\textwidth]{alpha1_gamma_s_t}}
-\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$ and b) the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 1$.}
+\centerline{\includegraphics[width=0.52\textwidth]{alpha1_gamma_s_t}}
+\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$ and b) the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 1, \kappa=1$.}
 \label{alpha1_gamma_s_t}
@@
 $\gamma_{s,k}$ and $\gamma_{t,k}$ for two different values of $\alpha$. In Fig.~\ref{alpha1e-5_gamma_s_t}, we show these growth rate spectra for $\alpha = 10^{-5}$, for which the system is
-highly non-normal, and in Fig.~\ref{alpha1_gamma_s_t} for $\alpha = 1$, for which the system is very close to being normal. Note the different scales used in all of the figures. In the highly
+highly non-normal, and in Fig.~\ref{alpha1_gamma_s_t} for $\alpha = 1$, for which the system is very close to being normal. All calculations in this section use $\kappa=1$. 
+Note the different scales used in all of the figures. In the highly
 non-normal case, the turbulent growth rate is somewhat higher than the eigenmode growth rate for most values of $k_x,k_y$, and about 6 times higher where the spectrum peaks. For the more normal
-case, in contrast, $\gamma_{t,k}$ is generally lower than $\gamma_{s,k}$, with the peak at less than a factor of 2 less, though the peaks are at slightly different values of $k_y$.
+case, in contrast, $\gamma_{t,k}$ is generally lower than $\gamma_{s,k}$, with the peak at nearly a factor of 2 less, with the peaks being at slightly different values of $k_y$.
 
@@
 \centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_alpha}}
-\caption{}
+\caption{The eigenmode $\gamma_{s}$ and turbulent $\gamma_{t}$ growth rates as a function of $\alpha$, which controls the normality of the linear operator.
+All growth rates are peaks in $k_x-k_y$ space.}
 \label{gamma_max_vs_alpha}
@@
 at the lowest $k_x$ available to our system, corresponding to half of a period of a sine wave. The value of $k_y$ that gives the highest growth rate migrates upward as $\alpha$ increases,
-as is clear from Figs.~\ref{alpha1e-5_gamma_s_t} and~\ref{alpha1_gamma_s_t}. Furthermore, as expected, when $\alpha$ is small $\lesssim 10^{-3}$, $\gamma_{t,k} > \gamma_{s,k}$, with
+as is clear from Figs.~\ref{alpha1e-5_gamma_s_t} and~\ref{alpha1_gamma_s_t}. Furthermore, as expected, when $\alpha$ is small ($\lesssim 10^{-3}$), $\gamma_{t,k} > \gamma_{s,k}$, with
 $\gamma_{t,k} \gg \gamma_{s,k}$ as $\alpha \to 0$. Additionally, when $\alpha$ becomes larger and the system becomes more normal ($\gtrsim 10^{-1}$), $\gamma_{t,k} < \gamma_{s,k}$.
+
 For both the normal and highly non-normal limits of the 2D HW model, $\gamma_{s,k}$ does a poor job of predicting $\gamma_{t,k}$ because $\gamma_{s,k}$ neglects the stable branch of the
-dispersion relation -- the damped eigenmode. The importance of the stable eigenmode branch has started to gain acceptance in the plasma community, although it has generally been used to explain
-turbulent saturation, as is the case for nearly normal systems~\cite{terry2006}. The increased turbulent excitation in the highly non-normal case, however, is not as well-known but it also
-important.
+dispersion relation -- the damped eigenmode. The importance of the stable eigenmode branch (or branches) 
+has started to gain acceptance in the plasma community, although it has generally been used in normal
+or nearly normal systems to explain turbulent saturation~\cite{terry2006b} ($\gamma_{t,k} < \gamma_{s,k}$). The case of highly non-normal turbulence and the associated increased turbulent excitation
+hasn't been understood as well. In fact, simulation and analysis of the 3D HW model (and extended models) is an area in which others have seen the effects of strong non-normal transient growth
+but have not realized them as such.
 
@@
 \centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_kz}}
-\caption{}
+\caption{The growth rates of the fastest growing eigenmode $\gamma_{s,k}$ and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model. The most remarkable feature is
+that the turbulent growth rate peaks at $k_z=0$ despite the stability of all linear eigenmodes at $k_z=0$.}
 \label{gamma_max_vs_kz}
@@
 
-The nonlinear simulation reveals a fascinating property of the turbulence -- 
-it is dominated by a nonlinear instability process despite being linearly unstable to drift waves~\cite{friedman2012b,friedman2013}.
-The nonlinear instability, which was discovered by Drake et al.~\cite{drake1995}, works as follows: 
-magnetic-field-aligned ($k_\para=0$) convective filaments transport density across the equilibrium density gradient, setting up $k_\para=0$ density filaments. 
-These filaments are unstable to secondary drift waves, which grow on the periphery of the filaments. 
-These drift waves, which have finite $k_\para$, nonlinearly couple to one another and generate new convective filaments.
+The 3D HW model nonlinear simulation has a fascinating property -- the turbulent energy injection is strongest at $k_z = 0$ despite the fact that there can be no unstable linear eigenmodes at $k_z=0$.
+We show this in Fig.~\ref{gamma_max_vs_kz} using the eigenmode $\gamma_{s,k}$ and turbulent $\gamma_{t,k}$ growth rates, plotting them as a function of $k_z$.
+The long-stated cause of this $k_z=0$ energy injection is a nonlinear instability, which was discovered by Biaskamp and Zeiler~\cite{biskamp1995} and explored further by Drake et al.~\cite{drake1995}.
+The nonlinear instability works as follows: magnetic-field-aligned ($k_z=0$) convective cells transport density across the equilibrium density gradient, setting up $k_z=0$ density structures. 
+These structures are unstable to drift waves, which grow as a secondary instability.
+The secondary drift waves, which have finite $k_z$, nonlinearly couple to one another and reinforce the original convective cells (or generate new ones), leading to self-sustainment.
 
-Although the instability is called a nonlinear instability, the first part of the mechanism -- the transport of background density by the convective filaments -- is a linear one.
-In fact, the other parts of the mechanism are driven by energetically conservative nonlinear interactions, 
-meaning that the convective transport is the only step responsible for energy injection into the fluctuations.
+Although the instability is a nonlinear instability, the first part of the mechanism -- the transport of background density by the convective cells -- is a linear one.
+Biskamp and Drake understood this and wondered how there could be linear energy injection despite the fact that the linear eigenvectors are stable.
+The answer to this came when Camargo et al.~\cite{camargo1998} stated that ``some properties of the drift-wave turbulence
+previously attributed to nonlinear terms could also be strongly influenced by the nonnormal character of the linear system.'' Although Camargo et al. focused only on the 2D HW model,
+based on this statement, it was likely apparent to them that the 3D HW nonlinear instability was a result of linear non-normality. 
+Therefore, when $k_z=0$, no matter the value of $\xi$, the adiabaticity parameter $\alpha = k_z^2 \xi = 0$, meaning the linear matrix operator is highly non-normal and subject to transient growth.
+So non-normal effects plays a role in the 3D HW system no matter what value of $\xi$ is chosen due to the presence of $k_z=0$ structures.
 
-This convective transport of density filaments is akin to the paradigmatic ``lift-up'' mechanism in hydrodynamic shear flows whereby streamwise vortices drive streamwise 
-streaks~\cite{trefethen1993,krommes1999}.
-Both are transient growth processes.
+\section{Non-Modal Linear Technique To Calculate Turbulent Growth Rate}
 
-\section{Non-Modal Linear Technique of Turbulent Growth Rate}
+In this section, we outline a procedure that uses quick and simple methods to approximate the turbulent growth rates $\gamma_{t,k}$.
+We contend that the key to doing this is to successfully model the effect that the nonlinearities have on the linear (modal and transient) processes.
+First, we note that the advective nonlinearity terms in Eqs.~\ref{n_eq} and~\ref{phi_eq} have the form of the state variable divided by a time $\tau_{nl} \sim (v_E k_\perp)^{-1}$. 
+This nonlinear time scale is generally associated with the eddy turnover or decorrelation time. We therefore present a heuristic model of the nonlinearities 
+as a randomizing force that acts on this characteristic nonlinear time scale. More specifically, this model is one in which 1) the turbulence begins as a random state, 
+2) evolves linearly for a time $\tau_{nl}$, and 3) randomizes by nonlinear energy transfer, at which point the steps repeat.
 
-We contend that the key to understanding and predicting turbulent properties through non-modal analysis is 
-to successfully model the effect that the nonlinearities have on the transient linear processes. 
-To this effect, we note that the advective nonlinearity in Eq.~\ref{fourier_en_eqn} has the form of the state vector divided by a time $\tau_{nl} \sim (v_E k_\perp)^{-1}$. This nonlinear
-time scale is generally associated with the eddy turnover or decorrelation time. We therefore present a heuristic model of the nonlinearities 
-as a randomizing force that acts on this characteristic nonlinear time scale. Again, this model is one in which 1) the turbulence begins as a random state, 
-2) evolves linearly for a time $\tau_{nl}$, and 3) randomizes by nonlinear energy transfer, at which point the steps repeat.
-In practice, we implement this model by starting with an ensemble of random initial conditions, which we evolve linearly for a time $\tau_{nl}$, 
-and then take the time and ensemble averaged growth rate of these curves.
-This procedure does not require actual linear simulations from an ensemble of random initial conditions. To see this, recall that the time evolution of the energy from an initial condition is
+This model can be used by starting with an ensemble of random initial conditions, evolving them linearly for a time $\tau_{nl}$, and
+then taking the time and ensemble averaged growth rate of these curves.
+In practice, this procedure can be greatly sped up and simplified. To see this, recall that the time evolution of the energy from an initial condition is
 
@@
 \eeq
-If the $\mathbf{u}(0)$ in the ensemble are random with uncorrelated components and normalized to unity, it follows that~\cite{camargo1998}
+Since we want to take the $\mathbf{u}(0)$ in the ensemble as random normalized vectors with uncorrelated components, it follows that~\cite{camargo1998}
 
@@
 \label{E_t_ensemble_avg}
-\left< E_k(t)/E_k(0) \right>_{{\rm{ens}}} = \frac{1}{2} {\rm{tr}} \{ e^{\mathbf{B}_k t} e^{\mathbf{B}^{\dagger}t} \}.
+\left< E_k(t) \right>_{{\rm{ens}}} = \frac{1}{2} {\rm{tr}} \{ e^{\mathbf{B}_k t} e^{\mathbf{B}^{\dagger}t} \},
 \eeq
-One point to note is that in global equation sets like ours -- where we discritize and use finite differences in the radial direction rather than a Fourier decomposition -- randomizing $\mathbf{u}(0)$
-amounts to setting the initial $k_r$ spectrum to a step function that goes to zero at the Nyquist wavenumber.
-This means that $\left< E(t)/E(0) \right>_{{\rm{ens}}}$ may depend on $N_r$, so we must be careful in
-choosing $N_r$ for this analysis. We choose $N_r$ such that the grid spacing equals $\rho_s$, the general Nyquist wavenumber of drift wave simulations~\cite{scott1992}.
-
-Mathematically, our procedure is to calculate $\gamma_{\rm{n-m}}$ by the following formula:
+where the normalization is $||\mathbf{u}(0)||^2 = 1$. Mathematically, we define the non-modal growth rate $\gamma_{{\rm{nm}},k}$ as the time averaged instantaneous growth rate:
 
 \beq
-\label{gamma_l_calc}
-\gamma_{\rm{nm},k} = \frac{1}{\tau_{nl}} \int_0^{\tau_{nl}} \frac{\pdiff{E_k(t)}{t}}{2 E_k(t)} dt = \frac{1}{2 \tau_{nl}} {\rm{Log}} \left[ \frac{E_k(\tau_{nl})}{E_k(0)}\right]
+\label{gamma_nm_calc}
+\gamma_{{\rm{nm}},k} = \frac{1}{\tau_{nl,k}} \int_0^{\tau_{nl,k}} \frac{\pdiff{E_k(t)}{t}}{2 E_k(t)} dt = \frac{1}{2 \tau_{nl,k}} \ {\rm{Log}} \left[ \frac{E_k(\tau_{nl,k})}{E_k(0)}\right].
 \eeq
-where $E(t)$ is the ensemble averaged energy calculated from Eq.~\ref{E_t_ensemble_avg}, and $E(0) = 1$ by our normalization.
-In order to move toward predictive capability, we must estimate $\tau_{nl}$ with only knowledge of linear (modal or non-modal) information. 
-We thus invoke the conjecture of \emph{critical balance}, which posits that the nonlinear time scale equals the linear time scale at all spatial scales~\cite{schekochihin2012}. This follows from 
-the previously derived steady-state result: $\gamma_l = - \gamma_{nl}$.
-
-Now there are two linear time scales that we may choose to test. The first linear time is the linear eigenmode frequency, labeled
-$\omega_s^{-1}$. However, $\omega_s = 0$ for $k_z=0$ linear eigenmodes, so we are forced to use $\omega_s$ at the lowest finite $k_z$ to get a meaningful time scale.
-
-Second, we use the steady-state condition $\gamma_l = - \gamma_{nl}$ and the approximation $\tau_{nl} \sim 1/|\gamma_{nl}|$ to get $\tau_{nl} = 1/|\gamma_{\rm{n-m}}|$. 
-Inserting this into Eq.~\ref{gamma_l_calc} gives
+Next, we substitute the ensemble averaged energy calculated from Eq.~\ref{E_t_ensemble_avg} for $E_k(t)$, and set $E_k(0) = 1$ by our normalization. Then,
 
 \beq
-\label{fourth_gamma_cond}
- {\rm{Log}} \left[ E_k(1/\gamma_{\rm{nm}})\right] = \pm 2 \ \rightarrow E_k(1/\gamma_{\rm{nm}}) = e^{\pm 2}.
+\label{gamma_nm}
+\gamma_{{\rm{nm}},k} = \frac{1}{2 \tau_{nl,k}} \ {\rm{Log}} \left<  E_k(\tau_{nl,k}) \right>_{\rm{ens}}
 \eeq
-In other words, we find the time at which $E_k(t)$ grows to the value of $e^2$ or decays to the value of $e^{-2}$, and then use this time to get $\gamma_{nm}$.
+Now, in order to obtain a simple and quick prediction of $\gamma_{{\rm{nm}},k}$, we must estimate $\tau_{nl,k}$ without having to calculate it directly from a simulation. 
+We thus invoke the conjecture of \emph{critical balance}, which posits that the nonlinear time scale equals the linear time scale at all spatial scales~\cite{schekochihin2012}. This is a fair
+assumption in light of our discussion above, in which we derived the relation: $\gamma_{l,k} = - \gamma_{nl,k}$.
+
+There are two fairly obvious linear time scales in the HW problem. First is the inverse of the linear eigenmode frequency $\omega_{s,k}$
+-- the imaginary part of the eigenvalues of the linear matrix operator $\mathbf{B}_k$.
+Recall that the imaginary part of the linear eigenvalues is the same for both eigenvalues of $\mathbf{B}_k$, so there is no question of which to use. We thus define the first non-modal
+growth rate $\gamma_{\omega,k}$ as
+
+\beq
+\label{gamma_omega_def}
+\gamma_{\omega,k} = \frac{\omega_{s,k}}{2} \ {\rm{Log}} \left< E_k(1/\omega_{s,k}) \right>_{\rm{ens}}.
+\eeq
+This growth rate is calculated by first finding $\omega_{s,k}$ of $\mathbf{B}_k$, and then inserting this into Eq.~\ref{gamma_omega_def} using the definition of 
+$\left< E_k(1/\omega_{s,k}) \right>_{\rm{ens}}$ from Eq.~\ref{E_t_ensemble_avg}.
+
+The second linear time scale that we may use is the inverse of the linear growth rate. However, the eigenvalues of $\mathbf{B}_k$ have different growth rates (real parts). 
+Furthermore, we have argued that the eigenmode growth rates have little bearing on turbulence, especially for highly non-normal turbulence, 
+so it would be innappropriate to use either of the linear eigenmode growth rates as a proxy for the nonlinear time scale.
+Rather, we use the result itself, $\gamma_{{\rm{nm}},k}$, as the inverse of $\tau_{nl,k}$. Substituting $1/|\gamma_{{\rm{nm}},k}|$ for $\tau_{nl,k}$ in Eq.~\ref{gamma_nm_calc}, we get
+a transcendental equation:
+
+\beq
+\label{trans_gamma_nm}
+\gamma_{{\rm{nm}},k} = \frac{|\gamma_{{\rm{nm}},k}|}{2} \ {\rm{Log}} \left< E_k(1/|\gamma_{{\rm{nm}},k}|) \right>_{\rm{ens}}.
+\eeq
+We label this second non-modal growth rate $\gamma_{\gamma,k}$, and ultimately simplify this equation to
 
 \begin{figure}
-\centerline{\includegraphics[width=0.55\textwidth]{alpha1e-5_gamma_spec_compare}}
-\caption{}
+\centerline{\includegraphics[width=0.52\textwidth]{alpha1e-5_gamma_spec_compare}}
+\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$,  b) the turbulent growth rate spectrum $\gamma_{t,k}$, c) the non-modal growth rate spectrum $\gamma_{\omega,k}$ d) and the 
+non-modal growth rate spectrum $\gamma_{\gamma,k}$ for the 2D HW model with $\alpha = 10^{-5}, \kappa=1$. Note the different scales.}
 \label{alpha1e-5_gamma_spec_compare}
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.55\textwidth]{alpha1_gamma_spec_compare}}
-\caption{}
+\centerline{\includegraphics[width=0.52\textwidth]{alpha1_gamma_spec_compare}}
+\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$,  b) the turbulent growth rate spectrum $\gamma_{t,k}$, c) the non-modal growth rate spectrum $\gamma_{\omega,k}$ d) and the 
+non-modal growth rate spectrum $\gamma_{\gamma,k}$ for the 2D HW model with $\alpha = 1, \kappa=1$.}
 \label{alpha1_gamma_spec_compare}
@@
 
+\beq
+\label{gamma_pm2}
+\left< E_k(1/|\gamma_{\gamma,k}|) \right>_{\rm{ens}} = e^{\pm 2},
+\eeq
+where if $\left< E_k(t) \right>_{\rm{ens}}$ first reaches $e^2$, $\gamma_{\gamma,k}$ is positive, and if it first reaches $e^{-2}$, $\gamma_{\gamma,k}$ is negative.
+Therefore, finding $\gamma_{\gamma,k}$ amounts to a root-finding problem, which is not trivial, but is much simpler than a nonlinear simulation.
+
+We test both of these non-modal growth rates to see if either or both produce good approximations of $\gamma_{t,k}$, especially in comparison to $\gamma_{s,k}$. 
+In Fig.~\ref{alpha1e-5_gamma_spec_compare}, we show the $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$ spectra for the 2D HW
+model for $\alpha=10^{-5}$ along with the eigenmode and turbulent spectra ($\gamma_{s,k}$ and $\gamma_{t,k}$), which we already provided in Fig.~\ref{alpha1e-5_gamma_s_t}. We do the same for
+$\alpha=1$ in Fig.~\ref{alpha1_gamma_s_t}. For the highly non-normal case ($\alpha = 10^{-5}$), $\gamma_{\omega,k}$ reproduces $\gamma_{t,k}$ in both magnitude and spectral shape. $\gamma_{\gamma,k}$
+is not close in any respect to $\gamma_{t,k}$. On the other hand, for the nearly normal case ($\alpha=1$), $\gamma_{\omega,k}$ does not resemble $\gamma_{t,k}$. $\gamma_{\gamma,k}$ actually is
+fairly similar to $\gamma_{s,k}$, but with a slightly decreased amplitude, making it a closer match to $\gamma_{t,k}$, though it is still not similar for all $k$.
+
 \begin{figure}
-\centerline{\includegraphics[width=0.55\textwidth]{gamma_max_with_nm}}
-\caption{}
+\centerline{\includegraphics[width=0.45\textwidth]{gamma_max_with_nm}}
+\caption{Comparison of the non-modal growth rates, $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$, to the eigenmode $\gamma_{s}$ and turbulent $\gamma_{t}$ growth rates as a function of $\alpha$
+for the 2D HW model. All growth rates are peaks in $k_x-k_y$ space.}
 \label{gamma_max_with_nm}
@@
 
+We further test how well the maxima of the non-modal growth rates predict the maxima of the turbulent growth rates. We show the result in Fig.~\ref{gamma_max_with_nm}. For $\alpha < 10^{-3}$,
+in which the system is fairly non-normal, $\gamma_{\omega,k}$ predicts $\gamma_{t,k}$ quite accurately. As the system becomes more normal, $\gamma_{t,k}$ and $\gamma_{\omega,k}$ diverge, with
+$\gamma_{t,k}$ taking on values between $\gamma_{s,k}$ and $\gamma_{\omega,k}$, but generally closer to $\gamma_{s,k}$ for $\alpha \gg 10^{-3}$. 
+This indicates that some assumption in our procedure doesn't work well for systems that are normal or nearly normal. Namely, our assumption
+of complete randomization by the nonlinearities breaks down. Apparently, the nonlinearities only partially randomize the turbulence. For the most part, the turbulence at a given $k$ consists
+primarily of the unstable eigenmode, while the stable eigenmode accounts for only a small amount of energy in the turbulence. Finally, $\gamma_{\gamma,k}$ is not useful in predicting $\gamma_{t,k}$.
+It is closest to $\gamma_{t,k}$ for $\alpha \ge 1$, however, it only provides a slight improvement over $\gamma_{s,k}$ in this range.
+
 \begin{figure}
 \centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_kz_with_nm}}
-\caption{}
+\caption{Comparison of the non-modal growth rates, $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$, to the growth rates of the fastest growing eigenmode $\gamma_{s,k}$ 
+and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model. The non-modal growth rates correctly predict the growth rate dominance of the $k_z=0$ mode
+despite the $k_z=0$ eigenmode stability.}
 \label{gamma_max_vs_kz_with_nm}
 \end{figure}
+
+We do the same comparison for the 3D HW model as we did for the 2D HW model. However, there is one additional subtelty when considering the 3D model. In the 3D model, for $k_z=0$, the linear
+eigenmode frequency is zero ($\omega_{s,k_z=0} = 0$), so $\gamma_{\omega,k_z=0}$ is not well defined (see Eq.~\ref{gamma_omega_def}). 
+We could, in fact, make the argument that since $\omega_{s,k_z=0} = 0$, or by critical balance $\tau_{nl,k_z=0} \to \infty$, it would be proper to equate $\gamma_{\omega,k_z=0}$ with $\gamma_{s,k_z=0}$
+since all transients die away in the long time limit. Rather, we use $\omega_{s,k_z=0.02}$ -- where $k_z=0.02$ is the lowest finite value of $k_z$ available to our system -- in place of
+$\omega_{s,k_z=0}$ in the calculation of $\gamma_{\omega,k_z=0}$. We contend that the nonlinear instability, which most strongly couples $k_z=0$ and $k_z=0.02$ modes, has its time scale set
+by the drift waves, not the convective cells, making $1/\omega_{s,k_z=0.02}$ a good proxy for $\tau_{nl,k_z=0}$. One may recall that the drift waves are actually secondary, and therefore, they feed
+off of the $k_z=0$ density structures, so the appropriate scale length to use for $\omega_{s,k_z=0.02}$ is not given by $\kappa$, but by the scale length of the density structures. We don't deal
+with this complication and assume the scale length of the fluctuating density structures is the same as the background.
+
+Despite these complications at $k_z=0$, our non-modal method of calculating $\gamma_{\omega,k}$ approximates $\gamma_{t,k}$ remarkably well as seen in Fig.~\ref{gamma_max_vs_kz_with_nm}.
+Most importantly, it correctly predicts that $\gamma_{\omega,k_z=0} > 0$ and $\gamma_{\omega,k_z=0} > \gamma_{\omega,k_z \ne 0}$. Therefore, non-modal calculations explain why the 3D HW model
+is dominated by the nonlinear instability rather than a linear drift wave instability. Simply put, transient growth is strongest at $k_z=0$, so the nonlinear instability is energetically
+favored over the linear instability. Again, $\gamma_{\gamma,k}$ does not provide enough accuracy to warrant its use.
 

### 2014-04-12 HW_non-modal.tex
April 12

@@
 -- the calculation of eigenvalues and eigenvectors of a linearized dynamical system -- has successfully predicted the onset criterion for many MHD and micro-instabilities.
-Nevertheless, it is well-known in the hydrodynamics community that normal mode analysis fails at predicting turbulent onset for a large class of instabilities that are called subcritical.
-Subcritical turbulence is certainly possible in plasmas, and several specific cases have been studied in the past~\cite{camargo1998,camporeale2010,schekochihin2012}. 
-For the most part, however, normal mode analysis dominates plasma turbulence prediction.
+Nevertheless, it is well-known in the hydrodynamics community that normal mode analysis fails at predicting turbulent onset for a large class of experiments that are deemed to have
+subcritical turbulence. Subcritical turbulence is certainly possible in plasmas, and several instances have been studied in the 
+past~\cite{waltz1985,scott1990,scott1992,nordman1993,biskamp1995,drake1995,camargo1998,krommes1999,scott2002,camporeale2009,camporeale2010,schekochihin2012,highcock2012}. 
+For the most part, however, normal mode analysis dominates plasma turbulence prediction despite the fact that it is not useful in all instances.
 
-Since the hydrodynamics community has long studied subcritical turbulence, it has much to offer in terms of understanding the phenomenon. Most importantly,
+Since the hydrodynamics community has long studied subcritical turbulence, it has much to offer in terms of understanding the phenomenon. Most significantly,
 in the early 1990's, Trefethen and others attributed the failure of normal mode analysis in predicting subcritical turbulent onset to the non-normality of linear operators of
@@
 Non-modal analysis generally employs tools such as pseudospectra and maximum possible transient growth curves, which have been used to successfully explain why turbulence is subcritical in various
-systems. However, to our knowledge, they have not been used to predict onset criteria in subcritical systems (e.g. the transition Reynold's number in pipe flows).
-
-Furthermore, non-modal treatments have generally been explanatory rather than predictive and have centered around the transition to turbulence in subcritical systems rather 
-than on properties of fully-developed turbulence.
-This paper takes up the task of developing an approach to understand turbulent properties using only non-modal linear calculations with the goal of 
+systems. However, to our knowledge, they have not been used to predict onset criteria in subcritical systems (e.g. the transition Reynold's number in pipe flows), although non-normal
+turbulent arguments have been applied in generating amplitude versus Reynold's number scalings for turbulent onset.
+This paper takes up the task of developing a new approach to understand turbulent properties using only non-modal linear calculations with the goal of 
 ultimately making quantitative predictions.  Our approach is to calculate an average growth rate of turbulent fluctuations due to linear processes, specifically due to transient growth.  We model the
@@
 nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
-arguments to use a characteristic \emph{linear} time. Since there is no obvious single linear time scale, we test several and compare the results to determine which works best.
-The procedure ultimately produces a growth rate that can be used
+arguments to use a characteristic \emph{linear} time. The procedure ultimately produces a growth rate that can be used
 to predict turbulent properties such as saturation levels and transport rates through mixing length arguments.
@@
 points opposite the density gradient is defined as the $x$ direction, and the $y$ direction is taken perpendicular to both of these.
+
 In the analysis and numerical simulations, which we perform with the BOUT++ code~\cite{dudson2009}, we use periodic boundary conditions in
-the $y$ and $z$ directions with zero-value boundaries in the $x$ direction. For the 2D version of the model, we replace $-\xi \gradpar^2$ by 
-$\alpha = k_z^2 \xi = 4 \pi k_z^2 \rho_s^2 \omega_{ce} /\nu_{ei}$, which is commonly known as the adiabaticity parameter~\cite{camargo1995,camargo1998}.
-In nonlinear simulations, the turbulence can laminarize due to the build-up of $k_y=0$ density and potential structures~\cite{biskamp1995}. To prevent this, we subtract out the $k_y=0$
-$n$ and $\phi$ components in the simulations, which is the equivalent of adding source terms to Eqs.~\ref{n_eq} and~\ref{phi_eq}~\cite{friedman2012b}.
+the $y$ and $z$ directions with zero-value boundaries in the $x$ direction. We use a domain size of $64 \rho_s \times 64 rho_s \times 300 \rho_s$.
+For the 2D version of the model, we replace $-\xi \gradpar^2$ by $\alpha = k_z^2 \xi = 4 \pi k_z^2 \rho_s^2 \omega_{ce} /\nu_{ei}$, 
+which is commonly known as the adiabaticity parameter~\cite{camargo1995,camargo1998}. Additionally we use a domain size of $50 \rho_s \times 50 \rho_s$.
+In numerical simulations, the turbulence can laminarize due to the build-up of $k_y=0$ density and potential structures~\cite{biskamp1995}. To prevent this, we subtract out the $k_y=0$
+$n$ and $\phi$ components, which is the equivalent of adding source terms to Eqs.~\ref{n_eq} and~\ref{phi_eq}~\cite{friedman2012b}.
 
@@
 
+\section{Prediction of Control Parameter for Subcritical Turbulent Onset}
+
+In both hydrodynamic and plasma systems, it is important to be able to predict the smallest value of a control parameter for which the system will become turbulent. 
+The control parameter is different for different systems (and models of those systems), and there may be several control parameters for a particular problem.
+When $d$ control parameters exist in a model,
+there is not a single point in the parameter space at which the system bifurcates to turbulence, rather, there is a $d-1$ manifold that separates the laminar and turbulent states.
+In many hydrodynamic systems driven turbulent by shear flow, there is generally a single control parameter -- the Reynold's number $Re$~\cite{drazin1981}.
+In plasmas, there are often multiple control parameters.
+For instance, kinetic solar wind turbulence models have two: $\beta$ and $T_\perp/T_\para$~\cite{camporeale2010}).
+One kinetic turbulence in the
+core of a tokamak has three: $\gamma_E$ (the normalized perpendicular flow shear), $R/L_T$ (the inverse ion temperature gradient scale length normalized to the major radius), and $q/\epsilon$
+(the ratio of the magnetic safety factor to the inverse aspect ratio)~\cite{highcock2012}.
+
+For normal systems, turbulence will be excited when the control parameter becomes large enough such that one of the linear eigenmodes becomes unstable (the real part of the eigenvalue crosses into the
+right half of the complex plane). In general, this is not the case for non-normal systems. In many non-normal systems, turbulence can be sustained when all linear eigenmodes are stable --
+the case of subcritical turbulence.
+
+\begin{figure}
+\centerline{\includegraphics[width=0.4\textwidth]{subcritical_diagram}}
+\caption{Diagram of a control parameter $R$ versus perturbation amplitude $A$ for two subcritical systems. The first curve (top blue one with $R^{-\Lambda}$ label) represents a system with no
+eigenmode instability for any $R$, but which is stable to finite amplitude perturbations for $R > R_g$. 
+The second (bottom red one with label of ``Eigenmode Unstable'') represents a system with a linear eigenmode instability for $R \ge R_c$, and finite amplitude
+instability for $R_g < R < R_c$.}
+\label{subcritical_diagram}
+\end{figure}
+
+Subcritical systems fall into two categories: those that have stable eigenmodes for all values of the control parameter, and those that have unstable eigenmodes
+when the control parameter becomes large enough. We illustrate these two categories with a diagram in Fig.~\ref{subcritical_diagram}. 
+The upper curve represents the first kind of subcritical system in which no unstable eigenmodes exist for any value of $R$. In this
+case, there is some transitional value of $R$, labeled $R_g$, below which turbulence cannot occur.
+When a perturbation with amplitude above the line occurs in the system due to external input or noise,
+turbulence will be excited. Perturbations with amplitude below the line will decay away. Clearly, the higher the value of $R$, the smaller the perturbation needs to be to excite turbulence.
+Experimental evidence and theoretical arguments show this curve should have a $R^{-\Lambda}$ dependence, with $\Lambda \sim 1$~\cite{grossmann2000}.
+A paradigmatic example of this kind of system is Hagen-Poisuille pipe flow -- flow in a cylindrical pipe with a parabolic flow profile -- for which $R_g \approx 2000$.
+
+The lower curve represents the second kind of subcritical system. Above a critical value of the control parameter $R_c$, the system is linearly unstable. But, for $R_g < R < R_c$, the system is
+unstable to finite amplitude perturbations, or ``nonlinearly'' unstable. A paradigmatic example of this is Poissuille channel flow -- the rectangular analog of pipe flow -- for which
+$R_c = 5772$ and $R_g \approx 1000$~\cite{grossmann2000}.
+Most subcritical magnetically-confined plasma systems fall into this category.
+It is perhaps easy to be fooled into believing that the turbulence
+in such systems is due to eigenmode instability. In either case, one of the most important predictions is the minimum turbulent control parameter $R_g$.
+
+The HW model in an unsheared magnetic slab is not subcritical in either the 2D or 3D cases, as implied by Figs.~\ref{gamma_max_vs_alpha} and~\ref{gamma_max_vs_kz}.
+It is subcritical in a highly sheared
+magnetic field~\cite{drake1995}, but the eigenmodes of the linear operator in that case are not sinusoidal, making our non-modal analysis more difficult.
+Biskamp et al., however, artificially modified the
+3D HW model to remove the linear drive by taking the $z$ average of the $\kappa \pdiff{\phi}{y}$ term in Eq.~\ref{n_eq}, effectively eliminating the $k_z \ne 0$ drive. This leaves only the $k_z = 0$
+component of the drive term. Since the $k_z = 0$ eigenmodes are stable but $k_z = 0$ structures can access the free energy (positive $\gamma_t$ in Fig.~\ref{gamma_max_vs_kz}), this system is subject
+to subcritical turbulence.
+
+We show that the system is subcritical in Fig.~\ref{init_en_ev} by plotting the energy versus time for fluctuations that are initialized with different amplitudes.
+When the fluctuations are initialized
+with high enough amplitude, the system eventually becomes turbulent.
+Otherwise, the fluctuations grow transiently, but do not reach high enough amplitude for the nonlinearities to bootstrap the process,
+so they decay away. The finite amplitude turbulence threashold, as described above, is the key indicator of subcritical turbulence.
+
+\begin{figure}
+\centerline{\includegraphics[width=0.4\textwidth]{init_en_ev}}
+\caption{Total energy evolution of fluctuations of the modified 3D HW model where the initial perturbations are set to different amplitudes. All initial perturbations are initialized randomly in
+the $y$ and $z$ directions and with a Guassian in the $x$ direction. There is a critical initial amplitude threashold above which turbulence develops and maintains itself
+and below which turbuelence never develops and the fluctuations decay away. This indicates that the modified model is subject to subcritical turbulence.}
+\label{init_en_ev}
+\end{figure}
+
+\begin{acknowledgments}
 This work was supported by the National Science Foundation (Grant PHY-1202007)
+\end{acknowledgments}
 

### 2014-04-15 HW_non-modal.tex
April 14

@@
 -- the calculation of eigenvalues and eigenvectors of a linearized dynamical system -- has successfully predicted the onset criterion for many MHD and micro-instabilities.
-Nevertheless, it is well-known in the hydrodynamics community that normal mode analysis fails at predicting turbulent onset for a large class of experiments that are deemed to have
-subcritical turbulence. Subcritical turbulence is certainly possible in plasmas, and several instances have been studied in the 
+Nevertheless, it is well-known in the hydrodynamics community that normal mode analysis fails at predicting turbulent onset for a large class of experiments, in which the
+turbulence is called subcritical. In these experiments, turbulence appears even though all linear eigenvectors are stable, meaning infinitesimal perturbations on the laminar state cannot grow
+exponentially. Nevertheless, finite amplitude perturbations can still excite turbulence.
+Subcritical turbulence is certainly possible in plasmas as well, and several instances have been studied in the 
 past~\cite{waltz1985,scott1990,scott1992,nordman1993,biskamp1995,drake1995,camargo1998,krommes1999,scott2002,camporeale2009,camporeale2010,schekochihin2012,highcock2012}. 
@@
 
-Since the hydrodynamics community has long studied subcritical turbulence, it has much to offer in terms of understanding the phenomenon. Most significantly,
-in the early 1990's, Trefethen and others attributed the failure of normal mode analysis in predicting subcritical turbulent onset to the non-normality of linear operators of
+Since the hydrodynamics community has long studied subcritical turbulence, it has much to offer in terms of understanding of the phenomenon. Most significantly,
+in the early 1990s, Trefethen and others attributed the failure of normal mode analysis in predicting subcritical turbulent onset to the non-normality of linear operators of
 dynamical systems~\cite{trefethen1993,schmid2007}. A non-normal operator has 
@@
 Non-modal analysis generally employs tools such as pseudospectra and maximum possible transient growth curves, which have been used to successfully explain why turbulence is subcritical in various
-systems. However, to our knowledge, they have not been used to predict onset criteria in subcritical systems (e.g. the transition Reynold's number in pipe flows), although non-normal
-turbulent arguments have been applied in generating amplitude versus Reynold's number scalings for turbulent onset.
+systems. However, they have not been used to predict onset criteria in subcritical systems (e.g. the transition Reynold's number in pipe flows), although non-normal
+turbulent arguments have been applied in generating amplitude versus Reynold's number scalings for turbulent onset~\cite{grossmann2000}.
 This paper takes up the task of developing a new approach to understand turbulent properties using only non-modal linear calculations with the goal of 
@@
 In the analysis and numerical simulations, which we perform with the BOUT++ code~\cite{dudson2009}, we use periodic boundary conditions in
-the $y$ and $z$ directions with zero-value boundaries in the $x$ direction. We use a domain size of $64 \rho_s \times 64 rho_s \times 300 \rho_s$.
-For the 2D version of the model, we replace $-\xi \gradpar^2$ by $\alpha = k_z^2 \xi = 4 \pi k_z^2 \rho_s^2 \omega_{ce} /\nu_{ei}$, 
+the $y$ and $z$ directions with zero-value boundaries in the $x$ direction. We use a domain size of $64 \rho_s \times 64 \rho_s \times 300 \rho_s$.
+For the 2D version of the model, we replace $-\xi \gradpar^2$ by $\alpha (= k_z^2 \xi = 4 \pi k_z^2 \rho_s^2 \omega_{ce} /\nu_{ei})$, 
 which is commonly known as the adiabaticity parameter~\cite{camargo1995,camargo1998}. Additionally we use a domain size of $50 \rho_s \times 50 \rho_s$.
@@
 Since the square $L_2$-norm of the new state vector $\frac{1}{\sqrt{2}} \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right)$ gives the energy contained in the mode $k$,
-$\mathbf{B}_k$ is the energetically-weighted matrix of interest. In the limit of large $\xi k_z^2$ as long as $k_\perp$ is also not too large:
+$\mathbf{B}_k$ is the energetically-weighted matrix of interest. In the limit of large $\xi k_z^2$ (as long as $k_\perp$ is also not too large):
 
@@
 
-Linear non-normality is important for the effect it has on turbulence. It is most often cited as the sufficient condition for turbulence in subcritical systems, 
-but it can also impact systems that have unstable linear eigenmodes like the HW system. Since certain superpositions of the linear eigenmodes can inject energy faster than the most unstable linear
+Linear non-normality is important for the effect it has on turbulence. It is most often cited as a necessary condition for turbulence in subcritical systems. 
+But it can also impact systems that have unstable linear eigenmodes like the HW system. Since certain superpositions of the linear eigenmodes can inject energy faster than the most unstable linear
 eigenmode, there is the possibility that the energy injection into the turbulence is faster and at different wavenumbers than is suggested by linear eigenmode analysis. 
@@
 Therefore, when $k_z=0$, no matter the value of $\xi$, the adiabaticity parameter $\alpha = k_z^2 \xi = 0$, meaning the linear matrix operator is highly non-normal and subject to transient growth.
-So non-normal effects plays a role in the 3D HW system no matter what value of $\xi$ is chosen due to the presence of $k_z=0$ structures.
+So non-normal effects play a role in the 3D HW system for all $\xi$ due to the presence of $k_z=0$ structures.
 
@@
 We contend that the key to doing this is to successfully model the effect that the nonlinearities have on the linear (modal and transient) processes.
-First, we note that the advective nonlinearity terms in Eqs.~\ref{n_eq} and~\ref{phi_eq} have the form of the state variable divided by a time $\tau_{nl} \sim (v_E k_\perp)^{-1}$. 
-This nonlinear time scale is generally associated with the eddy turnover or decorrelation time. We therefore present a heuristic model of the nonlinearities 
+First, we note that the advective nonlinearity terms in Eqs.~\ref{n_eq} and~\ref{phi_eq} have a common form -- a state variable multiplied by an inverse time scale:
+$\tau_{nl}^{-1} \sim v_E k_\perp$.
+This nonlinear time scale is generally associated with the eddy turnover or decorrelation time. We thus present a heuristic model of the nonlinearities 
 as a randomizing force that acts on this characteristic nonlinear time scale. More specifically, this model is one in which 1) the turbulence begins as a random state, 
@@
 
-This model can be used by starting with an ensemble of random initial conditions, evolving them linearly for a time $\tau_{nl}$, and
+We can calculate an effective growth rate from this model by starting with an ensemble of random initial conditions, evolving them linearly for a time $\tau_{nl}$, and
 then taking the time and ensemble averaged growth rate of these curves.
@@
 \eeq
-Since we want to take the $\mathbf{u}(0)$ in the ensemble as random normalized vectors with uncorrelated components, it follows that~\cite{camargo1998}
+Since we want the ensemble of initial conditions ($u(0)$'s) to be random normalized vectors with uncorrelated components, it follows that~\cite{camargo1998}
 
@@
 \eeq
-where the normalization is $||\mathbf{u}(0)||^2 = 1$. Mathematically, we define the non-modal growth rate $\gamma_{{\rm{nm}},k}$ as the time averaged instantaneous growth rate:
+where the normalization is $||u(0)||^2 = 1$. Mathematically, we define the non-modal growth rate $\gamma_{{\rm{nm}},k}$ as the time averaged instantaneous growth rate:
 
@@
 \eeq
-Next, we substitute the ensemble averaged energy calculated from Eq.~\ref{E_t_ensemble_avg} for $E_k(t)$, and set $E_k(0) = 1$ by our normalization. Then,
+Taking $E_k(t)$ as the ensemble averaged energy calculated from Eq.~\ref{E_t_ensemble_avg} and set $E_k(0) = 1$ for normalization, we find,
 
@@
 \eeq
-Now, in order to obtain a simple and quick prediction of $\gamma_{{\rm{nm}},k}$, we must estimate $\tau_{nl,k}$ without having to calculate it directly from a simulation. 
-We thus invoke the conjecture of \emph{critical balance}, which posits that the nonlinear time scale equals the linear time scale at all spatial scales~\cite{schekochihin2012}. This is a fair
-assumption in light of our discussion above, in which we derived the relation: $\gamma_{l,k} = - \gamma_{nl,k}$.
+Now, in order to obtain a simple and quick prediction of $\gamma_{{\rm{nm}},k}$, we must estimate $\tau_{nl,k}$ without having to calculate it directly from a turbulent simulation. 
+We thus invoke the conjecture of \emph{critical balance}, which posits that the nonlinear time scale equals a characteristic linear time scale at all spatial scales~\cite{schekochihin2012}. 
+This is a fair assumption in light of our discussion above, in which we showed that $\gamma_{l,k} = - \gamma_{nl,k}$.
 
@@
 model for $\alpha=10^{-5}$ along with the eigenmode and turbulent spectra ($\gamma_{s,k}$ and $\gamma_{t,k}$), which we already provided in Fig.~\ref{alpha1e-5_gamma_s_t}. We do the same for
-$\alpha=1$ in Fig.~\ref{alpha1_gamma_s_t}. For the highly non-normal case ($\alpha = 10^{-5}$), $\gamma_{\omega,k}$ reproduces $\gamma_{t,k}$ in both magnitude and spectral shape. $\gamma_{\gamma,k}$
-is not close in any respect to $\gamma_{t,k}$. On the other hand, for the nearly normal case ($\alpha=1$), $\gamma_{\omega,k}$ does not resemble $\gamma_{t,k}$. $\gamma_{\gamma,k}$ actually is
-fairly similar to $\gamma_{s,k}$, but with a slightly decreased amplitude, making it a closer match to $\gamma_{t,k}$, though it is still not similar for all $k$.
+$\alpha=1$ in Fig.~\ref{alpha1_gamma_spec_compare}. For the highly non-normal case ($\alpha = 10^{-5}$), $\gamma_{\omega,k}$ reproduces $\gamma_{t,k}$ in both magnitude and spectral shape. 
+$\gamma_{\gamma,k}$ is not close in any respect to $\gamma_{t,k}$. On the other hand, for the nearly normal case ($\alpha=1$), $\gamma_{\omega,k}$ does not resemble $\gamma_{t,k}$. 
+$\gamma_{\gamma,k}$ actually is fairly similar to $\gamma_{s,k}$, but with a slightly decreased amplitude, making it a closer match to $\gamma_{t,k}$, though it is still not similar for all $k$.
 
@@
 
-We further test how well the maxima of the non-modal growth rates predict the maxima of the turbulent growth rates. We show the result in Fig.~\ref{gamma_max_with_nm}. For $\alpha < 10^{-3}$,
+We further test how well the maxima of the non-modal growth rates predict the maxima of the turbulent growth rates. We show the result in Fig.~\ref{gamma_max_with_nm}. For $\alpha \le 10^{-4}$,
 in which the system is fairly non-normal, $\gamma_{\omega,k}$ predicts $\gamma_{t,k}$ quite accurately. As the system becomes more normal, $\gamma_{t,k}$ and $\gamma_{\omega,k}$ diverge, with
@@
 It is perhaps easy to be fooled into believing that the turbulence
-in such systems is due to eigenmode instability. In either case, one of the most important predictions is the minimum turbulent control parameter $R_g$.
+in such systems is due to eigenmode instability. In either case, one of the most important predictions is the minimum turbulent control parameter $R_g$. Given that this problem is many decades old
+and still unsolved, it is a tall order to predict $R_g$ without nonlinear simulations. We nevertheless make an attempt.
 
@@
 
+\begin{figure}
+\centerline{\includegraphics[width=0.4\textwidth]{kappa_en_ev}}
+\caption{}
+\label{kappa_en_ev}
+\end{figure}
+
 \begin{acknowledgments}

### 2014-04-18 HW_non-modal.tex
April 17

@@
 Nevertheless, it is well-known in the hydrodynamics community that normal mode analysis fails at predicting turbulent onset for a large class of experiments, in which the
-turbulence is called subcritical. In these experiments, turbulence appears even though all linear eigenvectors are stable, meaning infinitesimal perturbations on the laminar state cannot grow
-exponentially. Nevertheless, finite amplitude perturbations can still excite turbulence.
+turbulence is called subcritical~\cite{drazin1981}. 
+In these experiments, turbulence appears even though all linear eigenvectors are stable, meaning infinitesimal perturbations on the laminar state cannot grow exponentially. 
+Nevertheless, finite amplitude perturbations can still excite turbulence.
 Subcritical turbulence is certainly possible in plasmas as well, and several instances have been studied in the 
-past~\cite{waltz1985,scott1990,scott1992,nordman1993,biskamp1995,drake1995,camargo1998,krommes1999,scott2002,camporeale2009,camporeale2010,schekochihin2012,highcock2012}. 
+past~\cite{waltz1985,scott1990,nordman1993,biskamp1995,drake1995,itoh1996,camargo1998,krommes1999,camporeale2009,schekochihin2012,highcock2012}. 
 For the most part, however, normal mode analysis dominates plasma turbulence prediction despite the fact that it is not useful in all instances.
@@
 Since the hydrodynamics community has long studied subcritical turbulence, it has much to offer in terms of understanding of the phenomenon. Most significantly,
-in the early 1990s, Trefethen and others attributed the failure of normal mode analysis in predicting subcritical turbulent onset to the non-normality of linear operators of
-dynamical systems~\cite{trefethen1993,schmid2007}. A non-normal operator has 
+in the early 1990s, several researchers attributed the failure of normal mode analysis in predicting subcritical turbulent onset to the non-normality of linear operators of
+dynamical systems~\cite{gustavsson1991,butler1992,trefethen1993,reddy1993,henningson1994,schmid2007}. A non-normal operator has 
 eigenvectors that are not orthogonal to one another. One consequence of eigenvector nonorthogonality is that even when all eigenvectors decay exponentially under linear evolution, 
@@
 
-Non-modal analysis generally employs tools such as pseudospectra and maximum possible transient growth curves, which have been used to successfully explain why turbulence is subcritical in various
-systems. However, they have not been used to predict onset criteria in subcritical systems (e.g. the transition Reynold's number in pipe flows), although non-normal
-turbulent arguments have been applied in generating amplitude versus Reynold's number scalings for turbulent onset~\cite{grossmann2000}.
-This paper takes up the task of developing a new approach to understand turbulent properties using only non-modal linear calculations with the goal of 
-ultimately making quantitative predictions.  Our approach is to calculate an average growth rate of turbulent fluctuations due to linear processes, specifically due to transient growth.  We model the
-turbulent steady state as a series of processes:  (1) the turbulence starts as a spatially random state, (2) linear transient growth deterministically amplifies the turbulent energy (or
-decrease it in wavenumber ranges where linear damping dominates), (3) nonlinear transfer sets in at a specified timescale, terminating the
-transient growth process and re-randomizing the turbulent state (at which point the cycle repeats).    Optimally, the timescale for the final step would be the
-nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
-arguments to use a characteristic \emph{linear} time. The procedure ultimately produces a growth rate that can be used
-to predict turbulent properties such as saturation levels and transport rates through mixing length arguments.
+Non-modal analysis generally employs tools such as pseudospectra and maximum growth curves, which have been used to successfully explain why turbulence is subcritical in various instances. 
+However, they have not been used to predict onset criteria in subcritical systems (e.g. the transition Reynold's number in pipe flows).
+This paper outlines and illustrates a non-modal linear method to make quantitative predictions of turbulent properties including turbulent onset criteria. 
+Our approach is to define a non-modal growth rate that is analagous to an eigenmode growth rate, but which takes into account transient growth effects. To accomplish this, we make
+several simplifying assumptions regarding the nonlinearities; namely, the nonlinearities randomize the spatial structures of the turbulence on a characteristic nonlinear time scale, which can
+be estimated by a characteristic linear time scale.
+
+More specifically, we model the turbulent steady state as a series of processes:  (1) the turbulence starts as a spatially random state, 
+(2) linear transient growth deterministically amplifies the turbulent energy (or decrease it in wavenumber ranges where linear damping dominates), 
+(3) nonlinear transfer sets in at a specified timescale, terminating the transient growth process and re-randomizing the turbulent state (at which point the cycle repeats). 
+Optimally, the timescale for the final step would be the nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
+arguments to use a characteristic linear time. The procedure ultimately produces a growth rate that can be used to predict turbulent properties such as saturation levels and transport rates 
+through mixing length arguments. Furthermore, The sign of the growth rate predicts whether or not turbulence can exist -- a positive growth rate predicting turbulence.
 
@@
 dispersion relation -- the damped eigenmode. The importance of the stable eigenmode branch (or branches) 
-has started to gain acceptance in the plasma community, although it has generally been used in normal
-or nearly normal systems to explain turbulent saturation~\cite{terry2006b} ($\gamma_{t,k} < \gamma_{s,k}$). The case of highly non-normal turbulence and the associated increased turbulent excitation
+has started to gain acceptance in the plasma community~\cite{baver2002}, although it has generally been used in normal
+or nearly normal systems to explain turbulent saturation~\cite{terry2006b,hatch2011,makwana2011} ($\gamma_{t,k} < \gamma_{s,k}$). 
+The case of highly non-normal turbulence and the associated increased turbulent excitation
 hasn't been understood as well. In fact, simulation and analysis of the 3D HW model (and extended models) is an area in which others have seen the effects of strong non-normal transient growth

### 2014-05-08 HW_non-modal.tex
May 7

@@
 
-\title{}
+\title{A Procedure to Predict the Subcritical Turbulent Onset Criterion Applied to a Modified Hasegawa-Wakatani Model}
 
@@
 \section{Non-Normality of the Hasegawa-Wakatani Model}
+\label{sec_hw_model}
 
-Before employing the non-modal technique, we explore the effect of non-normality in the well-known and well-analyzed Hasegawa-Wakatani (HW) model~\cite{hasegawa1983}. 
+In order to illustrate and test our non-modal technique, we use as an example the well-known and well-analyzed Hasegawa-Wakatani (HW) model~\cite{hasegawa1983}.
+In this section, we analyze the linear normality of the model.
 We consider both the original 2D and the extended 3D~\cite{biskamp1995} versions of the model in a straight, unsheared magnetic field. 
@@
 and $\kappa = \rho_s/L_n$, where $L_n$ is the density scale length. The terms $D \gradperp^4 n$ and $D \gradperp^6 \phi$ are artificial hyper-diffusion and hyper-viscosity used to dissipate
-energy on small scales. We set $D \to 10^{-5}$ except when otherwise indicated.
+energy on small scales. We set $D \to 10^{-5}$ except in Sec.~\ref{sec_subcrit_prediction} where we set it to $0.5$.
 We associate the direction along $\mathbf{B}$ as the parallel or $z$ direction. The direction perpendicular $\mathbf{B}$ that
@@
 \section{The Effect of Non-Normality on Turbulence}
+\label{sec_non_norm_turb}
 
@@
 \centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_kz}}
-\caption{The growth rates of the fastest growing eigenmode $\gamma_{s,k}$ and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model. The most remarkable feature is
+\caption{The growth rates of the fastest growing eigenmode $\gamma_{s,k}$ and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model at $\xi = 10^3$.
+The most remarkable feature is
 that the turbulent growth rate peaks at $k_z=0$ despite the stability of all linear eigenmodes at $k_z=0$.}
@@
 
-The 3D HW model nonlinear simulation has a fascinating property -- the turbulent energy injection is strongest at $k_z = 0$ despite the fact that there can be no unstable linear eigenmodes at $k_z=0$.
+The 3D HW model nonlinear simulation has a fascinating property -- the turbulent energy injection is positive at $k_z = 0$ despite the fact that there can be no unstable linear eigenmodes at $k_z=0$.
+In fact, for high enough values of $\xi$ ($\xi \ge 10^2$), more energy is injected at $k_z = 0$ than at finite $k_z$ (see Fig.~\ref{gamma_max_vs_xi} below for more details).
 We show this in Fig.~\ref{gamma_max_vs_kz} using the eigenmode $\gamma_{s,k}$ and turbulent $\gamma_{t,k}$ growth rates, plotting them as a function of $k_z$.
@@
 
-\section{Non-Modal Linear Technique To Calculate Turbulent Growth Rate}
+\section{Non-Modal Linear Procedure To Calculate Turbulent Growth Rate}
+\label{sec_nm_procedure}
 
@@
 \caption{Comparison of the non-modal growth rates, $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$, to the growth rates of the fastest growing eigenmode $\gamma_{s,k}$ 
-and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model. The non-modal growth rates correctly predict the growth rate dominance of the $k_z=0$ mode
+and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model with $\xi = 10^3$. The non-modal growth rates correctly predict the growth rate dominance of the $k_z=0$ mode
 despite the $k_z=0$ eigenmode stability.}
@@
 
-Despite these complications at $k_z=0$, our non-modal method of calculating $\gamma_{\omega,k}$ approximates $\gamma_{t,k}$ remarkably well as seen in Fig.~\ref{gamma_max_vs_kz_with_nm}.
-Most importantly, it correctly predicts that $\gamma_{\omega,k_z=0} > 0$ and $\gamma_{\omega,k_z=0} > \gamma_{\omega,k_z \ne 0}$. Therefore, non-modal calculations explain why the 3D HW model
-is dominated by the nonlinear instability rather than a linear drift wave instability. Simply put, transient growth is strongest at $k_z=0$, so the nonlinear instability is energetically
+\begin{figure}
+\centerline{\includegraphics[width=0.55\textwidth]{gamma_max_vs_xi}}
+\caption{Comparison of the maximum non-modal growth rates, $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$, to the growth rates of the fastest growing eigenmode $\gamma_{s,k}$ 
+and the of the turbulence $\gamma_{t,k}$ as a function of $\xi$ for the 3D HW model at {\bf a)} $k_z \rho_s = 0$ and {\bf b)} $k_z \rho_s \approx 0.02$. 
+As in the 2D model, $\gamma_{\omega,k}$ does }
+\label{gamma_max_vs_xi}
+\end{figure}
+
+Despite these complications at $k_z=0$, our non-modal method of calculating $\gamma_{\omega,k}$ approximates $\gamma_{t,k}$ remarkably well as seen in Figs.~\ref{gamma_max_vs_kz_with_nm}
+and~\ref{gamma_max_vs_xi}.
+Most importantly, it correctly predicts that $\gamma_{\omega,k_z=0} > 0$ and $\gamma_{\omega,k_z=0} > \gamma_{\omega,k_z \ne 0}$ for $\xi \ge 10^{2}$. 
+Therefore, non-modal calculations explain why the 3D HW model is dominated by the nonlinear instability (for $\xi \ge 10^{2}$) rather than a linear drift wave instability. 
+Simply put, transient growth is strongest at $k_z=0$, so the nonlinear instability is energetically
 favored over the linear instability. Again, $\gamma_{\gamma,k}$ does not provide enough accuracy to warrant its use.
@@
 \section{Prediction of Control Parameter for Subcritical Turbulent Onset}
+\label{sec_subcrit_prediction}
 
-In both hydrodynamic and plasma systems, it is important to be able to predict the smallest value of a control parameter for which the system will become turbulent. 
-The control parameter is different for different systems (and models of those systems), and there may be several control parameters for a particular problem. 
-When $d$ control parameters exist in a model, 
-there is not a single point in the parameter space at which the system bifurcates to turbulence, rather, there is a $d-1$ manifold that separates the laminar and turbulent states.
-In many hydrodynamic systems driven turbulent by shear flow, there is generally a single control parameter -- the Reynold's number $Re$~\cite{drazin1981}.
-In plasmas, there are often multiple control parameters.
-For instance, kinetic solar wind turbulence models have two: $\beta$ and $T_\perp/T_\para$~\cite{camporeale2010}).
-One kinetic turbulence in the
-core of a tokamak has three: $\gamma_E$ (the normalized perpendicular flow shear), $R/L_T$ (the inverse ion temperature gradient scale length normalized to the major radius), and $q/\epsilon$
-(the ratio of the magnetic safety factor to the inverse aspect ratio)~\cite{highcock2012}.
-
-For normal systems, turbulence will be excited when the control parameter becomes large enough such that one of the linear eigenmodes becomes unstable (the real part of the eigenvalue crosses into the
-right half of the complex plane). In general, this is not the case for non-normal systems. In many non-normal systems, turbulence can be sustained when all linear eigenmodes are stable --
-the case of subcritical turbulence.
+In both hydrodynamic and plasma systems, it is important to be able to predict the control parameter space that divides the turbulent and laminar states.
+In many hydrodynamic systems with sheared flow, there is generally a single control parameter -- the Reynold's number $Re$~\cite{drazin1981}. For high enough $Re$ such systems
+can be turbulent, but for low $Re$ they are always laminar.
+Normal systems will undergo a bifurcation from a laminar to non-laminar state when the control parameters are such that at least one of the linear eigenmodes is unstable~\cite{grossmann2000}.
+In general, this is not true for non-normal systems. In many non-normal systems, turbulence can be sustained when all linear eigenmodes are stable -- the case of subcritical turbulence.
 
@@
 Subcritical systems fall into two categories: those that have stable eigenmodes for all values of the control parameter, and those that have unstable eigenmodes
-when the control parameter becomes large enough. We illustrate these two categories with a diagram in Fig.~\ref{subcritical_diagram}. 
+when the control parameter becomes large (or small) enough. We illustrate these two categories with a diagram in Fig.~\ref{subcritical_diagram}. 
 The upper curve represents the first kind of subcritical system in which no unstable eigenmodes exist for any value of $R$. In this
@@
 It is perhaps easy to be fooled into believing that the turbulence
-in such systems is due to eigenmode instability. In either case, one of the most important predictions is the minimum turbulent control parameter $R_g$. Given that this problem is many decades old
+in such systems is due to eigenmode instability. In either case, one of the most important predictions is the minimum turbulent control parameter $R_g$, which is sometimes called the ``transitional''
+vlue of the control parameter. Given that this problem is many decades old
 and still unsolved, it is a tall order to predict $R_g$ without nonlinear simulations. We nevertheless make an attempt.
@@
 Otherwise, the fluctuations grow transiently, but do not reach high enough amplitude for the nonlinearities to bootstrap the process,
-so they decay away. The finite amplitude turbulence threashold, as described above, is the key indicator of subcritical turbulence.
+so they decay away (we set the diffusion and viscosity coefficient $D$ to $0.5$ in this section so that the decay occurs on a short enough time scale to keep the simulations short). 
+The finite amplitude turbulence threashold, as described above, is a key indicator of subcritical turbulence.
 
@@
 
+In this subcritical system, we predict the line in $\xi-\kappa$ parameter space that divides the space into an area that permits turbulent solutions and an area that does not. Note that we are
+not concerend here with amplitude threasholds. We simply want to know for given values of $\xi$ and $\kappa$ whether turbuelence can exist given a sufficiently large perturbation. Our method
+of prediction is to find the line in $\xi-\kappa$ space at which $\gamma_\omega = 0$, where $\gamma_\omega$ is maximized over $k_x$ and $k_y$. We plot this line in Fig.~\ref{subcrit_param_space}
+
+\begin{figure}
+\centerline{\includegraphics[width=0.4\textwidth]{subcrit_param_space}}
+\caption{The non-modal curve $\gamma_\omega(\kappa,\xi) = 0$, which defines our prediction of the transitional value of $\kappa(\xi)$ for the subcritical 3D HW model. The points represent the
+transitional values of $\kappa(\xi)$ based on full nonlinear simulations.}
+\label{subcrit_param_space}
+\end{figure}
+
 \begin{figure}
 \centerline{\includegraphics[width=0.4\textwidth]{kappa_en_ev}}
-\caption{}
+\caption{$E_{\rm{tot}}(t)$ from full nonlinear simulations of the subcritical 3D HW model. The curves that reach a steady-state value
+reach sustained subcritically turbulent states. Those that decay for large time have values of $\kappa$ too small to sustain turbulence. The transitional value of $\kappa$ is the smallest value
+for which turbulence is sustained.}
 \label{kappa_en_ev}

### 2014-05-13 HW_non-modal.tex
May 12

@@
 \label{E_t_ensemble_avg}
-\left< E_k(t) \right>_{{\rm{ens}}} = \frac{1}{2} {\rm{tr}} \{ e^{\mathbf{B}_k t} e^{\mathbf{B}^{\dagger}t} \},
+\left< E_k(t) \right>_{{\rm{ens}}} = \frac{1}{2} {\rm{tr}} \{ e^{\mathbf{B}_k t} e^{\mathbf{B}_k^{\dagger}t} \},
 \eeq
@@
 
-There are two fairly obvious linear time scales in the HW problem. First is the inverse of the linear eigenmode frequency $\omega_{s,k}$
+There are a several possible linear time scales in the HW problem to use for $\tau_{nl,k}$ 
+including inverse linear eigenmode frequency and growth rates as well as certain times associated with transient growth curves. 
+After trying a few of these different time scales, we have concluded that the one that works best is the inverse of the linear eigenmode frequency $\omega_{s,k}$
 -- the imaginary part of the eigenvalues of the linear matrix operator $\mathbf{B}_k$.
-Recall that the imaginary part of the linear eigenvalues is the same for both eigenvalues of $\mathbf{B}_k$, so there is no question of which to use. We thus define the first non-modal
-growth rate $\gamma_{\omega,k}$ as
+Recall that the imaginary part of the linear eigenvalues is the same for both eigenvalues of $\mathbf{B}_k$, so there is no question of which to use.
+Because of this, unlike the eigenmode growth rates,
+the frequency is robust no matter how nonorthogonal the eigenvectors. We thus arrive at our final definition for the non-modal growth rate:
 
@@
 \label{gamma_omega_def}
-\gamma_{\omega,k} = \frac{\omega_{s,k}}{2} \ {\rm{Log}} \left< E_k(1/\omega_{s,k}) \right>_{\rm{ens}}.
+\gamma_{\rm{nm},k} = \frac{\omega_{s,k}}{4} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k t} e^{\mathbf{B}^{\dagger}t} \} \right].
 \eeq
-This growth rate is calculated by first finding $\omega_{s,k}$ of $\mathbf{B}_k$, and then inserting this into Eq.~\ref{gamma_omega_def} using the definition of 
-$\left< E_k(1/\omega_{s,k}) \right>_{\rm{ens}}$ from Eq.~\ref{E_t_ensemble_avg}.
+Note that the factor of $\frac{1}{4}$ is particular to two-field systems. Namely, the factor of $\frac{1}{2}$ coming from Eq.~\ref{E_t_ensemble_avg} is a consequence of the $2 \times 2$ size
+of $\mathbf{B}_k$.
 
-The second linear time scale that we may use is the inverse of the linear growth rate. However, the eigenvalues of $\mathbf{B}_k$ have different growth rates (real parts). 
-Furthermore, we have argued that the eigenmode growth rates have little bearing on turbulence, especially for highly non-normal turbulence, 
-so it would be innappropriate to use either of the linear eigenmode growth rates as a proxy for the nonlinear time scale.
-Rather, we use the result itself, $\gamma_{{\rm{nm}},k}$, as the inverse of $\tau_{nl,k}$. Substituting $1/|\gamma_{{\rm{nm}},k}|$ for $\tau_{nl,k}$ in Eq.~\ref{gamma_nm_calc}, we get
-a transcendental equation:
-
-\beq
-\label{trans_gamma_nm}
-\gamma_{{\rm{nm}},k} = \frac{|\gamma_{{\rm{nm}},k}|}{2} \ {\rm{Log}} \left< E_k(1/|\gamma_{{\rm{nm}},k}|) \right>_{\rm{ens}}.
-\eeq
-We label this second non-modal growth rate $\gamma_{\gamma,k}$, and ultimately simplify this equation to
-
-\begin{figure}
-\centerline{\includegraphics[width=0.52\textwidth]{alpha1e-5_gamma_spec_compare}}
-\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$,  b) the turbulent growth rate spectrum $\gamma_{t,k}$, c) the non-modal growth rate spectrum $\gamma_{\omega,k}$ d) and the 
-non-modal growth rate spectrum $\gamma_{\gamma,k}$ for the 2D HW model with $\alpha = 10^{-5}, \kappa=1$. Note the different scales.}
-\label{alpha1e-5_gamma_spec_compare}
-\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[width=0.52\textwidth]{alpha1_gamma_spec_compare}}
-\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$,  b) the turbulent growth rate spectrum $\gamma_{t,k}$, c) the non-modal growth rate spectrum $\gamma_{\omega,k}$ d) and the 
-non-modal growth rate spectrum $\gamma_{\gamma,k}$ for the 2D HW model with $\alpha = 1, \kappa=1$.}
-\label{alpha1_gamma_spec_compare}
-\end{figure}
-
-\beq
-\label{gamma_pm2}
-\left< E_k(1/|\gamma_{\gamma,k}|) \right>_{\rm{ens}} = e^{\pm 2},
-\eeq
-where if $\left< E_k(t) \right>_{\rm{ens}}$ first reaches $e^2$, $\gamma_{\gamma,k}$ is positive, and if it first reaches $e^{-2}$, $\gamma_{\gamma,k}$ is negative.
-Therefore, finding $\gamma_{\gamma,k}$ amounts to a root-finding problem, which is not trivial, but is much simpler than a nonlinear simulation.
-
-We test both of these non-modal growth rates to see if either or both produce good approximations of $\gamma_{t,k}$, especially in comparison to $\gamma_{s,k}$. 
-In Fig.~\ref{alpha1e-5_gamma_spec_compare}, we show the $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$ spectra for the 2D HW
-model for $\alpha=10^{-5}$ along with the eigenmode and turbulent spectra ($\gamma_{s,k}$ and $\gamma_{t,k}$), which we already provided in Fig.~\ref{alpha1e-5_gamma_s_t}. We do the same for
-$\alpha=1$ in Fig.~\ref{alpha1_gamma_spec_compare}. For the highly non-normal case ($\alpha = 10^{-5}$), $\gamma_{\omega,k}$ reproduces $\gamma_{t,k}$ in both magnitude and spectral shape. 
-$\gamma_{\gamma,k}$ is not close in any respect to $\gamma_{t,k}$. On the other hand, for the nearly normal case ($\alpha=1$), $\gamma_{\omega,k}$ does not resemble $\gamma_{t,k}$. 
-$\gamma_{\gamma,k}$ actually is fairly similar to $\gamma_{s,k}$, but with a slightly decreased amplitude, making it a closer match to $\gamma_{t,k}$, though it is still not similar for all $k$.
+%The second linear time scale that we may use is the inverse of the linear growth rate. However, the eigenvalues of $\mathbf{B}_k$ have different growth rates (real parts). 
+%Furthermore, we have argued that the eigenmode growth rates have little bearing on turbulence, especially for highly non-normal turbulence, 
+%so it would be innappropriate to use either of the linear eigenmode growth rates as a proxy for the nonlinear time scale.
+%Rather, we use the result itself, $\gamma_{{\rm{nm}},k}$, as the inverse of $\tau_{nl,k}$. Substituting $1/|\gamma_{{\rm{nm}},k}|$ for $\tau_{nl,k}$ in Eq.~\ref{gamma_nm_calc}, we get
+%a transcendental equation:
+%
+%\beq
+%\label{trans_gamma_nm}
+%\gamma_{{\rm{nm}},k} = \frac{|\gamma_{{\rm{nm}},k}|}{2} \ {\rm{Log}} \left< E_k(1/|\gamma_{{\rm{nm}},k}|) \right>_{\rm{ens}}.
+%\eeq
+%We label this second non-modal growth rate $\gamma_{\gamma,k}$, and ultimately simplify this equation to
+%
+%\begin{figure}
+%\centerline{\includegraphics[width=0.52\textwidth]{alpha1e-5_gamma_spec_compare}}
+%\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$,  b) the turbulent growth rate spectrum $\gamma_{t,k}$, c) the non-modal growth rate spectrum $\gamma_{\omega,k}$ d) and the 
+%non-modal growth rate spectrum $\gamma_{\gamma,k}$ for the 2D HW model with $\alpha = 10^{-5}, \kappa=1$. Note the different scales.}
+%\label{alpha1e-5_gamma_spec_compare}
+%\end{figure}
+%
+%\begin{figure}
+%\centerline{\includegraphics[width=0.52\textwidth]{alpha1_gamma_spec_compare}}
+%\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$,  b) the turbulent growth rate spectrum $\gamma_{t,k}$, c) the non-modal growth rate spectrum $\gamma_{\omega,k}$ d) and the 
+%non-modal growth rate spectrum $\gamma_{\gamma,k}$ for the 2D HW model with $\alpha = 1, \kappa=1$.}
+%\label{alpha1_gamma_spec_compare}
+%\end{figure}
+%
+%
+%\beq
+%\label{gamma_pm2}
+%\left< E_k(1/|\gamma_{\gamma,k}|) \right>_{\rm{ens}} = e^{\pm 2},
+%\eeq
+%where if $\left< E_k(t) \right>_{\rm{ens}}$ first reaches $e^2$, $\gamma_{\gamma,k}$ is positive, and if it first reaches $e^{-2}$, $\gamma_{\gamma,k}$ is negative.
+%Therefore, finding $\gamma_{\gamma,k}$ amounts to a root-finding problem, which is not trivial, but is much simpler than a nonlinear simulation. 
+%
+%We test both of these non-modal growth rates to see if either or both produce good approximations of $\gamma_{t,k}$, especially in comparison to $\gamma_{s,k}$. 
+%In Fig.~\ref{alpha1e-5_gamma_spec_compare}, we show the $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$ spectra for the 2D HW
+%model for $\alpha=10^{-5}$ along with the eigenmode and turbulent spectra ($\gamma_{s,k}$ and $\gamma_{t,k}$), which we already provided in Fig.~\ref{alpha1e-5_gamma_s_t}. We do the same for
+%$\alpha=1$ in Fig.~\ref{alpha1_gamma_spec_compare}. For the highly non-normal case ($\alpha = 10^{-5}$), $\gamma_{\omega,k}$ reproduces $\gamma_{t,k}$ in both magnitude and spectral shape. 
+%$\gamma_{\gamma,k}$ is not close in any respect to $\gamma_{t,k}$. On the other hand, for the nearly normal case ($\alpha=1$), $\gamma_{\omega,k}$ does not resemble $\gamma_{t,k}$. 
+%$\gamma_{\gamma,k}$ actually is fairly similar to $\gamma_{s,k}$, but with a slightly decreased amplitude, making it a closer match to $\gamma_{t,k}$, though it is still not similar for all $k$.
 
@@
 \centerline{\includegraphics[width=0.45\textwidth]{gamma_max_with_nm}}
-\caption{Comparison of the non-modal growth rates, $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$, to the eigenmode $\gamma_{s}$ and turbulent $\gamma_{t}$ growth rates as a function of $\alpha$
+\caption{Comparison of the non-modal growth rate $\gamma_{\rm{nm},k}$ to the eigenmode $\gamma_{s}$ and turbulent $\gamma_{t}$ growth rates as a function of $\alpha$
 for the 2D HW model. All growth rates are peaks in $k_x-k_y$ space.}
@@
 
-We further test how well the maxima of the non-modal growth rates predict the maxima of the turbulent growth rates. We show the result in Fig.~\ref{gamma_max_with_nm}. For $\alpha \le 10^{-4}$,
-in which the system is fairly non-normal, $\gamma_{\omega,k}$ predicts $\gamma_{t,k}$ quite accurately. As the system becomes more normal, $\gamma_{t,k}$ and $\gamma_{\omega,k}$ diverge, with
-$\gamma_{t,k}$ taking on values between $\gamma_{s,k}$ and $\gamma_{\omega,k}$, but generally closer to $\gamma_{s,k}$ for $\alpha \gg 10^{-3}$. 
+The non-modal growth rate prediction agrees fairly well with the turbulent growth rate for highly non-normal 2D HW turbulence as seen in Fig.~\ref{gamma_max_with_nm}. For $\alpha \le 10^{-4}$,
+in which the system is fairly non-normal, $\gamma_{\rm{nm},k}$ predicts $\gamma_{t,k}$ quite accurately, especially compared to the eigenmode prediction. 
+As the system becomes more normal, $\gamma_{t,k}$ and $\gamma_{\rm{nm},k}$ diverge, with $\gamma_{t,k}$ taking on values between $\gamma_{s,k}$ and $\gamma_{\rm{nm},k}$, 
+but generally closer to $\gamma_{s,k}$ for $\alpha \gg 10^{-3}$. 
 This indicates that some assumption in our procedure doesn't work well for systems that are normal or nearly normal. Namely, our assumption
-of complete randomization by the nonlinearities breaks down. Apparently, the nonlinearities only partially randomize the turbulence. For the most part, the turbulence at a given $k$ consists
-primarily of the unstable eigenmode, while the stable eigenmode accounts for only a small amount of energy in the turbulence. Finally, $\gamma_{\gamma,k}$ is not useful in predicting $\gamma_{t,k}$.
-It is closest to $\gamma_{t,k}$ for $\alpha \ge 1$, however, it only provides a slight improvement over $\gamma_{s,k}$ in this range.
+of complete randomization by the nonlinearities breaks down. Apparently, the nonlinearities only partially randomize the turbulence in the weakly non-normal case. 
+For the most part, the turbulence at a given $k$ consists primarily of the unstable eigenmode, while the stable eigenmode accounts for only a small amount of energy in the turbulence.
 
@@
 \centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_kz_with_nm}}
-\caption{Comparison of the non-modal growth rates, $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$, to the growth rates of the fastest growing eigenmode $\gamma_{s,k}$ 
-and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model with $\xi = 10^3$. The non-modal growth rates correctly predict the growth rate dominance of the $k_z=0$ mode
+\caption{Comparison of the non-modal growth rate $\gamma_{\rm{nm},k}$ to the growth rates of the fastest growing eigenmode $\gamma_{s,k}$ 
+and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model with $\xi = 10^3$. The non-modal growth rate correctly predict the growth rate dominance of the $k_z=0$ mode
 despite the $k_z=0$ eigenmode stability.}
@@
 We do the same comparison for the 3D HW model as we did for the 2D HW model. However, there is one additional subtelty when considering the 3D model. In the 3D model, for $k_z=0$, the linear
-eigenmode frequency is zero ($\omega_{s,k_z=0} = 0$), so $\gamma_{\omega,k_z=0}$ is not well defined (see Eq.~\ref{gamma_omega_def}). 
-We could, in fact, make the argument that since $\omega_{s,k_z=0} = 0$, or by critical balance $\tau_{nl,k_z=0} \to \infty$, it would be proper to equate $\gamma_{\omega,k_z=0}$ with $\gamma_{s,k_z=0}$
+eigenmode frequency is zero ($\omega_{s,k_z=0} = 0$), so $\gamma_{\rm{nm},k_z=0}$ is not well defined (see Eq.~\ref{gamma_omega_def}). 
+We could, in fact, make the argument that since $\omega_{s,k_z=0} = 0$, or by critical balance $\tau_{nl,k_z=0} \to \infty$, it would be proper to equate $\gamma_{\rm{nm},k_z=0}$ with $\gamma_{s,k_z=0}$
 since all transients die away in the long time limit. Rather, we use $\omega_{s,k_z=0.02}$ -- where $k_z=0.02$ is the lowest finite value of $k_z$ available to our system -- in place of
-$\omega_{s,k_z=0}$ in the calculation of $\gamma_{\omega,k_z=0}$. We contend that the nonlinear instability, which most strongly couples $k_z=0$ and $k_z=0.02$ modes, has its time scale set
-by the drift waves, not the convective cells, making $1/\omega_{s,k_z=0.02}$ a good proxy for $\tau_{nl,k_z=0}$. One may recall that the drift waves are actually secondary, and therefore, they feed
-off of the $k_z=0$ density structures, so the appropriate scale length to use for $\omega_{s,k_z=0.02}$ is not given by $\kappa$, but by the scale length of the density structures. We don't deal
-with this complication and assume the scale length of the fluctuating density structures is the same as the background.
+$\omega_{s,k_z=0}$ in the calculation of $\gamma_{\rm{nm},k_z=0}$. We contend that the nonlinear instability, which most strongly couples $k_z=0$ and $k_z=0.02$ modes, has its time scale set
+by the drift waves, not the convective cells, making $1/\omega_{s,k_z=0.02}$ a good proxy for $\tau_{nl,k_z=0}$.
+
+%One may recall that the drift waves are actually secondary, and therefore, they feed
+%off of the $k_z=0$ density structures, so the appropriate scale length to use for $\omega_{s,k_z=0.02}$ is not given by $\kappa$, but by the scale length of the density structures. We don't deal
+%with this complication and assume the scale length of the fluctuating density structures is the same as the background.
 
@@
 \centerline{\includegraphics[width=0.55\textwidth]{gamma_max_vs_xi}}
-\caption{Comparison of the maximum non-modal growth rates, $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$, to the growth rates of the fastest growing eigenmode $\gamma_{s,k}$ 
-and the of the turbulence $\gamma_{t,k}$ as a function of $\xi$ for the 3D HW model at {\bf a)} $k_z \rho_s = 0$ and {\bf b)} $k_z \rho_s \approx 0.02$. 
-As in the 2D model, $\gamma_{\omega,k}$ does }
+\caption{Comparison of the maximum non-modal growth rate $\gamma_{\rm{nm},k}$ to the growth rates of the fastest growing eigenmode $\gamma_{s,k}$ 
+and the of the turbulence $\gamma_{t,k}$ as a function of $\xi$ for the 3D HW model at {\bf a)} $k_z \rho_s = 0$ and {\bf b)} $k_z \rho_s \approx 0.02$. }
 \label{gamma_max_vs_xi}
@@
 
-Despite these complications at $k_z=0$, our non-modal method of calculating $\gamma_{\omega,k}$ approximates $\gamma_{t,k}$ remarkably well as seen in Figs.~\ref{gamma_max_vs_kz_with_nm}
-and~\ref{gamma_max_vs_xi}.
-Most importantly, it correctly predicts that $\gamma_{\omega,k_z=0} > 0$ and $\gamma_{\omega,k_z=0} > \gamma_{\omega,k_z \ne 0}$ for $\xi \ge 10^{2}$. 
-Therefore, non-modal calculations explain why the 3D HW model is dominated by the nonlinear instability (for $\xi \ge 10^{2}$) rather than a linear drift wave instability. 
-Simply put, transient growth is strongest at $k_z=0$, so the nonlinear instability is energetically
-favored over the linear instability. Again, $\gamma_{\gamma,k}$ does not provide enough accuracy to warrant its use.
+Despite these complications at $k_z=0$, our non-modal method of calculating $\gamma_{\rm{nm},k}$ approximates $\gamma_{t,k}$ remarkably well as seen in Figs.~\ref{gamma_max_vs_kz_with_nm}
+and~\ref{gamma_max_vs_xi}. Most importantly, it correctly predicts that $\gamma_{\rm{nm},k_z=0} > 0$. 
+Therefore, non-modal calculations explain why the 3D HW model can be dominated by the nonlinear instability rather than a linear drift wave instability. 
+Simply put, the transient growth rate is by far the largest at $k_z=0$ for $\xi \gg 10^2$, so the nonlinear instability is energetically favored over the primary drift wave instability.
+For $\xi \le 10^2$, the $k_z=0$ nonlinear instability has comparable growth rate to the drift wave instability, so neither dominates.
 
@@
 not concerend here with amplitude threasholds. We simply want to know for given values of $\xi$ and $\kappa$ whether turbuelence can exist given a sufficiently large perturbation. Our method
-of prediction is to find the line in $\xi-\kappa$ space at which $\gamma_\omega = 0$, where $\gamma_\omega$ is maximized over $k_x$ and $k_y$. We plot this line in Fig.~\ref{subcrit_param_space}
+of prediction is to find the line in $\xi-\kappa$ space at which $\gamma_{\rm{nm}} = 0$, where $\gamma_{\rm{nm}}$ is maximized over $k_x$ and $k_y$. We plot this line in Fig.~\ref{subcrit_param_space}
 
@@
 \centerline{\includegraphics[width=0.4\textwidth]{subcrit_param_space}}
-\caption{The non-modal curve $\gamma_\omega(\kappa,\xi) = 0$, which defines our prediction of the transitional value of $\kappa(\xi)$ for the subcritical 3D HW model. The points represent the
+\caption{The non-modal curve $\gamma_{\rm{nm}}(\kappa,\xi) = 0$, which defines our prediction of the transitional value of $\kappa(\xi)$ for the subcritical 3D HW model. The points represent the
 transitional values of $\kappa(\xi)$ based on full nonlinear simulations.}

### 2014-06-19 HW_non-modal.tex
June 19

@@
 %\documentclass[showpacs,preprintnumbers,amsmath,amssymb,superscriptaddress,aip]{revtex4-1}
-\documentclass[letter,scriptaddress,twocolumn, prl,showkeys]{revtex4}
+\documentclass[twocolumn,showkeys,superscriptaddress]{revtex4}
 \usepackage{graphicx}
+%\usepackage{authblk}
 %\usepackage{amssymb}
@@
 
+\affiliation{Department of Physics and Astronomy, University of California, Los Angeles, California 90095-1547, USA}
+\affiliation{Lawrence Livermore National Laboratory, Livermore, California 94550, USA}
+
 \author{T.A. Carter}
@@
 More specifically, we model the turbulent steady state as a series of processes:  (1) the turbulence starts as a spatially random state, 
-(2) linear transient growth deterministically amplifies the turbulent energy (or decrease it in wavenumber ranges where linear damping dominates), 
+(2) linear transient growth deterministically amplifies the turbulent energy (or depresses it, depending on the wavenumber), 
 (3) nonlinear transfer sets in at a specified timescale, terminating the transient growth process and re-randomizing the turbulent state (at which point the cycle repeats). 
@@
 arguments to use a characteristic linear time. The procedure ultimately produces a growth rate that can be used to predict turbulent properties such as saturation levels and transport rates 
-through mixing length arguments. Furthermore, The sign of the growth rate predicts whether or not turbulence can exist -- a positive growth rate predicting turbulence.
+through mixing length arguments. Furthermore, a finite positive growth rate (of a certain magnitude) indicates that turbulence can exist given a sufficient amount of noise in the system.
+Thus, the growth rate provides a turbulent onset criterion, even for subcritical systems.
+
+The paper is organized as follows: in Section~\ref{sec_hw_model}, we specify our illustrative model -- the 2D and 3D Hasegawa-Wakatani model -- 
+that we use throughout the paper and describe its linear non-normal properties,  while in Section~\ref{sec_non_norm_turb}, we explore the effect of linear non-normality in the model on turbulence. 
+In Section~\ref{sec_nm_procedure}, we cover the details of our non-modal predictive procedure and show its success in predicting properties of highly non-normal turbulence. 
+Finally, in Section~\ref{sec_subcrit_prediction}, we modify the 3D Hasegawa-Wakatani model to make it subcritical and show how our non-modal technique predicts the critical gradient for the onset of turbulence.
 
@@
 the $y$ and $z$ directions with zero-value boundaries in the $x$ direction. We use a domain size of $64 \rho_s \times 64 \rho_s \times 300 \rho_s$.
-For the 2D version of the model, we replace $-\xi \gradpar^2$ by $\alpha (= k_z^2 \xi = 4 \pi k_z^2 \rho_s^2 \omega_{ce} /\nu_{ei})$, 
+For the 2D version of the model, we replace $-\xi \gradpar^2$ with a single parameter $\alpha$ (definied as $\alpha = 4 \pi k_z^2 \rho_s^2 \omega_{ce} /\nu_{ei})$, 
 which is commonly known as the adiabaticity parameter~\cite{camargo1995,camargo1998}. Additionally we use a domain size of $50 \rho_s \times 50 \rho_s$.
@@
 \label{B_norm_limit0}
-\displaystyle\lim_{\xi k_z^2 \to 0} \mathbf{B}_k = \left( \begin{array}{cc} - D k_\perp^4 & -i \kappa \frac{k_y}{k_\perp} \\ 0  & - D k_\perp^4\end{array} \right)
+\displaystyle\lim_{\xi k_z^2 \to 0} \mathbf{B}_k = \left( \begin{array}{cc} - D k_\perp^4 & -i \kappa \frac{k_y}{k_\perp} \\ 0  & - D k_\perp^4\end{array} \right).
 \eeq
-in which case $\mathbf{B}_k$ is non-normal. In general, as the adiabaticity parameter is varied, the normality of $\mathbf{B}_k$ changes, but unfortunatley there is no absolute metric to quantify
+Thus, $\mathbf{B}_k$ is non-normal. In general, as the adiabaticity parameter is varied, the normality of $\mathbf{B}_k$ changes, but unfortunatley there is no absolute metric to quantify
 the normality of a matrix or operator~\cite{trefethen2005}. In practice, however, what matters is not how non-normal $\mathbf{B}_k$ is, but how much transient energy growth is possible in the linear
@@
 \eeq
-where $u_k = \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right)$. The solution to this $u_k(t) = e^{\mathbf{B}_k t} u(0)$ where the exponential of the matrix is defined in terms of
+where $u_k = \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right)$. The solution to this is $u_k(t) = e^{\mathbf{B}_k t} u(0)$ where the exponential of the matrix is defined in terms of
 its Taylor expansion. The solution is dependent on the initial condition $u(0)$. The energy growth as a function of time is
@@
 where the operator $|| \cdot ||$ is the $L_2$-norm. Different initial conditions provide optimal energy growth at different times, but the upper envelope of all such optimal growth curves is given by
-$G_{k,\rm{max}}(t) = ||e^{\mathbf{B}_k t}||^2$, which does not depend on the initial condition.
+$G_{k,\rm{max}}(t) = ||e^{\mathbf{B}_k t}||^2$, which does not depend on an initial condition.
 Moreover, $||e^{\mathbf{B}_k t}||^2 \ge e^{2 \gamma_{s,k} t}$, where $\gamma_{s,k}$ is the real part of
@@
 
-One explanation for this behavior is that non-normal matrices or operators have non-orthogonal linear eigenvectors. For the HW model, each Fourier matrix $\mathbf{B}_k$ has two eigenvectors associated
+Non-normality manifests itself in its eigenvectors -- non-normal matrices or operators have non-orthogonal linear eigenvectors.
+For the HW model, each Fourier matrix $\mathbf{B}_k$ has two eigenvectors associated
 with it. They are complex and take the form:
@@
 
-Linear non-normality is important for the effect it has on turbulence. It is most often cited as a necessary condition for turbulence in subcritical systems. 
-But it can also impact systems that have unstable linear eigenmodes like the HW system. Since certain superpositions of the linear eigenmodes can inject energy faster than the most unstable linear
+Linear non-normality is important for the effect it has on turbulence. It is most significantly a necessary condition for turbulence in subcritical systems that have only conservative nonlinearities 
+(e.g. advective nonlinearities). However, non-normality can also impact non-subcritical systems, which have unstable linear eigenmodes like the HW system. 
+Since certain superpositions of the linear eigenmodes can inject energy faster than the most unstable linear
 eigenmode, there is the possibility that the energy injection into the turbulence is faster and at different wavenumbers than is suggested by linear eigenmode analysis.
-To understand this concept and show that it occurs in the nonlinear HW model, we compare the fastest growing linear eigenmode growth rate $\gamma_{s,k}$ to the turbulent growth rate $\gamma_{t,k}$. 
-To define the turbulent growth rate in general and specifically for the HW model, we employ a Fourier-decomposed energy dynamics analysis~\cite{camargo1995,friedman2012b,friedman2013}. 
-To do this, we take Eqs.~\ref{n_eq} and~\ref{phi_eq} and substitute the following:
+
+To illustrate this in the nonlinear HW model, we compare the fastest growing linear eigenmode growth rate $\gamma_{s,k}$ to the turbulent growth rate $\gamma_{t,k}$. 
+In order to define the turbulent growth rate in general and specifically for the HW model, we employ a Fourier-decomposed energy dynamics analysis~\cite{camargo1995,friedman2012b,friedman2013}. 
+Specifically, we take Eqs.~\ref{n_eq} and~\ref{phi_eq} and substitute the following:
 
@@
 $\diff{E_{l,k}}{t}$ represents the injection of energy into the fluctuations from the free energy in the equilibrium gradients plus the dissipation from collisions and hyperdiffusion.
-$\diff{E_{nl,k}}{t}$ represents $\sum_{k'} T(k,k')$ in  Eq.~\ref{Explicit_En_eqn} and comes from the nonlinear terms in 
-Eqs.~\ref{n_eq} and~\ref{phi_eq}. The terms in $T(k,k')$ are triadic in the fluctuating quantities and
-account for the energy exchange between fluctuations with different $k$'s. Furthermore, it is conservative: $\sum_{k} \diff{E_{nl,k}}{t} = \sum_{k,k'} T(k,k') = 0$.
+$\diff{E_{nl,k}}{t}$ represents $\sum_{k'} T(k,k')$ in  Eq.~\ref{Explicit_En_eqn} and comes from the nonlinear terms in Eqs.~\ref{n_eq} and~\ref{phi_eq}. 
+The terms in $T(k,k')$ are triadic in the fluctuating quantities and account for the energy exchange between fluctuations with different $k$'s. 
+They are also conservative when summed over all wavenumbers: $\sum_{k} \diff{E_{nl,k}}{t} = \sum_{k,k'} T(k,k') = 0$~\cite{camargo1995}.
 
@@
 \gamma_k \equiv  \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_k/dt}{2 E_k} dt \nonumber \\
-= \displaystyle\lim_{T \to \infty} \frac{{\rm{Log}}\left[ E_k(T)/E_k(0) \right]}{2 T} = 0
+= \displaystyle\lim_{T \to \infty} \frac{{\rm{Log}}\left[ E_k(T)/E_k(0) \right]}{2 T} = 0.
 \eeqar
-where the equality to zero clearly holds as long as the turbulence is in a steady or quasi-steady state.
+The limit of the last expression vanishes -- as indicated -- for steady or quasi-steady state turbulence only.
 From, Eqs.~\ref{dEdt_def} and~\ref{steady_state}, it follows that $ \gamma_k = \gamma_{l,k} + \gamma_{nl,k} = 0$.
@@
 A turbulent system, which can be decomposed into a linear eigenvector basis, has $\gamma_{t,k} \le \gamma_{s,k}$ if the system is normal. Recall that in normal systems, the maximal
-energy growth is given by the most unstable linear eigenmode. In other words, maximal growth is achieved when the \emph{most stable} eigenmode has zero amplitude.
-On the other hand, non-normal systems may grow faster when the more stable eigenmode has finite amplitude than when it has none (see Fig.~\ref{max_en_growth}). 
+energy growth is given by the most unstable linear eigenmode. In other words, maximal growth is achieved when only the most unstable eigenmode has finite amplitude.
+On the other hand, non-normal systems may grow faster when the more stable eigenmodes have finite amplitude than when they have none (see Fig.~\ref{max_en_growth}). 
 It is therefore possible for $\gamma_{t,k} > \gamma_{s,k}$ in non-normal systems.
@@
 \centerline{\includegraphics[width=0.52\textwidth]{alpha1e-5_gamma_s_t}}
-\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$ and b) the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 10^{-5}, \kappa=1$. Note the different scales.}
+\caption{{\bf a)} The eigenmode growth rate spectrum $\gamma_{s,k}$ and {\bf b)} the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 10^{-5}, \kappa=1$. Note the different scales.}
 \label{alpha1e-5_gamma_s_t}
@@
 \centerline{\includegraphics[width=0.52\textwidth]{alpha1_gamma_s_t}}
-\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$ and b) the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 1, \kappa=1$.}
+\caption{{\bf a)} The eigenmode growth rate spectrum $\gamma_{s,k}$ and {\bf b)} the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 1, \kappa=1$.}
 \label{alpha1_gamma_s_t}
@@
 or nearly normal systems to explain turbulent saturation~\cite{terry2006b,hatch2011,makwana2011} ($\gamma_{t,k} < \gamma_{s,k}$). 
-The case of highly non-normal turbulence and the associated increased turbulent excitation
-hasn't been understood as well. In fact, simulation and analysis of the 3D HW model (and extended models) is an area in which others have seen the effects of strong non-normal transient growth
-but have not realized them as such.
+The case of highly non-normal turbulence and the associated increased turbulent excitation hasn't been understood as well. 
+In fact, simulation and analysis of the 3D HW model (and extended models) is an area in which others have seen the effects of strong non-normal transient growth
+but have not realized them as such~\cite{biskamp1995,drake1995,scott2002,scott2005,umansky2009,friedman2012b}.
 
@@
 
-The 3D HW model nonlinear simulation has a fascinating property -- the turbulent energy injection is positive at $k_z = 0$ despite the fact that there can be no unstable linear eigenmodes at $k_z=0$.
+As Biaskamp and Zeiler~\cite{biskamp1995} discovered, the 3D HW model nonlinear simulation has a fascinating property -- 
+the turbulent energy injection is positive at $k_z = 0$ despite the fact that there can be no unstable linear eigenmodes at $k_z=0$.
 In fact, for high enough values of $\xi$ ($\xi \ge 10^2$), more energy is injected at $k_z = 0$ than at finite $k_z$ (see Fig.~\ref{gamma_max_vs_xi} below for more details).
 We show this in Fig.~\ref{gamma_max_vs_kz} using the eigenmode $\gamma_{s,k}$ and turbulent $\gamma_{t,k}$ growth rates, plotting them as a function of $k_z$.
-The long-stated cause of this $k_z=0$ energy injection is a nonlinear instability, which was discovered by Biaskamp and Zeiler~\cite{biskamp1995} and explored further by Drake et al.~\cite{drake1995}.
+The long-stated cause of this $k_z=0$ energy injection is a nonlinear instability~\cite{biskamp1995,drake1995}. 
 The nonlinear instability works as follows: magnetic-field-aligned ($k_z=0$) convective cells transport density across the equilibrium density gradient, setting up $k_z=0$ density structures. 
@@
 
-Although the instability is a nonlinear instability, the first part of the mechanism -- the transport of background density by the convective cells -- is a linear one.
-Biskamp and Drake understood this and wondered how there could be linear energy injection despite the fact that the linear eigenvectors are stable.
-The answer to this came when Camargo et al.~\cite{camargo1998} stated that ``some properties of the drift-wave turbulence
-previously attributed to nonlinear terms could also be strongly influenced by the nonnormal character of the linear system.'' Although Camargo et al. focused only on the 2D HW model,
-based on this statement, it was likely apparent to them that the 3D HW nonlinear instability was a result of linear non-normality. 
-Therefore, when $k_z=0$, no matter the value of $\xi$, the adiabaticity parameter $\alpha = k_z^2 \xi = 0$, meaning the linear matrix operator is highly non-normal and subject to transient growth.
+Although the instability is a nonlinear instability, the first part of the mechanism -- the transport of background density by the convective cells -- is a linear one. This process occurs at $k_z=0$
+for which the adiabaticity parameter iz zero ($\alpha = k_z^2 \xi = 0$) meaning the linear matrix operator is highly non-normal and subject to transient growth.
 So non-normal effects play a role in the 3D HW system for all $\xi$ due to the presence of $k_z=0$ structures.
@@
 In this section, we outline a procedure that uses quick and simple methods to approximate the turbulent growth rates $\gamma_{t,k}$.
-We contend that the key to doing this is to successfully model the effect that the nonlinearities have on the linear (modal and transient) processes.
+We contend that the key to doing this is to successfully model the effect that the nonlinearities have on the linear (transient and possibly modal) processes.
 First, we note that the advective nonlinearity terms in Eqs.~\ref{n_eq} and~\ref{phi_eq} have a common form -- a state variable multiplied by an inverse time scale:
@@
 \eeq
-Taking $E_k(t)$ as the ensemble averaged energy calculated from Eq.~\ref{E_t_ensemble_avg} and set $E_k(0) = 1$ for normalization, we find,
+Taking $E_k(t)$ as the ensemble averaged energy calculated from Eq.~\ref{E_t_ensemble_avg} and setting $E_k(0) = 1$ for normalization, we find,
 
@@
 \label{gamma_nm}
-\gamma_{{\rm{nm}},k} = \frac{1}{2 \tau_{nl,k}} \ {\rm{Log}} \left<  E_k(\tau_{nl,k}) \right>_{\rm{ens}}
+\gamma_{{\rm{nm}},k} = \frac{1}{2 \tau_{nl,k}} \ {\rm{Log}} \left<  E_k(\tau_{nl,k}) \right>_{\rm{ens}}.
 \eeq
+
 Now, in order to obtain a simple and quick prediction of $\gamma_{{\rm{nm}},k}$, we must estimate $\tau_{nl,k}$ without having to calculate it directly from a turbulent simulation. 
@@
 There are a several possible linear time scales in the HW problem to use for $\tau_{nl,k}$ 
-including inverse linear eigenmode frequency and growth rates as well as certain times associated with transient growth curves. 
+including the inverse linear eigenmode frequency and growth rates as well as certain times associated with transient growth curves. 
 After trying a few of these different time scales, we have concluded that the one that works best is the inverse of the linear eigenmode frequency $\omega_{s,k}$
@@
 \label{gamma_omega_def}
-\gamma_{\rm{nm},k} = \frac{\omega_{s,k}}{4} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k t} e^{\mathbf{B}^{\dagger}t} \} \right].
+\gamma_{\rm{nm},k} = \frac{\omega_{s,k}}{4} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \omega_{s,k}^{-1}} e^{\mathbf{B}^{\dagger} \omega_{s,k}^{-1}} \} \right].
 \eeq
-Note that the factor of $\frac{1}{4}$ is particular to two-field systems. Namely, the factor of $\frac{1}{2}$ coming from Eq.~\ref{E_t_ensemble_avg} is a consequence of the $2 \times 2$ size
+Note that the factor of $\frac{1}{4}$ is particular to two-field systems. Namely, the factor of $\frac{1}{2}$ coming from Eq.~\ref{E_t_ensemble_avg} is a consequence of the $2 \times 2$ dimensionality
 of $\mathbf{B}_k$.
@@
 
-The non-modal growth rate prediction agrees fairly well with the turbulent growth rate for highly non-normal 2D HW turbulence as seen in Fig.~\ref{gamma_max_with_nm}. For $\alpha \le 10^{-4}$,
+The peak non-modal growth rate prediction agrees fairly well with the turbulent growth rate for highly non-normal 2D HW turbulence as seen in Fig.~\ref{gamma_max_with_nm}. For $\alpha \le 10^{-4}$,
 in which the system is fairly non-normal, $\gamma_{\rm{nm},k}$ predicts $\gamma_{t,k}$ quite accurately, especially compared to the eigenmode prediction. 
@@
 but generally closer to $\gamma_{s,k}$ for $\alpha \gg 10^{-3}$.
+
 This indicates that some assumption in our procedure doesn't work well for systems that are normal or nearly normal. Namely, our assumption
-of complete randomization by the nonlinearities breaks down. Apparently, the nonlinearities only partially randomize the turbulence in the weakly non-normal case. 
-For the most part, the turbulence at a given $k$ consists primarily of the unstable eigenmode, while the stable eigenmode accounts for only a small amount of energy in the turbulence.
+of complete randomization by the nonlinearities breaks down. The nonlinearities only partially randomize the turbulence in the weakly non-normal case such that
+the turbulence at a given $k$ always consists primarily of the unstable eigenmode. The stable eigenmode never accounts for a significant portion of the energy in the turbulence.
+Makwana et al.~\cite{makwana2011} developed a condition based on the ratio of the stable to unstable eigenmode growth rates that can be used to determine the significance of the stable eigenmode. 
+They determined that for $\alpha \ge 10^{-1}$, the stable eigenmode receives very little energy, explaining why our procedure does not work well in the normal limit for the 2D HW model.
 
@@
 \caption{Comparison of the non-modal growth rate $\gamma_{\rm{nm},k}$ to the growth rates of the fastest growing eigenmode $\gamma_{s,k}$ 
-and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model with $\xi = 10^3$. The non-modal growth rate correctly predict the growth rate dominance of the $k_z=0$ mode
+and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model with $\xi = 10^3$. The non-modal growth rate correctly predicts the growth rate dominance of the $k_z=0$ mode
 despite the $k_z=0$ eigenmode stability.}
@@
 eigenmode frequency is zero ($\omega_{s,k_z=0} = 0$), so $\gamma_{\rm{nm},k_z=0}$ is not well defined (see Eq.~\ref{gamma_omega_def}). 
-We could, in fact, make the argument that since $\omega_{s,k_z=0} = 0$, or by critical balance $\tau_{nl,k_z=0} \to \infty$, it would be proper to equate $\gamma_{\rm{nm},k_z=0}$ with $\gamma_{s,k_z=0}$
-since all transients die away in the long time limit. Rather, we use $\omega_{s,k_z=0.02}$ -- where $k_z=0.02$ is the lowest finite value of $k_z$ available to our system -- in place of
+%We could, in fact, make the argument that since $\omega_{s,k_z=0} = 0$, or by critical balance $\tau_{nl,k_z=0} \to \infty$, it would be proper to equate $\gamma_{\rm{nm},k_z=0}$ with $\gamma_{s,k_z=0}$
+%since all transients die away in the long time limit. 
+Thus, we use $\omega_{s,k_z=0.02}$ -- where $k_z=0.02$ is the lowest finite value of $k_z$ available to our system -- in place of
 $\omega_{s,k_z=0}$ in the calculation of $\gamma_{\rm{nm},k_z=0}$. We contend that the nonlinear instability, which most strongly couples $k_z=0$ and $k_z=0.02$ modes, has its time scale set
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.55\textwidth]{gamma_max_vs_xi}}
-\caption{Comparison of the maximum non-modal growth rate $\gamma_{\rm{nm},k}$ to the growth rates of the fastest growing eigenmode $\gamma_{s,k}$ 
-and the of the turbulence $\gamma_{t,k}$ as a function of $\xi$ for the 3D HW model at {\bf a)} $k_z \rho_s = 0$ and {\bf b)} $k_z \rho_s \approx 0.02$. }
+\centerline{\includegraphics[width=0.52\textwidth]{gamma_max_vs_xi}}
+\caption{Comparison of the peak non-modal growth rate $\gamma_{\rm{nm},k}$ to that of the fastest growing eigenmode $\gamma_{s,k}$ 
+and the turbulence $\gamma_{t,k}$ as a function of $\xi$ for the 3D HW model at {\bf a)} $k_z \rho_s = 0$ and {\bf b)} $k_z \rho_s \approx 0.02$. }
 \label{gamma_max_vs_xi}
@@
 Despite these complications at $k_z=0$, our non-modal method of calculating $\gamma_{\rm{nm},k}$ approximates $\gamma_{t,k}$ remarkably well as seen in Figs.~\ref{gamma_max_vs_kz_with_nm}
-and~\ref{gamma_max_vs_xi}. Most importantly, it correctly predicts that $\gamma_{\rm{nm},k_z=0} > 0$. 
+and~\ref{gamma_max_vs_xi}. Most importantly, it correctly predicts that $\gamma_{t,k_z=0} > 0$. 
 Therefore, non-modal calculations explain why the 3D HW model can be dominated by the nonlinear instability rather than a linear drift wave instability. 
@@
 
-In both hydrodynamic and plasma systems, it is important to be able to predict the control parameter space that divides the turbulent and laminar states.
+In both hydrodynamic and plasma systems, it is important to be able to predict for a given set of parameters whether the system will be turbulent or not.
+Put another way, it is useful to map out the point, line, or manifold in parameter space that divides turbulent and laminar solutions.
 In many hydrodynamic systems with sheared flow, there is generally a single control parameter -- the Reynold's number $Re$~\cite{drazin1981}. For high enough $Re$ such systems
@@
 \centerline{\includegraphics[width=0.4\textwidth]{subcritical_diagram}}
-\caption{Diagram of a control parameter $R$ versus perturbation amplitude $A$ for two subcritical systems. The first curve (top blue one with $R^{-\Lambda}$ label) represents a system with no
+\caption{Diagram of a control parameter $R$ versus perturbation amplitude $A$ for two subcritical systems. The first curve (top blue one with $(R-R_g)^{-\lambda}$ label) represents a system with no
 eigenmode instability for any $R$, but which is stable to finite amplitude perturbations for $R > R_g$. 
@@
 when the control parameter becomes large (or small) enough. We illustrate these two categories with a diagram in Fig.~\ref{subcritical_diagram}. 
-The upper curve represents the first kind of subcritical system in which no unstable eigenmodes exist for any value of $R$. In this
+The upper curve represents the first kind of subcritical system in which no unstable eigenmodes exist for any value of the control parameter $R$. In this
 case, there is some transitional value of $R$, labeled $R_g$, below which turbulence cannot occur.
@@
 turbulence will be excited. Perturbations with amplitude below the line will decay away. Clearly, the higher the value of $R$, the smaller the perturbation needs to be to excite turbulence.
-Experimental evidence and theoretical arguments show this curve should have a $R^{-\Lambda}$ dependence, with $\Lambda \sim 1$~\cite{grossmann2000}.
+Experimental evidence and theoretical arguments show this curve should have a $(R-R_g)^{-\lambda}$ dependence, with $\lambda \sim 1$~\cite{grossmann2000}.
 A paradigmatic example of this kind of system is Hagen-Poisuille pipe flow -- flow in a cylindrical pipe with a parabolic flow profile -- for which $R_g \approx 2000$.
@@
 It is perhaps easy to be fooled into believing that the turbulence
-in such systems is due to eigenmode instability. In either case, one of the most important predictions is the minimum turbulent control parameter $R_g$, which is sometimes called the ``transitional''
-vlue of the control parameter. Given that this problem is many decades old
+in such systems is due to eigenmode instability even though it may not be. 
+In either case, one of the most important predictions is the minimum turbulent control parameter $R_g$, which is sometimes called the ``transitional'' value of the control parameter.
+Given that this problem is many decades old
 and still unsolved, it is a tall order to predict $R_g$ without nonlinear simulations. We nevertheless make an attempt.
@@
 magnetic field~\cite{drake1995}, but the eigenmodes of the linear operator in that case are not sinusoidal, making our non-modal analysis more difficult. 
-Biskamp et al., however, artificially modified the
+Biskamp et al.~\cite{biskamp1995}, however, artificially modified the
 3D HW model to remove the linear drive by taking the $z$ average of the $\kappa \pdiff{\phi}{y}$ term in Eq.~\ref{n_eq}, effectively eliminating the $k_z \ne 0$ drive. This leaves only the $k_z = 0$
@@
 
-We show that the system is subcritical in Fig.~\ref{init_en_ev} by plotting the energy versus time for fluctuations that are initialized with different amplitudes.
-When the fluctuations are initialized
-with high enough amplitude, the system eventually becomes turbulent.
+We show that the system is subcritical in Fig.~\ref{crit_amp_dependence}.
+In Fig.~\ref{crit_amp_dependence} a), we plot the energy versus time for fluctuations that are initialized with different amplitudes.
+The fluctuations initialized with high enough amplitude lead to turbulence. 
 Otherwise, the fluctuations grow transiently, but do not reach high enough amplitude for the nonlinearities to bootstrap the process,
 so they decay away (we set the diffusion and viscosity coefficient $D$ to $0.5$ in this section so that the decay occurs on a short enough time scale to keep the simulations short). 
-The finite amplitude turbulence threashold, as described above, is a key indicator of subcritical turbulence.
+This finite amplitude turbulence threashold, as described above, is a key indicator of subcritical turbulence.
+
+In Fig.~\ref{crit_amp_dependence} b), we show the amplitude threashold curve -- like that of Fig.~\ref{subcritical_diagram} -- for the subcritical 3D HW model.
+We obtain the points from nonlinear simulations
+and fit them to a function of the form $A = a (\kappa - \kappa_g)^{- \lambda}$, where $A$ is the lowest initial energy amplitude that leads to turbulence for each value of $\kappa$.
+The least-squares fitting provides optimal values
+for the parameters $a, \kappa_g, \rm{and} \lambda$. As expected for shear flows, $\lambda \sim 1$.
 
 \begin{figure}
-\centerline{\includegraphics[width=0.4\textwidth]{init_en_ev}}
-\caption{Total energy evolution of fluctuations of the modified 3D HW model where the initial perturbations are set to different amplitudes. All initial perturbations are initialized randomly in
-the $y$ and $z$ directions and with a Guassian in the $x$ direction. There is a critical initial amplitude threashold above which turbulence develops and maintains itself
-and below which turbuelence never develops and the fluctuations decay away. This indicates that the modified model is subject to subcritical turbulence.}
-\label{init_en_ev}
+\centerline{\includegraphics[width=0.52\textwidth]{crit_amp_dependence}}
+\caption{{\bf a)} Total energy evolution of fluctuations of the modified 3D HW model with $\xi=10$ and $\kappa=1$ where the initial perturbations are set to different amplitudes. 
+All initial perturbations are initialized randomly in the $y$ and $z$ directions and with a Guassian in the $x$ direction. 
+There is a critical initial amplitude threashold above which turbulence develops and maintains itself and below which turbuelence never develops and the fluctuations decay away. 
+{\bf b)} The amplitude threashold as a function of $\kappa$ for $\xi=10$.
+The points are obtained from nonlinear simulations like those in part a) with each point indicating the minimum amplitude needed for the simulation
+to become turbulent. The line is a fit to the points with the functional form $a (\kappa-\kappa_g)^{- \lambda}$.}
+\label{crit_amp_dependence}
 \end{figure}
 
-In this subcritical system, we predict the line in $\xi-\kappa$ parameter space that divides the space into an area that permits turbulent solutions and an area that does not. Note that we are
-not concerend here with amplitude threasholds. We simply want to know for given values of $\xi$ and $\kappa$ whether turbuelence can exist given a sufficiently large perturbation. Our method
+In this subcritical system, we predict the line in $\xi-\kappa$ parameter space that divides the space into an area that permits turbulent solutions and an area that does not. That is, we predict
+$\kappa_g(\xi)$.
+We show the answer that we obtain from nonlinear simulations along with our non-modal prediction in Fig.~\ref{subcrit_param_space}.
+To obtain $\kappa_g(\xi)$ from the nonlinear simulations,
+rather than mapping out curves like those in Fig.~\ref{crit_amp_dependence} b), we simply start all simulations with a large initial amplitude ($A = 10^4$) and vary $\kappa$ until we find the minimum $\kappa$ that
+sustains turbulence (see Fig.~\ref{kappa_en_ev} for an example).
+
+Our method
 of prediction is to find the line in $\xi-\kappa$ space at which $\gamma_{\rm{nm}} = 0$, where $\gamma_{\rm{nm}}$ is maximized over $k_x$ and $k_y$. We plot this line in Fig.~\ref{subcrit_param_space}
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.4\textwidth]{kappa_en_ev}}
-\caption{$E_{\rm{tot}}(t)$ from full nonlinear simulations of the subcritical 3D HW model. The curves that reach a steady-state value
-reach sustained subcritically turbulent states. Those that decay for large time have values of $\kappa$ too small to sustain turbulence. The transitional value of $\kappa$ is the smallest value
-for which turbulence is sustained.}
+\centerline{\includegraphics[width=0.52\textwidth]{kappa_en_ev}}
+\caption{$E_{\rm{tot}}(t)$ from full nonlinear simulations of the subcritical 3D HW model. The curves that reach a quasi-steady-state value represent sustained subcritically turbulent states. 
+Those that decay for large time have values of $\kappa$ too small to sustain turbulence. The transitional value $\kappa_g$ is the smallest value of $\kappa$ for which turbulence is sustained.}
 \label{kappa_en_ev}

### 2014-08-19 HW_non-modal.tex
Aug 18

@@
 \begin{abstract}
-
+Linear eigenmode analysis fails to account for turbulence in subcritical systems in many cases because such systems have non-normal linear operators, and thus
+nonorthogonal eigenmodes, which can cause fluctuations to transiently grow faster than expected from eigenmode analysis. When combined with energetically conservative nonlinear mode mixing, 
+transient growth can lead to sustained turbulence even in the absence of eigenmode instability. 
+Since linear operators ultimately provide the turbulent fluctuations with energy, it is useful to define a growth rate that takes into account non-modal effects, allowing for
+prediction of energy injection and transport levels, and even turbulent onset in the subcritical regime. 
+We define such a non-modal growth rate by modelling the statistical effect that the nonlinearities have on cross-phases and amplitude ratios of the system state variables. 
+In particular, we model the nonlinearities as delta-function-like, periodic forces that randomize the state variables once every eddy turnover time. Furthermore, we estimate the eddy turnover
+time to be the inverse of the least stable eigenmode frequency or growth rate, which allows for prediction without nonlinear numerical simulation.
+We test this procedure on the 2D and 3D Hasegawa-Wakatani model
+[A. Hasegawa and M. Wakatani, Phys. Rev. Lett. 50, 682 (1983)] as well as on a modified subcritical version of the 3D Hasegawa-Wakatani model and find that
+the non-modal growth rate is a good predictor of energy injection rates in some regimes, though not yet in the subcritical regime.
 \end{abstract}
@@
 
-Turbulence in magnetically confined plasmas is almost always attributed to linear instabilities. The calculation of linear instabilities through normal mode analysis
--- the calculation of eigenvalues and eigenvectors of a linearized dynamical system -- has successfully predicted the onset criterion for many MHD and micro-instabilities.
-Nevertheless, it is well-known in the hydrodynamics community that normal mode analysis fails at predicting turbulent onset for a large class of experiments, in which the
+The calculation of linear instabilities through normal mode analysis, which solves for the eigenvalues and eigenvectors of a linearized dynamical system,
+has successfully predicted the turbulent onset criterion for many plasma and fluid systems.
+Nevertheless, it is well-known in the hydrodynamics community that normal mode analysis has failed to predict turbulent onset in many experiments -- those in which the
 turbulence is called subcritical~\cite{drazin1981}. 
-In these experiments, turbulence appears even though all linear eigenvectors are stable, meaning infinitesimal perturbations on the laminar state cannot grow exponentially. 
-Nevertheless, finite amplitude perturbations can still excite turbulence.
-Subcritical turbulence is certainly possible in plasmas as well, and several instances have been studied in the 
-past~\cite{waltz1985,scott1990,nordman1993,biskamp1995,drake1995,itoh1996,camargo1998,krommes1999,camporeale2009,schekochihin2012,highcock2012}. 
-For the most part, however, normal mode analysis dominates plasma turbulence prediction despite the fact that it is not useful in all instances.
+In subcritical systems, turbulence develops even though all linear eigenvectors are stable such that infinitesimal perturbations on the laminar state cannot grow exponentially. 
+Nevertheless, finite amplitude perturbations still excite turbulence through nonlinear instabilities.
+Subcritical turbulence is common in plasmas as well, and several instances have been studied in the literature~\cite{waltz1985,scott1990,nordman1993,biskamp1995,drake1995,itoh1996,camargo1998,krommes1999,camporeale2009,schekochihin2012,highcock2012}.
 
@@
 Such behavior is obscured by traditional normal mode analysis, which only effectively describes the long time asymptotic behavior of fluctuations under  
-action of the linear operator. Transient growth, which can dominate turbulent evolution, can be discovered only through non-modal calculations.
+action of the linear operator. Transient growth, which can dominate turbulent evolution, must be found through non-modal calculations.
 
@@
 However, they have not been used to predict onset criteria in subcritical systems (e.g. the transition Reynold's number in pipe flows).
-This paper outlines and illustrates a non-modal linear method to make quantitative predictions of turbulent properties including turbulent onset criteria. 
-Our approach is to define a non-modal growth rate that is analagous to an eigenmode growth rate, but which takes into account transient growth effects. To accomplish this, we make
-several simplifying assumptions regarding the nonlinearities; namely, the nonlinearities randomize the spatial structures of the turbulence on a characteristic nonlinear time scale, which can
-be estimated by a characteristic linear time scale.
+This paper outlines and illustrates a non-modal method to make quantitative predictions of turbulent properties including turbulent onset criteria. 
+Our approach is to define a non-modal growth rate that is analagous to an eigenmode growth rate, but which takes into account transient growth effects.
 
-More specifically, we model the turbulent steady state as a series of processes:  (1) the turbulence starts as a spatially random state, 
-(2) linear transient growth deterministically amplifies the turbulent energy (or depresses it, depending on the wavenumber), 
-(3) nonlinear transfer sets in at a specified timescale, terminating the transient growth process and re-randomizing the turbulent state (at which point the cycle repeats). 
-Optimally, the timescale for the final step would be the nonlinear decorrelation time of the turbulent system, but in order to enable predictive capability, we employ critical balance
-arguments to use a characteristic linear time. The procedure ultimately produces a growth rate that can be used to predict turbulent properties such as saturation levels and transport rates 
-through mixing length arguments. Furthermore, a finite positive growth rate (of a certain magnitude) indicates that turbulence can exist given a sufficient amount of noise in the system.
-Thus, the growth rate provides a turbulent onset criterion, even for subcritical systems.
+To accomplish this, we model the effect that the nonlinearities have on the complex ratio of the state variables. Our model is one in which the 
+nonlinearities randomize the complex ratios on a characteristic nonlinear time scale.
+We estimate the nonlinear time scale by a characteristic linear time scale, taking either the inverse linear eigenmode frequency or growth rate as the linear time scale.
+Between the nonlinear randomization, the system evolves linearly, and we use this evolution to calculate the non-modal growth rate.
+This non-modal growth rate takes the place of the eigenmode growth, which can be used
+to give energy injection rates, turbulent saturation levels, and the turbulent onset criteria.
+That is, we argue that a finite positive non-modal growth rate indicates that turbulence can exist, given, of course a sufficient perturbative kick.
 
-The paper is organized as follows: in Section~\ref{sec_hw_model}, we specify our illustrative model -- the 2D and 3D Hasegawa-Wakatani model -- 
+The paper is organized as follows: in Section~\ref{sec_hw_model}, we specify our illustrative model, the 2D and 3D Hasegawa-Wakatani model,
 that we use throughout the paper and describe its linear non-normal properties,  while in Section~\ref{sec_non_norm_turb}, we explore the effect of linear non-normality in the model on turbulence. 
-In Section~\ref{sec_nm_procedure}, we cover the details of our non-modal predictive procedure and show its success in predicting properties of highly non-normal turbulence. 
-Finally, in Section~\ref{sec_subcrit_prediction}, we modify the 3D Hasegawa-Wakatani model to make it subcritical and show how our non-modal technique predicts the critical gradient for the onset of turbulence.
+In Section~\ref{sec_nm_procedure}, we cover the details of our procedure to calculate the non-modal growth rate and compare show it to growth rates derived directly from nonlinear numerical simulations.
+Finally, in Section~\ref{sec_subcrit_prediction}, we modify the 3D Hasegawa-Wakatani model to make it subcritical and show to what degree our non-modal technique predicts the critical gradient for the onset of turbulence.
 
@@
 In the analysis and numerical simulations, which we perform with the BOUT++ code~\cite{dudson2009}, we use periodic boundary conditions in
-the $y$ and $z$ directions with zero-value boundaries in the $x$ direction. We use a domain size of $64 \rho_s \times 64 \rho_s \times 300 \rho_s$.
+the $y$ and $z$ directions with zero-value boundaries in the $x$ direction. We use a domain size of $64 \rho_s \times 64 \rho_s \times 320 \rho_s$.
 For the 2D version of the model, we replace $-\xi \gradpar^2$ with a single parameter $\alpha$ (definied as $\alpha = 4 \pi k_z^2 \rho_s^2 \omega_{ce} /\nu_{ei})$, 
@@
 In numerical simulations, the turbulence can laminarize due to the build-up of $k_y=0$ density and potential structures~\cite{biskamp1995}. To prevent this, we subtract out the $k_y=0$
-$n$ and $\phi$ components, which is the equivalent of adding source terms to Eqs.~\ref{n_eq} and~\ref{phi_eq}~\cite{friedman2012b}.
+density and potential components, which is the equivalent of adding source terms to Eqs.~\ref{n_eq} and~\ref{phi_eq}~\cite{friedman2012b}.
 
 The degree of non-normality of the linear operator in the HW model is largely a function of the adiabaticity parameter~\cite{camargo1998}, or in the 3D case, of $\xi \gradpar^2$. 
-Note that a matrix is normal if it commutes with its adjoint: $\mathbf{D} \mathbf{D}^\dagger = \mathbf{D}^\dagger \mathbf{D}$ for $\mathbf{D}$ normal. 
-The HW model has a normal linear operator in the limit of $\alpha, \xi \gradpar^2 \to \infty$. 
-To see this, we first transform the linear operator into a linear matrix by using a Fourier decomposition. Because we use a zero-value boundary condition in the $x$ direction, the
+A normal matrix is on that commutes with its adjoint: $\mathbf{D} \mathbf{D}^\dagger = \mathbf{D}^\dagger \mathbf{D}$.
+The HW model has a normal linear operator in the adiabatic limit: $\alpha, \xi \to \infty$. 
+This is evident when we transform the linear operator into a linear matrix by using a Fourier decomposition. Because we use a zero-value boundary condition in the $x$ direction, the
 Fourier decomposition takes the form:
@@
 Since the square $L_2$-norm of the new state vector $\frac{1}{\sqrt{2}} \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right)$ gives the energy contained in the mode $k$,
-$\mathbf{B}_k$ is the energetically-weighted matrix of interest. In the limit of large $\xi k_z^2$ (as long as $k_\perp$ is also not too large):
+$\mathbf{B}_k$ is the energetically-weighted matrix of interest. In the adiabatic limit (as long as $k_\perp$ remains finite):
 
@@
 \eeq
-where it is easy to see that $\mathbf{B}_k$ is self-adjoint and therefore normal. In the opposite limit:
+where it is easy to see that $\mathbf{B}_k$ is self-adjoint and therefore normal. In the opposite (hydrodynamic) limit:
 
@@
 \eeq
-Thus, $\mathbf{B}_k$ is non-normal. In general, as the adiabaticity parameter is varied, the normality of $\mathbf{B}_k$ changes, but unfortunatley there is no absolute metric to quantify
-the normality of a matrix or operator~\cite{trefethen2005}. In practice, however, what matters is not how non-normal $\mathbf{B}_k$ is, but how much transient energy growth is possible in the linear
+Thus, $\mathbf{B}_k$ is non-normal.
+In general, as the adiabaticity parameter is varied, the normality of $\mathbf{B}_k$ changes.
+There are several ways to assess the non-normality of a matrix~\cite{trefethen2005}.
+Perhaps the most intuitive is to use the scalar measure proposed by Henrici~\cite{henrici1962}:
+
+\beq
+\label{henrici_num}
+H = \frac{||\mathbf{B} \mathbf{B}^\dagger - \mathbf{B}^\dagger \mathbf{B}||}{|| \mathbf{B} ||^2}
+\eeq
+where the operator $|| \cdot ||$ is the $L_2$-norm.
+This scalar normality measure can take on values: $0 \le H < 1$, where $H=0$ for a normal matrix.
+Since we are generally interested in highly non-normal
+matrices where $H$ can be very close to $1$, we will look at the quantity $1-H$ rather than $H$ itself.
+Furthermore, in practice, what matters is how much transient energy growth is possible in the linear
 system. The linear HW system is
@@
 \eeq
-where $u_k = \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right)$. The solution to this is $u_k(t) = e^{\mathbf{B}_k t} u(0)$ where the exponential of the matrix is defined in terms of
-its Taylor expansion. The solution is dependent on the initial condition $u(0)$. The energy growth as a function of time is
+where $u_k = \left( n_k, k_\perp \phi_k \right)^T$.
+The solution to this is $u_k(t) = e^{\mathbf{B}_k t} u_k(0)$ where the exponential of the matrix is defined in terms of
+its Taylor expansion. The solution is dependent on the initial condition $u_k(0)$. The energy as a function of time is
 
 \beq
-\label{Gt}
-G_k(t) = \frac{||u(t)||^2}{||u(0)||^2} = \frac{||e^{\mathbf{B}_k t} u(0)||^2}{||u(0)||^2}
+\label{Et}
+E_k(t) = \frac{||u_k(t)||^2}{||u_k(0)||^2} = \frac{||e^{\mathbf{B}_k t} u_k(0)||^2}{||u_k(0)||^2}
 \eeq
-where the operator $|| \cdot ||$ is the $L_2$-norm. Different initial conditions provide optimal energy growth at different times, but the upper envelope of all such optimal growth curves is given by
-$G_{k,\rm{max}}(t) = ||e^{\mathbf{B}_k t}||^2$, which does not depend on an initial condition.
+Different initial conditions provide optimal energy growth at different times, but the upper envelope of all such energy evolution curves is given by
+$E_{k,\rm{max}}(t) = ||e^{\mathbf{B}_k t}||^2$, which does not depend on an initial condition.
 Moreover, $||e^{\mathbf{B}_k t}||^2 \ge e^{2 \gamma_{s,k} t}$, where $\gamma_{s,k}$ is the real part of
 the fastest growing (or least damped) eigenmode of $\mathbf{B}_k$ (and of $\mathbf{A}_k$, since $\mathbf{B}_k$ is just a similarity transform of $\mathbf{A}_k$~\cite{camargo1998}). 
-The equality holds when $\mathbf{B}_k$ is normal, meaning the normal case provides a lower bound to the maximal energy growth. 
+The equality holds when $\mathbf{B}_k$ is normal. 
 Thus, non-normality in practice allows a system to undergo more energy growth than is dictated by the most unstable linear eigenmode. We show this in Fig.~\ref{max_en_growth} where we plot
-$||e^{\mathbf{B}_k t}||^2$ for different values of $\alpha$ along with $e^{2 \gamma_{s,k} t}$, showing that non-modal growth is possible for small adiabaticity parameter $\alpha$. Furthermore,
-the short time transient growth becomes larger as $\alpha$ becomes smaller.
+$||e^{\mathbf{B}_k t}||^2$ for different values of $\alpha$ along with $e^{2 \gamma_{s,k} t}$, showing that transient non-modal growth is strong for small values of the scalar normality measure $1-H$, which scales monotonically
+with the adiabaticity parameter $\alpha$.
 
@@
 \caption{$G_{k,\alpha,\rm{max}}(t) = ||e^{\mathbf{B}_{k,\alpha} t}||^2$ for $k_x=0, k_y=1, \kappa=1$ for different values of the 2D adiabaticity parameter 
-$\alpha$ compared to the energy growth $e^{2 \gamma_{s,\alpha,k} t}$ of the least stable eigenmode of  $\mathbf{B}_{k,\alpha}$. 
+$\alpha$ compared to the energy growth $e^{2 \gamma_{s,\alpha,k} t}$ of the least stable eigenmode of  $\mathbf{B}_{k,\alpha}$. Also indicated for each matrix is its scalar normality measure $1-H$.
 The time axis is normalized to $\gamma_{s,\alpha}$, which is different for each of the curves. This normalization makes the $e^{2 \gamma_{s,\alpha,k} t}$ line the same for all $\alpha$. 
@@
 
-Non-normality manifests itself in its eigenvectors -- non-normal matrices or operators have non-orthogonal linear eigenvectors.
+Non-normality manifests itself in its eigenvectors -- non-normal matrices or operators have nonorthogonal linear eigenvectors.
 For the HW model, each Fourier matrix $\mathbf{B}_k$ has two eigenvectors associated
@@
 The two eigenvalues have different growth rates (the real part), but the same frequency, though with opposite sign.
-As $\alpha \to 0$, the two eigenvectors become less orthogonal -- more parallel or anti-parallel to one another. When the eigenvectors are nearly anti-parallel and initialized with finite amplitude,
+As $\alpha, 1-H \to 0$, the two eigenvectors become less orthogonal -- more parallel or anti-parallel to one another.
+When the eigenvectors are nearly anti-parallel and initialized with finite amplitude,
 their superposition, which constitutes the total system at $k$, may linearly evolve so that the norm of the superposition grows faster than either individual eigenvector. A paradigmatic illustration
@@
 
-Linear non-normality is important for the effect it has on turbulence. It is most significantly a necessary condition for turbulence in subcritical systems that have only conservative nonlinearities 
-(e.g. advective nonlinearities). However, non-normality can also impact non-subcritical systems, which have unstable linear eigenmodes like the HW system. 
-Since certain superpositions of the linear eigenmodes can inject energy faster than the most unstable linear
+Linear non-normality is important for the effect it has on turbulence. For one, non-normality is a necessary condition for turbulence in subcritical systems that have only conservative nonlinearities 
+(e.g. advective nonlinearities). But it can also impact non-subcritical systems -- those which do have unstable linear eigenmodes (e.g. the 2D and 3D HW model). 
+Since certain superpositions of the nonorthogonal linear eigenmodes can inject energy faster than the most unstable linear
 eigenmode, there is the possibility that the energy injection into the turbulence is faster and at different wavenumbers than is suggested by linear eigenmode analysis.
@@
 In order to define the turbulent growth rate in general and specifically for the HW model, we employ a Fourier-decomposed energy dynamics analysis~\cite{camargo1995,friedman2012b,friedman2013}. 
-Specifically, we take Eqs.~\ref{n_eq} and~\ref{phi_eq} and substitute the following:
+First, we take Eqs.~\ref{n_eq} and~\ref{phi_eq} and substitute the following:
 
 \beqar
-\label{subs_fourier}
+\label{subs_fourier_n}
 n = \sum_{k} n_k \ {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z} \\
+\label{subs_fourier_phi}
 \phi = \sum_{k} \phi_k \ {\rm sin} \left( k_x x \right) e^{i k_y y + i k_z z},
@@
 where, again, $k$ stands for $(k_x,k_y,k_z)$.
-Then, we multiply the resulting first equation by $n^*_{k'} \ {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$ and the second equation by 
+Then, we multiply Eq.~\ref{subs_fourier_n} by $n^*_{k'} \ {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$ and Eq.~\ref{subs_fourier_phi} by 
 $-\phi^*_{k'} \ {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$, take the volume integral and add the two equations together. The result is
@@
 \frac{1}{2} \diff{}{t} \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right) = -i k_y \kappa \phi_k n^*_k - \xi k_z^2 |n_k - \phi_k|^2 \nonumber \\
-- D k^4_\perp \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right) + \sum_{k'} T(k,k'), \quad \quad
+- D k^4_\perp \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right) + \sum_{k'} T(k,k'). \quad \quad
 \eeqar
-where, again, the energy $E_k$ is $\frac{1}{2} \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right)$, 
-and the term $\sum_{k'} T(k,k')$ represents the nonlinear contribution (three-wave transfers between different $k$'s), 
+Recall that the energy $E_k$ is $\frac{1}{2} \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right)$, so the left hand side is the time derivative of the energy.
+The term $\sum_{k'} T(k,k')$ represents the nonlinear contribution (three-wave transfers between different $k$'s), 
 which we do not write explicitly. This energy evolution equation may be written symbolically as
@@
 \eeq
-where $\diff{E_{l,k}}{t}$ represents the terms that come from the linear terms in Eqs.~\ref{n_eq} and~\ref{phi_eq}, which are quadratic in the fluctuating
-quantities $n_k$ and $\phi_k$ in Eq.~\ref{Explicit_En_eqn}. 
-$\diff{E_{l,k}}{t}$ represents the injection of energy into the fluctuations from the free energy in the equilibrium gradients plus the dissipation from collisions and hyperdiffusion.
+where $\diff{E_{l,k}}{t}$ represents the the linear terms on the RHS of Eq.~\ref{Explicit_En_eqn}, which are quadratic in the fluctuating quantities $n_k$ and $\phi_k$.
+$\diff{E_{l,k}}{t}$ describes the injection of energy into the fluctuations from the free energy in the equilibrium gradients plus the dissipation from collisions and hyperdiffusion/viscosity.
 $\diff{E_{nl,k}}{t}$ represents $\sum_{k'} T(k,k')$ in  Eq.~\ref{Explicit_En_eqn} and comes from the nonlinear terms in Eqs.~\ref{n_eq} and~\ref{phi_eq}. 
@@
 Now, in quasi-steady state turbulence, the rate of net energy injection (or dissipation) into the fluctuations at each $k$ by the linear terms must be balanced by
-the rate of energy removal (or deposition) from the nonlinear terms. This may be represented formally using growth rates:
+the rate of energy removal (or deposition) from the nonlinear terms. Formally,
 
@@
 \label{steady_state}
+\gamma_k = \gamma_{l,k} + \gamma_{nl,k} = 0, \\
 \gamma_k \equiv  \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_k/dt}{2 E_k} dt \nonumber \\
@@
 \eeqar
-The limit of the last expression vanishes -- as indicated -- for steady or quasi-steady state turbulence only.
-From, Eqs.~\ref{dEdt_def} and~\ref{steady_state}, it follows that $ \gamma_k = \gamma_{l,k} + \gamma_{nl,k} = 0$.
+$\gamma_k$ vanishes, as indicated, for steady or quasi-steady state turbulence, meaning $\gamma_{l,k} = - \gamma_{nl,k}$.
 
-We now associate the turbulent growth rate $\gamma_{t,k}$ discussed above with the linear rate of energy injection into the fluctuations $\gamma_{l,k}$ because $\gamma_{l.k}$
+It is the linear rate of energy injection into the fluctuations $\gamma_{l,k}$ that we wish to use as the turbulent growth rate $\gamma_{t,k}$ because $\gamma_{l.k}$
 determines how fast and where in wavenumber space energy is injected into the turbulence.
 $\gamma_{l,k}$, in other words, is the turbulence equivalent to the linear eigenmode growth rate $\gamma_{s,k}$~\cite{friedman2012b,terry2006b}. 
-That is, in a linear simulation run for a sufficiently long time where transients die away, the calculation of $\gamma_{l,k}$ gives exactly $\gamma_{s,k}$.
-Moreover, $\gamma_{l,k}$ is calculated from the spatial structures of the plasma state variables, so it is always well-defined, even in a turbulent plasma.
+That is, in a linear simulation run for a sufficiently long time where transients die away, the calculation of $\gamma_{l,k}$ from the simulated $n_k$ and $\phi_k$ gives exactly $\gamma_{s,k}$, which is generally
+calculated from the eigenvalue problem rather than a direct simulation.
+However, we always calculate $\gamma_{t,k}$ from steady state turbulent $n_k$ and $\phi_k$ fluctuations.
 
-A turbulent system, which can be decomposed into a linear eigenvector basis, has $\gamma_{t,k} \le \gamma_{s,k}$ if the system is normal. Recall that in normal systems, the maximal
+The turbulent growth rate $\gamma_{t,k}$ is an informative quantity. 
+Note that in Eq.~\ref{Explicit_En_eqn} all of the linear terms on the right hand side of the equation are negative, except for the first term, which can have either sign. 
+This term is proportional to the cross-field particle flux $\Gamma_x = \left< v_{E,x} n \right> \sim -i k_y \phi_k n_k$, meaning that a positive $\gamma_{t,k}$ indicates an outward particle flux
+due to $k$ wavenumber flows. Particle transport, then, requires that $\gamma_{t,k} > 0$ for at least one value of $k$. Furthermore, since the nonlinearities are energetically conservative,
+positivity of at least one $\gamma_{t,k}$ is necessary for the fluctuations to sustain themselves, meaning that it is a necessary condition for turbulence and transport. 
+Finally, through the mixing length approximation $|\phi| \sim \frac{\gamma_{t,k}}{k_\perp^2}$, $\gamma_{t,k}$ can provide an estimate of the turbulent saturation level.
+
+In turbulent systems with normal linear operators $\gamma_{t,k} \le \gamma_{s,k}$. This is evident if one recalls that in normal systems, the maximal
 energy growth is given by the most unstable linear eigenmode. In other words, maximal growth is achieved when only the most unstable eigenmode has finite amplitude.
-On the other hand, non-normal systems may grow faster when the more stable eigenmodes have finite amplitude than when they have none (see Fig.~\ref{max_en_growth}). 
+If the stable eigenmode has finite amplitude, it decreases the turbulent growth rate.
+On the other hand, non-normal systems may grow faster when the stable eigenmodes have finite amplitude than when they have none (see Fig.~\ref{max_en_growth}). 
 It is therefore possible for $\gamma_{t,k} > \gamma_{s,k}$ in non-normal systems.
 
+For the 2D HW model, we show
+$\gamma_{s,k}$ and $\gamma_{t,k}$ for two different values of $\alpha$. In Fig.~\ref{alpha1e-5_gamma_s_t}, we show these growth rate spectra for $\alpha = 10^{-5}$, for which the system is
+highly non-normal, and in Fig.~\ref{alpha1_gamma_s_t} for $\alpha = 1$, for which the system is closer to being normal (see the inset of Fig.~\ref{gamma_max_vs_alpha}).
+All calculations in this section use $\kappa=1$.
+Note the different scales used in all of the figures. In the highly
+non-normal case, the turbulent growth rate is somewhat higher than the eigenmode growth rate for most values of $k_x,k_y$, and about 6 times higher where the spectrum peaks. For the more normal
+case, in contrast, $\gamma_{t,k}$ is generally lower than $\gamma_{s,k}$, with the peak at nearly a factor of 2 less, with the peaks being at slightly different values of $k_y$.
 \begin{figure}
@@
 \label{alpha1e-5_gamma_s_t}
-\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[width=0.52\textwidth]{alpha1_gamma_s_t}}
-\caption{{\bf a)} The eigenmode growth rate spectrum $\gamma_{s,k}$ and {\bf b)} the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 1, \kappa=1$.}
-\label{alpha1_gamma_s_t}
-\end{figure}
-
-For the 2D HW model, we show
-$\gamma_{s,k}$ and $\gamma_{t,k}$ for two different values of $\alpha$. In Fig.~\ref{alpha1e-5_gamma_s_t}, we show these growth rate spectra for $\alpha = 10^{-5}$, for which the system is
-highly non-normal, and in Fig.~\ref{alpha1_gamma_s_t} for $\alpha = 1$, for which the system is very close to being normal. All calculations in this section use $\kappa=1$. 
-Note the different scales used in all of the figures. In the highly
-non-normal case, the turbulent growth rate is somewhat higher than the eigenmode growth rate for most values of $k_x,k_y$, and about 6 times higher where the spectrum peaks. For the more normal
-case, in contrast, $\gamma_{t,k}$ is generally lower than $\gamma_{s,k}$, with the peak at nearly a factor of 2 less, with the peaks being at slightly different values of $k_y$.
-
-\begin{figure}
-\centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_alpha}}
-\caption{The eigenmode $\gamma_{s}$ and turbulent $\gamma_{t}$ growth rates as a function of $\alpha$, which controls the normality of the linear operator.
-All growth rates are peaks in $k_x-k_y$ space.}
-\label{gamma_max_vs_alpha}
 \end{figure}
@@
 For both the normal and highly non-normal limits of the 2D HW model, $\gamma_{s,k}$ does a poor job of predicting $\gamma_{t,k}$ because $\gamma_{s,k}$ neglects the stable branch of the
-dispersion relation -- the damped eigenmode. The importance of the stable eigenmode branch (or branches) 
-has started to gain acceptance in the plasma community~\cite{baver2002}, although it has generally been used in normal
+dispersion relation -- the damped eigenmode.
+Additionally, in the moderate normality region where $\gamma_{s,k}$ does match $\gamma_{t,k}$ well, it is misleading to think of the system as being composed of only the
+unstable eigenmode.
+The damped mode simply has both damping and transient growth effects depending on its relative magnitude at a particular time.
+These effects nearly cancel out on average when $\gamma_{t,k} \approx \gamma_{s,k}$.
+
+\begin{figure}
+\centerline{\includegraphics[width=0.52\textwidth]{alpha1_gamma_s_t}}
+\caption{{\bf a)} The eigenmode growth rate spectrum $\gamma_{s,k}$ and {\bf b)} the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 1, \kappa=1$.}
+\label{alpha1_gamma_s_t}
+\end{figure}
+
+The importance of the stable eigenmode branch~\cite{baver2002} (or branches) has started to gain acceptance in the plasma community, although it has generally been used in normal
 or nearly normal systems to explain turbulent saturation~\cite{terry2006b,hatch2011,makwana2011} ($\gamma_{t,k} < \gamma_{s,k}$). 
@@
 In fact, simulation and analysis of the 3D HW model (and extended models) is an area in which others have seen the effects of strong non-normal transient growth
-but have not realized them as such~\cite{biskamp1995,drake1995,scott2002,scott2005,umansky2009,friedman2012b}.
-
-\begin{figure}
-\centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_kz}}
-\caption{The growth rates of the fastest growing eigenmode $\gamma_{s,k}$ and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model at $\xi = 10^3$.
-The most remarkable feature is
-that the turbulent growth rate peaks at $k_z=0$ despite the stability of all linear eigenmodes at $k_z=0$.}
-\label{gamma_max_vs_kz}
-\end{figure}
+but have not connected them to the nonorthogonal eigenmode transient growth theory~\cite{biskamp1995,drake1995,scott2002,scott2005,umansky2009,friedman2012b}.
 
@@
 
+\begin{figure}
+\centerline{\includegraphics[width=0.4\textwidth]{gamma_max_nn}}
+\caption{The eigenmode $\gamma_{s}$ and turbulent $\gamma_{t}$ growth rates as a function of $\alpha$, which controls the normality of the linear operator.
+All growth rates are peaks in $k_x-k_y$ space.
+The insert shows the scalar normality measure $1-H$ as a function of $\alpha$ for the matrices $\mathbf{B}_k$ that give the $\gamma_{s}$ curve.}
+\label{gamma_max_vs_alpha}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_kz}}
+\caption{The growth rates of the fastest growing eigenmode $\gamma_{s,k}$ and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model at $\xi = 10^3$.
+The most remarkable feature is
+that the turbulent growth rate peaks at $k_z=0$ despite the stability of all linear eigenmodes at $k_z=0$.}
+\label{gamma_max_vs_kz}
+\end{figure}
+
 \section{Non-Modal Linear Procedure To Calculate Turbulent Growth Rate}
@@
 
-In this section, we outline a procedure that uses quick and simple methods to approximate the turbulent growth rates $\gamma_{t,k}$.
-We contend that the key to doing this is to successfully model the effect that the nonlinearities have on the linear (transient and possibly modal) processes.
-First, we note that the advective nonlinearity terms in Eqs.~\ref{n_eq} and~\ref{phi_eq} have a common form -- a state variable multiplied by an inverse time scale:
-$\tau_{nl}^{-1} \sim v_E k_\perp$.
-This nonlinear time scale is generally associated with the eddy turnover or decorrelation time. We thus present a heuristic model of the nonlinearities 
-as a randomizing force that acts on this characteristic nonlinear time scale. More specifically, this model is one in which 1) the turbulence begins as a random state, 
-2) evolves linearly for a time $\tau_{nl}$, and 3) randomizes by nonlinear energy transfer, at which point the steps repeat.
+In this section, we outline a procedure that uses quick and simple methods to approximate the turbulent growth rate spectrum $\gamma_{t,k}$. To do this, first recall that $\gamma_{t,k}$ is calculated
+from only linear terms in the original model equations (Eqs.~\ref{n_eq} and~\ref{phi_eq}).
+The exact form of the nonlinearities do not enter the calculation of $\gamma_{t,k}$.
+Rather, the nonlinearities
+simply influence the quantities $n_k$ and $\phi_k$ that are used to calculate $\gamma_{t,k}$. Furthermore, $\gamma_{t,k}$ is a statistical (time-averaged) quantity, and therefore, it is not
+necessary to know its instantaneous value, or likewise, the instantaneous values of $n_k$ and $\phi_k$.
+In fact, the only quantity that affects $\gamma_{t,k}$ is the average complex ratio $n_k/\phi_k$.
+Thus, it is feasible that $\gamma_{t,k}$ may be approximated without large numerical simulations.
 
-We can calculate an effective growth rate from this model by starting with an ensemble of random initial conditions, evolving them linearly for a time $\tau_{nl}$, and
-then taking the time and ensemble averaged growth rate of these curves.
-In practice, this procedure can be greatly sped up and simplified. To see this, recall that the time evolution of the energy from an initial condition is
+Our method of approximation involves modelling the nonlinearities as instantaneous randomizing forces that act periodically.
+In this regard, note that the advective nonlinearity terms in Eqs.~\ref{n_eq} and~\ref{phi_eq} have a certain time scale associated with them: $\tau_{nl}^{-1} \sim v_E k_\perp$.
+This nonlinear time scale is generally associated with the eddy turnover or decorrelation time. Our heuristic model of the nonlinearities is then one
+in which they randomize the complex ratio $n_k/\phi_k$ once every $\tau_{nl}$.
+This model equivalently describes a system in which 1) the turbulence begins as a random state, 
+2) evolves linearly for a time $\tau_{nl}$, and 3) randomizes again -- due to nonlinear energy transfer -- at which point the step 2 is repeated and so on.
+
+What makes this model truly appealing is its simplicity. In fact, we can use it to \emph{analytically} calculate an effective growth rate. It requires no numerical simulation.
+To see this, recall that the linear time evolution of the energy from an initial condition is
 
@@
 \eeq
-Since we want the ensemble of initial conditions ($u(0)$'s) to be random normalized vectors with uncorrelated components, it follows that~\cite{camargo1998}
+where $e^{\mathbf{B}_k^{\dagger}t}$ is shorthand for $\left( e^{\mathbf{B}_k t} \right)^{\dagger}$.
+Since $\gamma_{t,k}$ is a time-averaged quantity and the nonlinear model randomizes the system periodically, we can use an ensemble of initial conditions ($u(0)$'s that are random normalized vectors with uncorrelated components),
+and take two averages (over $t=0$ to $t = \tau_{nl}$ and over the ensemble) in place of the single time average that is generally used to calculate $\gamma_{t,k}$.
+The two averages commute, but for simplicity,
+we look at each average separately and then put them together.
+
+For the ensemble average, Camargo et al.~\cite{camargo1998,friedman2014} showed that the ensemble average of the energy evolution in Eq.~\ref{E_t_from_u0} could be simplified as,
 
@@
 \eeq
-where the normalization is $||u(0)||^2 = 1$. Mathematically, we define the non-modal growth rate $\gamma_{{\rm{nm}},k}$ as the time averaged instantaneous growth rate:
+where we have initialized $||u(0)||^2 = 1$ for convenience.
+
+Second, the short-time averaged growth rate from $t=0$ to $t = \tau_{nl}$ is simplified as,
 
@@
 \label{gamma_nm_calc}
-\gamma_{{\rm{nm}},k} = \frac{1}{\tau_{nl,k}} \int_0^{\tau_{nl,k}} \frac{\pdiff{E_k(t)}{t}}{2 E_k(t)} dt = \frac{1}{2 \tau_{nl,k}} \ {\rm{Log}} \left[ \frac{E_k(\tau_{nl,k})}{E_k(0)}\right].
+\frac{1}{\tau_{nl,k}} \int_0^{\tau_{nl,k}} \frac{\pdiff{E_k(t)}{t}}{2 E_k(t)} dt = \frac{1}{2 \tau_{nl,k}} \ {\rm{Log}} \left[ \frac{E_k(\tau_{nl,k})}{E_k(0)}\right].
 \eeq
-Taking $E_k(t)$ as the ensemble averaged energy calculated from Eq.~\ref{E_t_ensemble_avg} and setting $E_k(0) = 1$ for normalization, we find,
+
+Now, putting these commuting averages together, and again setting $E_k(0) = 1$ for normalization, we arrive at the non-modal growth rate:
 
@@
 \label{gamma_nm}
-\gamma_{{\rm{nm}},k} = \frac{1}{2 \tau_{nl,k}} \ {\rm{Log}} \left<  E_k(\tau_{nl,k}) \right>_{\rm{ens}}.
+\gamma_{{\rm{nm}},k} = \frac{1}{4 \tau_{nl,k}} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \tau_{nl,k}} e^{\mathbf{B}^{\dagger} \tau_{nl,k}} \} \right].
 \eeq
-
-Now, in order to obtain a simple and quick prediction of $\gamma_{{\rm{nm}},k}$, we must estimate $\tau_{nl,k}$ without having to calculate it directly from a turbulent simulation. 
-We thus invoke the conjecture of \emph{critical balance}, which posits that the nonlinear time scale equals a characteristic linear time scale at all spatial scales~\cite{schekochihin2012}. 
-This is a fair assumption in light of our discussion above, in which we showed that $\gamma_{l,k} = - \gamma_{nl,k}$.
-
-There are a several possible linear time scales in the HW problem to use for $\tau_{nl,k}$ 
-including the inverse linear eigenmode frequency and growth rates as well as certain times associated with transient growth curves. 
-After trying a few of these different time scales, we have concluded that the one that works best is the inverse of the linear eigenmode frequency $\omega_{s,k}$
--- the imaginary part of the eigenvalues of the linear matrix operator $\mathbf{B}_k$.
-Recall that the imaginary part of the linear eigenvalues is the same for both eigenvalues of $\mathbf{B}_k$, so there is no question of which to use.
-Because of this, unlike the eigenmode growth rates,
-the frequency is robust no matter how nonorthogonal the eigenvectors. We thus arrive at our final definition for the non-modal growth rate:
-
-\beq
-\label{gamma_omega_def}
-\gamma_{\rm{nm},k} = \frac{\omega_{s,k}}{4} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \omega_{s,k}^{-1}} e^{\mathbf{B}^{\dagger} \omega_{s,k}^{-1}} \} \right].
-\eeq
-Note that the factor of $\frac{1}{4}$ is particular to two-field systems. Namely, the factor of $\frac{1}{2}$ coming from Eq.~\ref{E_t_ensemble_avg} is a consequence of the $2 \times 2$ dimensionality
-of $\mathbf{B}_k$.
-
-%The second linear time scale that we may use is the inverse of the linear growth rate. However, the eigenvalues of $\mathbf{B}_k$ have different growth rates (real parts). 
-%Furthermore, we have argued that the eigenmode growth rates have little bearing on turbulence, especially for highly non-normal turbulence, 
-%so it would be innappropriate to use either of the linear eigenmode growth rates as a proxy for the nonlinear time scale.
-%Rather, we use the result itself, $\gamma_{{\rm{nm}},k}$, as the inverse of $\tau_{nl,k}$. Substituting $1/|\gamma_{{\rm{nm}},k}|$ for $\tau_{nl,k}$ in Eq.~\ref{gamma_nm_calc}, we get
-%a transcendental equation:
-%
-%\beq
-%\label{trans_gamma_nm}
-%\gamma_{{\rm{nm}},k} = \frac{|\gamma_{{\rm{nm}},k}|}{2} \ {\rm{Log}} \left< E_k(1/|\gamma_{{\rm{nm}},k}|) \right>_{\rm{ens}}.
-%\eeq
-%We label this second non-modal growth rate $\gamma_{\gamma,k}$, and ultimately simplify this equation to
-%
-%\begin{figure}
-%\centerline{\includegraphics[width=0.52\textwidth]{alpha1e-5_gamma_spec_compare}}
-%\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$,  b) the turbulent growth rate spectrum $\gamma_{t,k}$, c) the non-modal growth rate spectrum $\gamma_{\omega,k}$ d) and the 
-%non-modal growth rate spectrum $\gamma_{\gamma,k}$ for the 2D HW model with $\alpha = 10^{-5}, \kappa=1$. Note the different scales.}
-%\label{alpha1e-5_gamma_spec_compare}
-%\end{figure}
-%
-%\begin{figure}
-%\centerline{\includegraphics[width=0.52\textwidth]{alpha1_gamma_spec_compare}}
-%\caption{a) The eigenmode growth rate spectrum $\gamma_{s,k}$,  b) the turbulent growth rate spectrum $\gamma_{t,k}$, c) the non-modal growth rate spectrum $\gamma_{\omega,k}$ d) and the 
-%non-modal growth rate spectrum $\gamma_{\gamma,k}$ for the 2D HW model with $\alpha = 1, \kappa=1$.}
-%\label{alpha1_gamma_spec_compare}
-%\end{figure}
-%
-%
-%\beq
-%\label{gamma_pm2}
-%\left< E_k(1/|\gamma_{\gamma,k}|) \right>_{\rm{ens}} = e^{\pm 2},
-%\eeq
-%where if $\left< E_k(t) \right>_{\rm{ens}}$ first reaches $e^2$, $\gamma_{\gamma,k}$ is positive, and if it first reaches $e^{-2}$, $\gamma_{\gamma,k}$ is negative.
-%Therefore, finding $\gamma_{\gamma,k}$ amounts to a root-finding problem, which is not trivial, but is much simpler than a nonlinear simulation. 
-%
-%We test both of these non-modal growth rates to see if either or both produce good approximations of $\gamma_{t,k}$, especially in comparison to $\gamma_{s,k}$. 
-%In Fig.~\ref{alpha1e-5_gamma_spec_compare}, we show the $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$ spectra for the 2D HW
-%model for $\alpha=10^{-5}$ along with the eigenmode and turbulent spectra ($\gamma_{s,k}$ and $\gamma_{t,k}$), which we already provided in Fig.~\ref{alpha1e-5_gamma_s_t}. We do the same for
-%$\alpha=1$ in Fig.~\ref{alpha1_gamma_spec_compare}. For the highly non-normal case ($\alpha = 10^{-5}$), $\gamma_{\omega,k}$ reproduces $\gamma_{t,k}$ in both magnitude and spectral shape. 
-%$\gamma_{\gamma,k}$ is not close in any respect to $\gamma_{t,k}$. On the other hand, for the nearly normal case ($\alpha=1$), $\gamma_{\omega,k}$ does not resemble $\gamma_{t,k}$. 
-%$\gamma_{\gamma,k}$ actually is fairly similar to $\gamma_{s,k}$, but with a slightly decreased amplitude, making it a closer match to $\gamma_{t,k}$, though it is still not similar for all $k$.
+Note that the factor of $\frac{1}{4}$ is particular to two-field systems.
+Namely, the factor of $\frac{1}{2}$ coming from Eq.~\ref{E_t_ensemble_avg} is a consequence of $\mathbf{B}_k$ being a rank-$2$ matrix.
 
 \begin{figure}
-\centerline{\includegraphics[width=0.45\textwidth]{gamma_max_with_nm}}
-\caption{Comparison of the non-modal growth rate $\gamma_{\rm{nm},k}$ to the eigenmode $\gamma_{s}$ and turbulent $\gamma_{t}$ growth rates as a function of $\alpha$
-for the 2D HW model. All growth rates are peaks in $k_x-k_y$ space.}
+\centerline{\includegraphics[width=0.4\textwidth]{gamma_max_with_nm_nn}}
+\caption{Comparison of the non-modal growth rates $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$ to the eigenmode $\gamma_{s}$ and turbulent $\gamma_{t}$ growth rates as a function of $\alpha$ for the 2D HW model.
+All growth rates are peaks in $k_x-k_y$ space.}
 \label{gamma_max_with_nm}
@@
 
-The peak non-modal growth rate prediction agrees fairly well with the turbulent growth rate for highly non-normal 2D HW turbulence as seen in Fig.~\ref{gamma_max_with_nm}. For $\alpha \le 10^{-4}$,
-in which the system is fairly non-normal, $\gamma_{\rm{nm},k}$ predicts $\gamma_{t,k}$ quite accurately, especially compared to the eigenmode prediction. 
-As the system becomes more normal, $\gamma_{t,k}$ and $\gamma_{\rm{nm},k}$ diverge, with $\gamma_{t,k}$ taking on values between $\gamma_{s,k}$ and $\gamma_{\rm{nm},k}$, 
-but generally closer to $\gamma_{s,k}$ for $\alpha \gg 10^{-3}$.
+Finally, in order to obtain a simple and quick prediction of $\gamma_{{\rm{nm}},k}$, we must estimate $\tau_{nl,k}$ without having to calculate it directly from a turbulent simulation. 
+We thus invoke the conjecture of \emph{critical balance}, which posits that the nonlinear time scale equals a characteristic linear time scale at all spatial scales~\cite{schekochihin2012}. 
+There are a few possible linear time scales in the HW problem to use for $\tau_{nl,k}$ 
+including the inverse linear eigenmode frequency and growth rates as well as certain times associated with transient growth curves.
 
-This indicates that some assumption in our procedure doesn't work well for systems that are normal or nearly normal. Namely, our assumption
-of complete randomization by the nonlinearities breaks down. The nonlinearities only partially randomize the turbulence in the weakly non-normal case such that
-the turbulence at a given $k$ always consists primarily of the unstable eigenmode. The stable eigenmode never accounts for a significant portion of the energy in the turbulence.
-Makwana et al.~\cite{makwana2011} developed a condition based on the ratio of the stable to unstable eigenmode growth rates that can be used to determine the significance of the stable eigenmode. 
-They determined that for $\alpha \ge 10^{-1}$, the stable eigenmode receives very little energy, explaining why our procedure does not work well in the normal limit for the 2D HW model.
+We choose to test two of these linear time scales -- the inverse linear eigenmode frequency $\omega_{s,k}^{-1}$ and the inverse least stable eigenmode growth rate $\gamma_{s,k}^{-1}$, corresponding to the imaginary and real parts
+of the dominant eigenvalue of the linear matrix operator $\mathbf{B}_k$.
+The reason for using the least stable eigenmode growth rate rather than the most stable growth rate is that the least stable one should contribute more to the nonlinearity at long time after the randomization step.
+As for the eigenmode frequency, recall that the imaginary part of the linear eigenvalues is the same for both eigenvalues of $\mathbf{B}_k$, so there is no question of which eigenfrequency to use. 
+We thus arrive at our final definition for the non-modal growth rates:
+
+\beqar
+\label{gamma_omega_def}
+\gamma_{\omega,k} = \frac{\omega_{s,k}}{4} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \omega_{s,k}^{-1}} e^{\mathbf{B}^{\dagger} \omega_{s,k}^{-1}} \} \right] \\
+\label{gamma_gamma_def}
+\gamma_{\gamma,k} = \frac{\gamma_{s,k}}{4} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \gamma_{s,k}^{-1}} e^{\mathbf{B}^{\dagger} \gamma_{s,k}^{-1}} \} \right].
+\eeqar
+Eqs.~\ref{gamma_omega_def} and~\ref{gamma_gamma_def} are analytic expressions that require no numerical simulation. 
+Furthermore, since the $k$'s have been decoupled by the model, one does not have to perform calculations for many values of $k$, making all calculations extremely fast compared to nonlinear simulations.
 
@@
 \centerline{\includegraphics[width=0.4\textwidth]{gamma_max_vs_kz_with_nm}}
-\caption{Comparison of the non-modal growth rate $\gamma_{\rm{nm},k}$ to the growth rates of the fastest growing eigenmode $\gamma_{s,k}$ 
-and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model with $\xi = 10^3$. The non-modal growth rate correctly predicts the growth rate dominance of the $k_z=0$ mode
+\caption{Comparison of the non-modal growth rates $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$ to the growth rates of the fastest growing eigenmode $\gamma_{s,k}$ 
+and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model with $\xi = 10^3$. The non-modal growth rates correctly predict the growth rate dominance of the $k_z=0$ mode
 despite the $k_z=0$ eigenmode stability.}
@@
 
-We do the same comparison for the 3D HW model as we did for the 2D HW model. However, there is one additional subtelty when considering the 3D model. In the 3D model, for $k_z=0$, the linear
-eigenmode frequency is zero ($\omega_{s,k_z=0} = 0$), so $\gamma_{\rm{nm},k_z=0}$ is not well defined (see Eq.~\ref{gamma_omega_def}). 
-%We could, in fact, make the argument that since $\omega_{s,k_z=0} = 0$, or by critical balance $\tau_{nl,k_z=0} \to \infty$, it would be proper to equate $\gamma_{\rm{nm},k_z=0}$ with $\gamma_{s,k_z=0}$
-%since all transients die away in the long time limit. 
+We test the model against the full direct numerical simulations (DNS) by comparing $\gamma_{\omega,k} \rm{and} \ \gamma_{\gamma,k}$ to $\gamma_{t,k}$. 
+As seen in Fig.~\ref{gamma_max_with_nm}, the peak non-modal growth rate predictions for 2D HW turbulence generally agree within a factor of 2 with the turbulent growth rate. For $\alpha \le 10^{-4}$,
+in which the system is fairly non-normal, $\gamma_{\omega,k} \rm{and} \ \gamma_{\gamma,k}$ predict $\gamma_{t,k}$ quite accurately, especially compared to the eigenmode prediction. 
+This limit, of course, is important in subcritical systems, making the non-modal growth rates quite useful.
+As the system becomes more normal, $\gamma_{t,k}$ diverges from $\gamma_{\omega,k} \rm{and} \ \gamma_{\gamma,k}$, with $\gamma_{\gamma,k}$ doing a better job than $\gamma_{\omega,k}$.
+
+Next, we do the same comparison for the 3D HW model as we did for the 2D HW model.
+However, there is one additional subtelty when considering the 3D model.
+In the 3D model, for $k_z=0$, the least stable linear
+eigenmode is approximately zero ($\omega_{s,k_z=0} = 0, \gamma_{s,k_z=0} \approx 0$), so $\gamma_{\omega,k_z=0}$ is not well defined (see Eq.~\ref{gamma_omega_def}), and $|\gamma_{s,k_z=0}^{-1}|$ is much too large to
+represent a meaningful time scale. 
 Thus, we use $\omega_{s,k_z=0.02}$ -- where $k_z=0.02$ is the lowest finite value of $k_z$ available to our system -- in place of
-$\omega_{s,k_z=0}$ in the calculation of $\gamma_{\rm{nm},k_z=0}$. We contend that the nonlinear instability, which most strongly couples $k_z=0$ and $k_z=0.02$ modes, has its time scale set
-by the drift waves, not the convective cells, making $1/\omega_{s,k_z=0.02}$ a good proxy for $\tau_{nl,k_z=0}$.
+$\omega_{s,k_z=0}$ in the calculation of $\gamma_{\omega,k_z=0}$ (and similarly for $\gamma_{s,k_z=0}$). 
+We contend that the nonlinear instability, which most strongly couples $k_z=0$ and $k_z=0.02$ modes, has its time scale set
+by the drift waves, not the convective cells, making $\omega_{s,k_z=0.02}^{-1}$ and $\gamma_{s,k_z=0.02}^{-1}$ good proxies for $\tau_{nl,k_z=0}$~\cite{friedman2014}.
 
-%One may recall that the drift waves are actually secondary, and therefore, they feed
-%off of the $k_z=0$ density structures, so the appropriate scale length to use for $\omega_{s,k_z=0.02}$ is not given by $\kappa$, but by the scale length of the density structures. We don't deal
-%with this complication and assume the scale length of the fluctuating density structures is the same as the background.
+Despite these complications at $k_z=0$,  $\gamma_{\omega,k} \rm{and} \ \gamma_{\gamma,k}$ approximate $\gamma_{t,k}$ remarkably well, even when the linear matrix is only moderately ($1-H \approx 10^{-1}$) non-normal
+(see Figs.~\ref{gamma_max_vs_kz_with_nm} and~\ref{gamma_max_vs_xi}).
+And again, $\gamma_{\gamma,k}$ is better at higher $\xi$ than $\gamma_{\omega,k}$, though here $\gamma_{\omega,k}$ gives the better prediction
+at some intermediate values of $\xi$. Most importantly, however, they both correctly predict that $\gamma_{t,k_z=0} > 0$.
+Therefore, non-modal calculations explain why the 3D HW model can be dominated by the nonlinear instability rather than a linear drift wave instability. 
+Simply put, the transient growth rate may be largest at $k_z=0$ so the nonlinear instability is energetically favored over the primary drift wave instability.
 
@@
 \centerline{\includegraphics[width=0.52\textwidth]{gamma_max_vs_xi}}
-\caption{Comparison of the peak non-modal growth rate $\gamma_{\rm{nm},k}$ to that of the fastest growing eigenmode $\gamma_{s,k}$ 
+\caption{Comparison of the peak non-modal growth rates $\gamma_{\omega,k} \ $ and $\gamma_{\gamma,k}$ to that of the fastest growing eigenmode $\gamma_{s,k}$ 
 and the turbulence $\gamma_{t,k}$ as a function of $\xi$ for the 3D HW model at {\bf a)} $k_z \rho_s = 0$ and {\bf b)} $k_z \rho_s \approx 0.02$. }
@@
 \end{figure}
-
-Despite these complications at $k_z=0$, our non-modal method of calculating $\gamma_{\rm{nm},k}$ approximates $\gamma_{t,k}$ remarkably well as seen in Figs.~\ref{gamma_max_vs_kz_with_nm}
-and~\ref{gamma_max_vs_xi}. Most importantly, it correctly predicts that $\gamma_{t,k_z=0} > 0$. 
-Therefore, non-modal calculations explain why the 3D HW model can be dominated by the nonlinear instability rather than a linear drift wave instability. 
-Simply put, the transient growth rate is by far the largest at $k_z=0$ for $\xi \gg 10^2$, so the nonlinear instability is energetically favored over the primary drift wave instability.
-For $\xi \le 10^2$, the $k_z=0$ nonlinear instability has comparable growth rate to the drift wave instability, so neither dominates.
 
@@
 In both hydrodynamic and plasma systems, it is important to be able to predict for a given set of parameters whether the system will be turbulent or not.
-Put another way, it is useful to map out the point, line, or manifold in parameter space that divides turbulent and laminar solutions.
+Ideally, one would like to find the point, line, or manifold in parameter space that divides turbulent and laminar solutions.
 In many hydrodynamic systems with sheared flow, there is generally a single control parameter -- the Reynold's number $Re$~\cite{drazin1981}. For high enough $Re$ such systems
@@
 In general, this is not true for non-normal systems. In many non-normal systems, turbulence can be sustained when all linear eigenmodes are stable -- the case of subcritical turbulence.
+
+Subcritical systems fall into two categories: those that have stable eigenmodes for all values of the control parameter, and those that have unstable eigenmodes
+when the control parameter becomes large (or small) enough. We illustrate these two categories with a diagram in Fig.~\ref{subcritical_diagram}. 
+The upper curve represents the first kind of subcritical system in which no unstable eigenmodes exist for any value of the control parameter $R$. In this
+case, there is some transitional value of $R$, labeled $R_g$, below which turbulence cannot occur.
+When a perturbation with amplitude above the line occurs in the system due to external input or noise,
+turbulence will be excited. Perturbations with amplitude below the line will decay away. The higher the value of $R$, the smaller the perturbation needs to be to excite turbulence.
+Experimental evidence and theoretical arguments show this curve should have a $(R-R_g)^{-\lambda}$ dependence (for asymptotically large $R$), with $\lambda \sim 1$~\cite{waleffe1995b,grossmann2000,hof2003}.
+A paradigmatic example of this kind of system is Hagen-Poisuille pipe flow -- flow in a cylindrical pipe with a parabolic flow profile -- for which $R_g \approx 2000$.
+
+The lower curve represents the second kind of subcritical system. Above a critical value of the control parameter $R_c$, the system is linearly unstable. But, for $R_g < R < R_c$, the system is
+unstable to finite amplitude perturbations, or ``nonlinearly'' unstable. A paradigmatic example of this is Poissuille channel flow -- the rectangular analog of pipe flow -- for which
+$R_c = 5772$ and $R_g \approx 1000$~\cite{grossmann2000}.
+Most subcritical magnetically-confined plasma systems fall into this category.
+It is perhaps easy to be fooled into believing that the turbulence
+in such systems is due to eigenmode instability even though it may not be. 
+In either case, one of the most important predictions is the minimum turbulent control parameter $R_g$, which is sometimes called the ``transitional'' value of the control parameter.
+Given that this problem is many decades old
+and still unsolved, it is a tall order to predict $R_g$ without nonlinear simulations. We nevertheless make an attempt using our procedure for calculating $\gamma_{\rm{nm}}$.
+
+The HW model in an unsheared magnetic slab is not subcritical in either the 2D or 3D cases, as implied by Figs.~\ref{gamma_max_vs_alpha} and~\ref{gamma_max_vs_kz}.
+It is subcritical in a highly sheared
+magnetic field~\cite{drake1995}, but the eigenmodes of the linear operator in that case are not sinusoidal, making our non-modal analysis more difficult. 
+Biskamp et al.~\cite{biskamp1995}, however, artificially modified the
+3D HW model to remove the linear drive by taking the $z$-average of the $\kappa \pdiff{\phi}{y}$ term in Eq.~\ref{n_eq}, effectively eliminating the $k_z \ne 0$ drive. This leaves only the $k_z = 0$
+component of the drive term. Since the $k_z = 0$ eigenmodes are stable but $k_z = 0$ structures can access the free energy (positive $\gamma_t$ in Fig.~\ref{gamma_max_vs_kz}), this system is subject
+to subcritical turbulence.
 
@@
 
-Subcritical systems fall into two categories: those that have stable eigenmodes for all values of the control parameter, and those that have unstable eigenmodes
-when the control parameter becomes large (or small) enough. We illustrate these two categories with a diagram in Fig.~\ref{subcritical_diagram}. 
-The upper curve represents the first kind of subcritical system in which no unstable eigenmodes exist for any value of the control parameter $R$. In this
-case, there is some transitional value of $R$, labeled $R_g$, below which turbulence cannot occur.
-When a perturbation with amplitude above the line occurs in the system due to external input or noise,
-turbulence will be excited. Perturbations with amplitude below the line will decay away. Clearly, the higher the value of $R$, the smaller the perturbation needs to be to excite turbulence.
-Experimental evidence and theoretical arguments show this curve should have a $(R-R_g)^{-\lambda}$ dependence, with $\lambda \sim 1$~\cite{grossmann2000}.
-A paradigmatic example of this kind of system is Hagen-Poisuille pipe flow -- flow in a cylindrical pipe with a parabolic flow profile -- for which $R_g \approx 2000$.
-
-The lower curve represents the second kind of subcritical system. Above a critical value of the control parameter $R_c$, the system is linearly unstable. But, for $R_g < R < R_c$, the system is
-unstable to finite amplitude perturbations, or ``nonlinearly'' unstable. A paradigmatic example of this is Poissuille channel flow -- the rectangular analog of pipe flow -- for which
-$R_c = 5772$ and $R_g \approx 1000$~\cite{grossmann2000}.
-Most subcritical magnetically-confined plasma systems fall into this category.
-It is perhaps easy to be fooled into believing that the turbulence
-in such systems is due to eigenmode instability even though it may not be. 
-In either case, one of the most important predictions is the minimum turbulent control parameter $R_g$, which is sometimes called the ``transitional'' value of the control parameter.
-Given that this problem is many decades old
-and still unsolved, it is a tall order to predict $R_g$ without nonlinear simulations. We nevertheless make an attempt.
-
-The HW model in an unsheared magnetic slab is not subcritical in either the 2D or 3D cases, as implied by Figs.~\ref{gamma_max_vs_alpha} and~\ref{gamma_max_vs_kz}.
-It is subcritical in a highly sheared
-magnetic field~\cite{drake1995}, but the eigenmodes of the linear operator in that case are not sinusoidal, making our non-modal analysis more difficult. 
-Biskamp et al.~\cite{biskamp1995}, however, artificially modified the
-3D HW model to remove the linear drive by taking the $z$ average of the $\kappa \pdiff{\phi}{y}$ term in Eq.~\ref{n_eq}, effectively eliminating the $k_z \ne 0$ drive. This leaves only the $k_z = 0$
-component of the drive term. Since the $k_z = 0$ eigenmodes are stable but $k_z = 0$ structures can access the free energy (positive $\gamma_t$ in Fig.~\ref{gamma_max_vs_kz}), this system is subject
-to subcritical turbulence.
-
 We show that the system is subcritical in Fig.~\ref{crit_amp_dependence}.
-In Fig.~\ref{crit_amp_dependence} a), we plot the energy versus time for fluctuations that are initialized with different amplitudes.
+In Fig.~\ref{crit_amp_dependence} a), we plot the energy versus time for fluctuations that are initialized with the same functional form, but with
+different amplitudes.
 The fluctuations initialized with high enough amplitude lead to turbulence. 
-Otherwise, the fluctuations grow transiently, but do not reach high enough amplitude for the nonlinearities to bootstrap the process,
-so they decay away (we set the diffusion and viscosity coefficient $D$ to $0.5$ in this section so that the decay occurs on a short enough time scale to keep the simulations short). 
+Otherwise, the fluctuations grow transiently, but do not reach high enough amplitude for the nonlinearities to bootstrap the process, so they decay away. 
 This finite amplitude turbulence threashold, as described above, is a key indicator of subcritical turbulence.
 
-In Fig.~\ref{crit_amp_dependence} b), we show the amplitude threashold curve -- like that of Fig.~\ref{subcritical_diagram} -- for the subcritical 3D HW model.
-We obtain the points from nonlinear simulations
-and fit them to a function of the form $A = a (\kappa - \kappa_g)^{- \lambda}$, where $A$ is the lowest initial energy amplitude that leads to turbulence for each value of $\kappa$.
-The least-squares fitting provides optimal values
-for the parameters $a, \kappa_g, \rm{and} \lambda$. As expected for shear flows, $\lambda \sim 1$.
+In Fig.~\ref{crit_amp_dependence} b), we show the amplitude threashold curve -- like that of Fig.~\ref{subcritical_diagram}.
+Each point represents the minimum amplitude that is needed for the simulation to eventually become
+turbulent. We fit the points to a function of the form $A = a (\kappa - \kappa_g)^{- \lambda}$, 
+where $A$ is the lowest initial energy amplitude that leads to turbulence for each value of $\kappa$.
+The least-squares fitting provides optimal values for the parameters $a, \kappa_g, \rm{and} \ \lambda$.
+As in the case of shear flows, $\lambda \sim 1$.
+Note that this power law exponent shouldn't be taken too seriously because we have not used ``optimal'' initial perturbations to get the points~\cite{cossu2005}.
+In any case, this is not our main concern; we are simply demonstrating that this modified 3D HW model is subcritical.
 
@@
 The points are obtained from nonlinear simulations like those in part a) with each point indicating the minimum amplitude needed for the simulation
-to become turbulent. The line is a fit to the points with the functional form $a (\kappa-\kappa_g)^{- \lambda}$.}
+to become turbulent. The line is a fit to the points with the functional form $a (\kappa-\kappa_g)^{- \lambda}$.
+We set the diffusion and viscosity coefficient $D$ to $0.5$ in these figures so that the decay occurs on a short enough time scale to be able to see the results.}
 \label{crit_amp_dependence}
@@
 
-In this subcritical system, we predict the line in $\xi-\kappa$ parameter space that divides the space into an area that permits turbulent solutions and an area that does not. That is, we predict
-$\kappa_g(\xi)$.
-We show the answer that we obtain from nonlinear simulations along with our non-modal prediction in Fig.~\ref{subcrit_param_space}.
-To obtain $\kappa_g(\xi)$ from the nonlinear simulations,
+For this subcritical system, we predict the transitional dividing line $\kappa_g(\xi)$ that separates the parameter space into turbulent and laminar sections. In the turbulent section, 
+fluctuations initialized with high enough amplitude can drive turbulence. In the laminar section, fluctuations will all eventually decay no matter how large they are initialized. 
+In the turbulent section, it is not necessarily the case that the system will look turbulent in the usual sense, at least not strongly turbulent.
+So more accurately, in the turbulent section, the system will undergo some sort of bifurcation that brings with it sustained finite transport.
+
+We show the answer that we obtain from nonlinear simulations along with our non-modal predictions in Fig.~\ref{subcrit_param_space}. To obtain $\kappa_g(\xi)$ from the nonlinear simulations,
 rather than mapping out curves like those in Fig.~\ref{crit_amp_dependence} b), we simply start all simulations with a large initial amplitude ($A = 10^4$) and vary $\kappa$ until we find the minimum $\kappa$ that
-sustains turbulence (see Fig.~\ref{kappa_en_ev} for an example).
-
-Our method
-of prediction is to find the line in $\xi-\kappa$ space at which $\gamma_{\rm{nm}} = 0$, where $\gamma_{\rm{nm}}$ is maximized over $k_x$ and $k_y$. We plot this line in Fig.~\ref{subcrit_param_space}
+permits a sustained positive value of the turbulent growth rate $\gamma_{t,k}$.
+Similarly, we find the lines (with a root-finding routine) in $\xi-\kappa$ space at which $\gamma_{\omega} = 0$ and $\gamma_{\gamma} = 0$, using the lowest permissible values of $k_x$ and $k_y$, which maximize the growth rates.
+We plot these lines in Fig.~\ref{subcrit_param_space}.
 
@@
 \centerline{\includegraphics[width=0.4\textwidth]{subcrit_param_space}}
-\caption{The non-modal curve $\gamma_{\rm{nm}}(\kappa,\xi) = 0$, which defines our prediction of the transitional value of $\kappa(\xi)$ for the subcritical 3D HW model. The points represent the
+\caption{The non-modal curves $\gamma_{\omega}(\kappa,\xi) = 0$ and $\gamma_{\gamma}(\kappa,\xi) = 0$ predict the transitional value of $\kappa(\xi)$ for the subcritical 3D HW model.
+The points represent the
 transitional values of $\kappa(\xi)$ based on full nonlinear simulations.}
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=0.52\textwidth]{kappa_en_ev}}
-\caption{$E_{\rm{tot}}(t)$ from full nonlinear simulations of the subcritical 3D HW model. The curves that reach a quasi-steady-state value represent sustained subcritically turbulent states. 
-Those that decay for large time have values of $\kappa$ too small to sustain turbulence. The transitional value $\kappa_g$ is the smallest value of $\kappa$ for which turbulence is sustained.}
-\label{kappa_en_ev}
-\end{figure}
+Clearly, our non-modal procedure with our particular choice of linear time scales does not provide even an order-of-magnitude accurate prediction of $\kappa_g$ for all values of $\xi$.
+Nevertheless, the non-modal predictions display some qualitative agreement with the DNS results in that the non-modal predictions trend fairly accurately with $\xi$. 
+And they do provide a prediction to the transition density gradient $\kappa_g$ in contrast to the eigenmode growth rates, which are always negative in this system and thus do not predict turbulence for any $\kappa$.
+So, there is some value in the non-modal predictions, especially in light of the fact that it took over $10^5$ computer hours to calculate the $\kappa_g$ line using DNS and only seconds using the non-modal procedure.
+Furthermore, it is likely that more sophisticated models for the nonlinearities will improve the predictions in the future.
+
+\section{Conclusion}
+
+When the nonlinear terms in the model equations are energetically conservative, as they are in the HW model, the linear terms must on average inject energy into at least one wavenumber to sustain turbulent fluctuations.
+This is manifest in the positivity of the maximum turbulent growth rate, which is defined in terms of only the linear terms in the energetic model equations.
+Thus, an estimation of the turbulent growth rate provides vital
+information into the nature of the system. Furthermore, such an estimate almost surely requires non-modal techniques when the system is non-normal.
+
+Our method for analytically esimating this growth rate involves using a type of closure, which means modelling the effect that the nonlinearities have
+on the time-averaged complex ratio of the Fourier-decomposed state variables. 
+We present one specific phenomenological closure, in which the nonlinearities instantaneously randomize the state variables once every eddy turnover time, between
+which, the linear terms deterministically evolve them. We then estimate the eddy turnover time as the inverse linear eigenmode frequency or growth rate.
+
+While our nonlinear closure model is successful in some instances, such as the highly non-normal limit in the 2D HW model and in the 3D HW model when the gradient scale length factor $\kappa$ is large, it is relatively unsuccessful
+in others, especially when the system is not highly non-normal. This prevents an accurate prediction of the transitional value $\kappa_g$ that marks where the system first becomes turbulent.
+Perhaps in the future, more advanced closure techniques will allow for better approximation of the turbulent growth rate.
 

### 2014-08-27 HW_non-modal.tex
Aug 27

@@
 All calculations in this section use $\kappa=1$.
-Note the different scales used in all of the figures. In the highly
-non-normal case, the turbulent growth rate is somewhat higher than the eigenmode growth rate for most values of $k_x,k_y$, and about 6 times higher where the spectrum peaks. For the more normal
+In the highly non-normal case, the turbulent growth rate is very similar to the eigenmode growth rate for most values of $k_x,k_y$, but about 6 times higher where the spectrum peaks.
+For the more normal
 case, in contrast, $\gamma_{t,k}$ is generally lower than $\gamma_{s,k}$, with the peak at nearly a factor of 2 less, with the peaks being at slightly different values of $k_y$.
@@
 \centerline{\includegraphics[width=0.52\textwidth]{alpha1e-5_gamma_s_t}}
-\caption{{\bf a)} The eigenmode growth rate spectrum $\gamma_{s,k}$ and {\bf b)} the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 10^{-5}, \kappa=1$. Note the different scales.}
+\caption{{\bf a)} The eigenmode growth rate spectrum $\gamma_{s,k}$ and {\bf b)} the turbulent growth rate spectrum $\gamma_{t,k}$ for $\alpha = 10^{-5}, \kappa=1$.}
 \label{alpha1e-5_gamma_s_t}
@@
 
-We show the peak values for $\gamma_{s,k}$ and $\gamma_{t,k}$ in Fig.~\ref{gamma_max_vs_alpha} as a function of $\alpha$. The peak values always occur in $k$-space
-at the lowest $k_x$ available to our system, corresponding to half of a period of a sine wave. The value of $k_y$ that gives the highest growth rate migrates upward as $\alpha$ increases,
-as is clear from Figs.~\ref{alpha1e-5_gamma_s_t} and~\ref{alpha1_gamma_s_t}. Furthermore, as expected, when $\alpha$ is small ($\lesssim 10^{-3}$), $\gamma_{t,k} > \gamma_{s,k}$, with
-$\gamma_{t,k} \gg \gamma_{s,k}$ as $\alpha \to 0$. Additionally, when $\alpha$ becomes larger and the system becomes more normal ($\gtrsim 10^{-1}$), $\gamma_{t,k} < \gamma_{s,k}$.
+Since the peak growth rates are the most important when considering whether a system can sustain turbulent, we show the peak values for $\gamma_{s,k}$ and $\gamma_{t,k}$ in Fig.~\ref{gamma_max_vs_alpha} as a function of $\alpha$.
+The peak values always occur in $k$-space at the lowest $k_x$ available to our system, corresponding to half of a period of a sine wave.
+The value of $k_y$ that gives the highest growth rate migrates upward as $\alpha$ increases,
+as is clear from Figs.~\ref{alpha1e-5_gamma_s_t} and~\ref{alpha1_gamma_s_t}. Henceforth, we discuss only the peak growth rates.
 
-For both the normal and highly non-normal limits of the 2D HW model, $\gamma_{s,k}$ does a poor job of predicting $\gamma_{t,k}$ because $\gamma_{s,k}$ neglects the stable branch of the
+As expected, when $\alpha$ is small ($\lesssim 10^{-3}$), $\gamma_{t} > \gamma_{s}$, with
+$\gamma_{t} \gg \gamma_{s}$ as $\alpha \to 0$. Additionally, when $\alpha$ becomes larger and the system becomes more normal ($\gtrsim 10^{-1}$), $\gamma_{t} < \gamma_{s}$.
+For both the normal and highly non-normal limits of the 2D HW model, $\gamma_{s}$ does a poor job of predicting $\gamma_{t}$ because $\gamma_{s}$ neglects the stable branch of the
 dispersion relation -- the damped eigenmode.
-Additionally, in the moderate normality region where $\gamma_{s,k}$ does match $\gamma_{t,k}$ well, it is misleading to think of the system as being composed of only the
+Additionally, in the moderate normality region where $\gamma_{s}$ does match $\gamma_{t}$ well, it is misleading to think of the system as being composed of only the
 unstable eigenmode.
 The damped mode simply has both damping and transient growth effects depending on its relative magnitude at a particular time.
-These effects nearly cancel out on average when $\gamma_{t,k} \approx \gamma_{s,k}$.
+These effects nearly cancel out on average when $\gamma_{t} \approx \gamma_{s}$.
 
@@
 The importance of the stable eigenmode branch~\cite{baver2002} (or branches) has started to gain acceptance in the plasma community, although it has generally been used in normal
-or nearly normal systems to explain turbulent saturation~\cite{terry2006b,hatch2011,makwana2011} ($\gamma_{t,k} < \gamma_{s,k}$). 
+or nearly normal systems to explain turbulent saturation~\cite{terry2006b,hatch2011,makwana2011} ($\gamma_{t} < \gamma_{s}$). 
 The case of highly non-normal turbulence and the associated increased turbulent excitation hasn't been understood as well. 

### 2014-11-20 HW_non-modal.tex
Nov 20

@@
 \author{B. Friedman}
-\email{friedman@physics.ucla.edu}
+\email{friedman11@llnl.gov}
 
@@
 \begin{abstract}
-Linear eigenmode analysis fails to account for turbulence in subcritical systems in many cases because such systems have non-normal linear operators, and thus
+Linear eigenmode analysis often fails to describe turbulence in model
+systems that have non-normal linear operators, and thus
 nonorthogonal eigenmodes, which can cause fluctuations to transiently grow faster than expected from eigenmode analysis. When combined with energetically conservative nonlinear mode mixing, 
@@
 Since linear operators ultimately provide the turbulent fluctuations with energy, it is useful to define a growth rate that takes into account non-modal effects, allowing for
-prediction of energy injection and transport levels, and even turbulent onset in the subcritical regime. 
-We define such a non-modal growth rate by modelling the statistical effect that the nonlinearities have on cross-phases and amplitude ratios of the system state variables. 
+prediction of energy injection, transport levels, and possibly even turbulent onset in the subcritical regime. 
+We define such a non-modal growth rate using a relatively simple model
+of the statistical effect that the nonlinearities have on cross-phases and amplitude ratios of the system state variables. 
 In particular, we model the nonlinearities as delta-function-like, periodic forces that randomize the state variables once every eddy turnover time. Furthermore, we estimate the eddy turnover
@@
 turbulence is called subcritical~\cite{drazin1981}. 
-In subcritical systems, turbulence develops even though all linear eigenvectors are stable such that infinitesimal perturbations on the laminar state cannot grow exponentially. 
+In subcritical systems, turbulence develops even though all linear eigenvectors are stable such that infinitesimal perturbations on the equilibrium state cannot grow exponentially. 
 Nevertheless, finite amplitude perturbations still excite turbulence through nonlinear instabilities.
@@
 superpositions of eigenvectors can grow, albeit transiently.
-In other words, certain fluctuations of the laminar state can access free energy from background gradients even though normal mode fluctuations cannot.
+In other words, certain fluctuation structures can access free energy from background gradients even though normal mode fluctuations cannot.
 When combined with nonlinear effects, this allows for sustained subcritical turbulence by a bootstrapping process.
@@
 
-Non-modal analysis generally employs tools such as pseudospectra and maximum growth curves, which have been used to successfully explain why turbulence is subcritical in various instances. 
+Non-modal analysis generally employs tools such as pseudospectra and maximum growth curves, which have been used to successfully explain
+why turbulence is subcritical in various instances~\cite{trefethen1993,camargo1998}. 
 However, they have not been used to predict onset criteria in subcritical systems (e.g. the transition Reynold's number in pipe flows).
@@
 to give energy injection rates, turbulent saturation levels, and the turbulent onset criteria.
-That is, we argue that a finite positive non-modal growth rate indicates that turbulence can exist, given, of course a sufficient perturbative kick.
+That is, we argue that a finite positive non-modal growth rate
+indicates that turbulence can exist given a sufficiently large perturbative kick.
 
@@
 \eeqar
-The normalizations are $\phi \to e \phi/T_e, n \to n/n_0, \grad \to \rho_s \grad , x,y,z \to x,y,z/\rho_s, t \to \omega_{ci} t $. Additionally,
-$\mathbf{v_E} = \mathbf{b} \times \grad \phi$, $\xi = 4 \pi \omega_{ce}/\nu_{ei}$, 
+The normalizations are $\phi \to e \phi/T_e, n \to n/n_0, \grad \to \rho_s \grad , x,y,z \to x,y,z/\rho_s, t \to \omega_{ci} t $, where $\rho_s$ is the ion sound gyroradius at the electron temperature and $\omega_{ci}$ is
+the ion cyclotron frequency.
+Additionally, $\mathbf{v_E} = \mathbf{b} \times \grad \phi$, where $\mathbf{b}$ is the magnetic field direction, $\xi = 4 \pi \omega_{ce}/\nu_{ei}$, where $\omega_{ce}$ is the electron cyclotron
+frequency and $\nu_{ei}$ is the electron-ion collision frequency,
 and $\kappa = \rho_s/L_n$, where $L_n$ is the density scale length. The terms $D \gradperp^4 n$ and $D \gradperp^6 \phi$ are artificial hyper-diffusion and hyper-viscosity used to dissipate
 energy on small scales. We set $D \to 10^{-5}$ except in Sec.~\ref{sec_subcrit_prediction} where we set it to $0.5$.
-We associate the direction along $\mathbf{B}$ as the parallel or $z$ direction. The direction perpendicular $\mathbf{B}$ that
+We associate the direction $\mathbf{b}$ as the $z$ direction. The direction perpendicular $\mathbf{b}$ that
 points opposite the density gradient is defined as the $x$ direction, and the $y$ direction is taken perpendicular to both of these.
@@
 which is commonly known as the adiabaticity parameter~\cite{camargo1995,camargo1998}. Additionally we use a domain size of $50 \rho_s \times 50 \rho_s$.
-In numerical simulations, the turbulence can laminarize due to the build-up of $k_y=0$ density and potential structures~\cite{biskamp1995}. To prevent this, we subtract out the $k_y=0$
+In numerical simulations, the turbulence can be suppressed due to the build-up of $k_y=0$ density and potential structures~\cite{biskamp1995}. To prevent this, we subtract out the $k_y=0$
 density and potential components, which is the equivalent of adding source terms to Eqs.~\ref{n_eq} and~\ref{phi_eq}~\cite{friedman2012b}.
@@
 
-Non-normality manifests itself in its eigenvectors -- non-normal matrices or operators have nonorthogonal linear eigenvectors.
+Non-normality manifests itself in the eigenvectors -- non-normal matrices or operators have nonorthogonal linear eigenvectors.
 For the HW model, each Fourier matrix $\mathbf{B}_k$ has two eigenvectors associated
@@
 
-It is the linear rate of energy injection into the fluctuations $\gamma_{l,k}$ that we wish to use as the turbulent growth rate $\gamma_{t,k}$ because $\gamma_{l.k}$
-determines how fast and where in wavenumber space energy is injected into the turbulence.
+It is the linear rate of energy injection into the fluctuations
+$\gamma_{l,k}$ that we wish to use as the turbulent growth rate
+$\gamma_{t,k}$ because $\gamma_{l,k}$ determines how fast and where in wavenumber space energy is injected into the turbulence.
 $\gamma_{l,k}$, in other words, is the turbulence equivalent to the linear eigenmode growth rate $\gamma_{s,k}$~\cite{friedman2012b,terry2006b}. 
@@
 
-The turbulent growth rate $\gamma_{t,k}$ is an informative quantity. 
+The turbulent growth rate $\gamma_{t,k}$ is a useful and informative quantity. 
 Note that in Eq.~\ref{Explicit_En_eqn} all of the linear terms on the right hand side of the equation are negative, except for the first term, which can have either sign. 
@@
 
-Since the peak growth rates are the most important when considering whether a system can sustain turbulent, we show the peak values for $\gamma_{s,k}$ and $\gamma_{t,k}$ in Fig.~\ref{gamma_max_vs_alpha} as a function of $\alpha$.
+Since the peak growth rates are the most important when considering whether a system can sustain turbulence, we show the peak values for $\gamma_{s,k}$ and $\gamma_{t,k}$ in Fig.~\ref{gamma_max_vs_alpha} as a function of $\alpha$.
 The peak values always occur in $k$-space at the lowest $k_x$ available to our system, corresponding to half of a period of a sine wave.
@@
 For both the normal and highly non-normal limits of the 2D HW model, $\gamma_{s}$ does a poor job of predicting $\gamma_{t}$ because $\gamma_{s}$ neglects the stable branch of the
-dispersion relation -- the damped eigenmode.
+dispersion relation -- the damped eigenmode~\cite{makwana2011}. 
 Additionally, in the moderate normality region where $\gamma_{s}$ does match $\gamma_{t}$ well, it is misleading to think of the system as being composed of only the
@@
 The upper curve represents the first kind of subcritical system in which no unstable eigenmodes exist for any value of the control parameter $R$. In this
-case, there is some transitional value of $R$, labeled $R_g$, below which turbulence cannot occur.
-When a perturbation with amplitude above the line occurs in the system due to external input or noise,
+case, there is some transitional value of $R$, labeled $R_g$, below
+which turbulence cannot occur. When a perturbation with amplitude
+($A$) above the line occurs in the system due to external input or noise, 
 turbulence will be excited. Perturbations with amplitude below the line will decay away. The higher the value of $R$, the smaller the perturbation needs to be to excite turbulence.
@@
 
-We show that the system is subcritical in Fig.~\ref{crit_amp_dependence}.
+We show that the modifed 3D HW system is subcritical in Fig.~\ref{crit_amp_dependence}.
 In Fig.~\ref{crit_amp_dependence} a), we plot the energy versus time for fluctuations that are initialized with the same functional form, but with
@@
 Otherwise, the fluctuations grow transiently, but do not reach high enough amplitude for the nonlinearities to bootstrap the process, so they decay away. 
-This finite amplitude turbulence threashold, as described above, is a key indicator of subcritical turbulence.
+This finite amplitude turbulence threshold, as described above, is a key indicator of subcritical turbulence.
 

### 2014-12-24 Referee_Responses/referee_response.tex
Dec 23

\documentclass[12pt]{article}
\usepackage{graphicx}
\setlength{\parindent}{0in}

% General Latex  --------------------------------------------------
\def\beq{\begin{equation}}
\def\eeq{\end{equation}}
\def\beqar{\begin{eqnarray}}
\def\eeqar{\end{eqnarray}}
\def\nn{\nonumber}
\def\ol{\overline}
\def\para{\parallel}

% Operators  ------------------------------------------------------
\newcommand{\diff}[2]{\frac{d#1}{d#2}}
\newcommand{\diffs}[2]{\frac{d^2#1}{d#2^2}}
\newcommand{\pdiff}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\pdiffs}[2]{\frac{\partial^2#1}{\partial#2^2}}
\newcommand{\pdiffxy}[3]{\frac{\partial^2#1}{\partial#2 \partial#3}}
\newcommand{\pdt}{\partial_t}
\newcommand{\pdr}{\partial_r}
\newcommand{\pdth}{\partial_\theta}
\newcommand{\pdrr}{\partial^2_r}

\newcommand{\enum}[2]{{#1}\times10^{#2}} % 4.2x10^{3} = \enum{4.2}{3}

\newcommand{\vect}[1]{{\bf #1}}
%\newcommand{\vect}{\overrightarrow}
%\newcommand{\vect}{\vec}
\def\div{\nabla\cdot}
\def\grad{\nabla}
\def\curl{\nabla\times}
\newcommand{\gradpar}{\grad_\parallel}
\newcommand{\gradperp}{\grad_\perp}
\newcommand{\gradr}{\grad_r}
\newcommand{\defeq}{\ensuremath{\stackrel{\text{\tiny def}}{=}}}

\newcommand{\savg}[1]{\left<{#1}\right>}
\newcommand{\vavg}[1]{\left<{#1}\right>_V}
\newcommand{\thavg}[1]{\left<{#1}\right>_\theta}

% Variable names  -------------------------------------------------
\newcommand{\vpar} {v_\parallel}
\newcommand{\Apar} {A_\parallel}
\newcommand{\jpar} {j_\parallel}
\newcommand{\kpar} {k_\parallel}
\newcommand{\kperp} {k_\perp }
\newcommand{\vperp} {v_\perp }
\newcommand{\kthe}{k_\theta}

\newcommand{\Evec}{\ensuremath{\boldsymbol{{\rm E}}}}
\newcommand{\Bvec}{\ensuremath{\boldsymbol{{\rm B}}}}
\newcommand{\Jvec}{\ensuremath{\boldsymbol{{\rm J}}}}
\newcommand{\Fvec}{\ensuremath{\boldsymbol{{\rm F}}}}
\newcommand{\fvec}{\ensuremath{\boldsymbol{{\rm f}}}}
\newcommand{\vE}{\ensuremath{\boldsymbol{{\rm v}_{E}}}}
\newcommand{\bo}{\ensuremath{\boldsymbol{{\rm b}_0}}}
\newcommand{\bvec}{\ensuremath{\boldsymbol{{\rm b}}}}
\newcommand{\xvec}{\ensuremath{\boldsymbol{{\rm x}}}}
\newcommand{\yvec}{\ensuremath{\boldsymbol{{\rm y}}}}
\newcommand{\zvec}{\ensuremath{\boldsymbol{{\rm z}}}}
\newcommand{\vvec}{\ensuremath{\boldsymbol{{\rm v}}}}
\newcommand{\jvec}{\ensuremath{\boldsymbol{{\rm j}}}}

\newcommand{\bxgp}{\bvec\times\gradperp}

\newcommand{\vve}{\ensuremath{\boldsymbol{{\rm v}}_{e}}}
\newcommand{\vvi}{\ensuremath{\boldsymbol{{\rm v}}_{i}}}
\newcommand{\vpe}{v_{\parallel e}}
\newcommand{\vpi}{v_{\parallel i}}
\newcommand{\vvE}{\ensuremath{\boldsymbol{{\rm v}}_{E}}}
\newcommand{\vvD}{\ensuremath{\boldsymbol{{\rm v}}_{D}}}

\newcommand{\nuei}{\nu_{ei}}
\newcommand{\nuii}{\nu_{ii}}
\newcommand{\nue}{\nu_{e}}
\newcommand{\nuen}{\nu_{en}}
\newcommand{\nuin}{\nu_{in}}
\newcommand{\kpe}{\kappa_{\parallel e}}

\newcommand{\rs}{\rho_{s}}
\newcommand{\ri}{\rho_{i}}
\newcommand{\wci}{\Omega_{i}}
\newcommand{\wcix}{\Omega_{ix}}
\newcommand{\wce}{\Omega_{e}}
\newcommand{\tomega}{\tilde\omega}
\newcommand{\Isat}{I_{\rm sat}}
\newcommand{\fmie}{\frac{m_i}{m_e}}
\newcommand{\fmei}{\frac{m_e}{m_i}}

% Often used dimensions
\newcommand{\cm}{\rm cm}
\newcommand{\mm}{\rm mm}
\newcommand{\cmn}{{\rm cm}^{-3}}
\newcommand{\mn}{{\rm m}^{-3}}
\newcommand{\eV}{\rm eV}
\newcommand{\G}{\rm G}
\newcommand{\T}{\rm T}

\begin{document}

{\bf Response to Referees \\
 \\} 
Brett Friedman and Troy Carter
%Updated: \today \\

\hrulefill

We would like to thank the referee for his/her thorough reading of our paper and clear and insightful comments. We have responded to each comment below and made requisite changes to the text.
Most of the changes are relatively minor and involve corrections pointed out by the referee, changes in wording, and further clarification on certain points.

The referee's comment 14 is a rather significant one, suggesting the possible removal of Section V of the paper. 
We agree with the points but don't want to go so far as to remove the entire section because even though the result is weak (unsuccessful), 
we believe that it is still valuable and provides some important insight. What we have done, then, is move it to an appendix. 
We have also title, abstract, and introduction to de-emphasize that part of the paper. \\ \\

\emph{1) page 2, second column: $N_k$ is termed 'the nonlinear matrix'. Since matrices are generally associated to linear operator, it would be more appropriate to call it nonlinear operator. } \\

This is a good point. We have made this change. \\

\emph{2) page 3, Eq. 13. This is defined as an energy. However, it is a normalized (adimensional) quantity, and as such it looks more like an energy amplification factor.}\\

This is correct. We did not do a good job with our explanation. Really, we always set the initial amplitude to 1 so that the energy and the energy amplification are synonymous.
We have added this explanation in the text and changed Eq. 13. \\

\emph{3) page 3, 2nd column. The argument about non-normality of an operator implicitly assumes the choice of a norm (in this case $L_2$ norm, I presume). One can always choose a norm that makes a non-normal matrix normal, so the particular choice (and the physical reason behind it) should be spelled out. }\\

We did try to spell out what norm we were using around Eqs. 6-8, but I guess we didn't make it clear enough. So we have changed and added some text to try to make it clearer. We are using the $L_2$ norm. However, we have redefined our state vector (and the linear and nonlinear operators for consistency) so that the $L_2$ norm of the state vector gives the energy. This is a common thing that is done in many of our references.\\

\emph{4) page 4: the derivation of Eq. 17 is confusing. It is probably just a wrong equation reference ("we multiply eq. 15 and 16..."). I guess the procedure is simply to substitute eqs. 15 and 16 in 1 and 2, use the basis orthogonality and sum the two equations together. }\\

Yes, we did just reference the wrong equations. Sorry.\\

\emph{5) From section 3 on, several growth rate 'gamma' are introduced (with different subscripts). I suggest to list all of them in a table, with an explanation of their meaning. Otherwise it is too easy to get lost. }\\

Good idea. We've added a table now.\\

\emph{6) I am a bit confused by the use of eq. 18. The authors say that this is the way the energy evolution equation can be 'symbolically' (I suggest "formally") written. However, by inspection of eq. 17, one can explicitly write down an equation for the linear term $dE_l/dt$. Is not only the third term on the RHS of eq. 17 linear in the energy? In other words, it seems to me that the first and second terms on RHS cannot be written as linear combinations of the energy. This automatically gives the linear rate of energy injection $\gamma_{l,k}$. }\\

We attempt to make some clarifications in the text, namely by explicitly writing the equations for $dE_l/dt$ and $dE_{nl}/dt$. We do not mean that certain terms on the RHS
can be written as linear combinations of the energy. We mean that the first three terms come from the linear terms of the original HW equations. 
They can be used to evaluate the turbulent growth rate, which is somewhat analogous to the eigenmode growth rate. For instance, if $n_k, \phi_k$ are taken from one of the linear
eigenmodes and inserted into the first three terms of Eq. 17, and then one divides by (twice) the energy (Eq. 6) using the same $n_k, \phi_k$, one recovers the growth rate of that eigenmode.\\

\emph{7) page 4, last paragraph. "$\gamma_{t,k}<\gamma_{s,k}$". Here the authors are using $\gamma_t$ in lieu of $\gamma_l$, as explained in the previous paragraphs. However, this is legitimate only in quasi-steady state. Otherwise, $\gamma_t$ is a nonlinear quantity that changes in time, and can be larger than $\gamma_s$ even for systems with normal linear operators. }\\

We clarify in the text that $\gamma_t$ is defined in quasi-steady state. It is time-independent, as are all of the growth rates we define.\\

\emph{8) On page 5 the authors compare plots of $\gamma_s$ and $\gamma_t$, but they have not yet explained how these are calculated. I suggest to postpone any discussion about growth rate behavior, after explaining how they are calculated. }\\

The calculations of these growth rates are explained before the plots, and now they are hopefully explained better.\\

\emph{9) Page 5, first column: "moderate normality" should be "moderate non-normality" (i.e. normality is only for the Henrici measure equal to 0, while non-normality can be moderate, heavy, etc.) }\\

Good catch.\\

\emph{10) I have found quite confusing the explanation of how the turbulent growth rate $\gamma_t$ is calculated. On page 6, first column, it says: "may be approximated without large numerical simulations", which suggests that one needs to run small simulations. It turns out that, on the contrary, it can be evaluated analytically. However, there is a refrain in the paper referring to the complex ratio $n_k/\phi_k$. But I don't see where and why this needs to be calculated. }\\

We hope the edits have made this clear in the paper. $\gamma_t$ is calculated from the nonlinear simulation. Its calculation requires $n_k$ and $\phi_k$ as functions of time, although only the ratio $n_k/\phi_k$ is actually necessary. Our analytic method gives us a non-modal growth rate, which is an \emph{approximation} (good or bad) of the turbulent growth rate. So there is no need for small simulations, although we see how that was implied and we have thus removed the word ``large.''\\

\emph{11) Eq. 21 misses a factor 1/2 }\\

We had meant to define $u_k = 1/ \sqrt{2} (n_k, k \phi_k)^T$ so that the factor of $1/2$ wasn't necessary, but we missed the $\sqrt{2}$ factor in an earlier expression, so that was the incorrect one.\\

\emph{12) On Eq. 22. There is an unspelled assumption here on the distribution of the ensemble. Are all states equally distributed? Is such distribution physical meaningful? Also, I would emphasize the fact that by averaging over an ensemble of states, the information about the initial condition is lost (or averaged out). This is a crucial point because one of the main weaknesses of non-modal analysis with respect to the standard normal mode analysis, is the need to specify a given initial condition. }\\

Yes all states are equally distributed initially. Remember that each state at a particular $k$ is completely classified by the ratio $n_k/\phi_k$ as far as the growth rate is concerned.
So each member of the ensemble is initially one possible value of this complex ratio. But once the initial states have evolved under the linear operator for some time, not all of
the possible values of the ratio are populated. In fact, after a long enough time, this ratio is nearly equal to that for the faster growing eigenmode for all members of the ensemble.
The time average, then, produces something meaningful -- that is, an average over many ratios between a random ratio and the eigenmode ratio.

In physical terms, the system starts out random, then evolves toward the faster growing linear eigenmode structure until it reaches a nonlinear decorrelation time, at which point it becomes
random again. Remember, though, that we use an ensemble of initially random states in place of a long time series that involves randomization followed by linear evolution followed by
randomization, etc.

And yes, all information about the initial condition is lost, which is exactly our intention. Having to specify an initial condition is the main weakness of traditional non-modal analysis.
Our analysis uses ideas from non-modal analysis but attempts to average out the initial condition, making it more analogous to normal-mode analysis. So maybe when we say non-modal analysis,
we give the wrong impression, but we have no better term.\\

\emph{13) About Figure 6 and 7: these represent probably the most crucial figures of the paper. Yet, there is still no explanation on how $\gamma_t$ is evaluated from the full nonlinear simulations. }\\

Again, we've clarified it in the paper now.\\

\emph{14) I am very perplexed about the whole section V, to the extent that I would probably take it out. I appreciate the effort of using the discussed non-modal growth rate as indicator of subcritical turbulence onset. But the results of this section are very weak. First, the author have to introduce a yet different model from the ones previously described, in order to get something that can actually sustain subcritical turbulence. And more importantly, figure 11 suggest that their method does not work! So I don't understand why one would like to read this section. }\\

We agree that the section overall is weak due to the lack of success of the method in the subcritical regime, and the section isn't essential for the rest of the paper. But the issue
of predicting the subcritical onset is such an important one that it merits discussion. And although our particular method does not work, it seems plausible that some of the ideas
we use may be helpful in crafting a better method. And from doing the subcritical work, we have learned where the method breaks down. This information shouldn't be abandoned.
So we think turning Section V into an appendix may be the best solution to keep it from muddying the rest of the paper.\\

\emph{15) In figure 10a) I think it is more interesting to show the ratio of fluctuation over total energy, otherwise the numbers do not convey any information. }\\

Well the numbers are not that important, but this energy seems to be the natural normalized one: $|n|^2 + |\gradperp \phi|^2$. So we're not sure what you mean. \\

\emph{16) Finally, this manuscript contain too many typos!! Please run a spell-check before submitting (and, possibly, enumerate lines). }\\

We've now run spell-check and tried to catch any typos.

\bibliographystyle{unsrt}
\bibliography{refs}

%
%

\end{document}

### 2014-12-24 HW_non-modal.tex
Dec 23

@@
 
-\title{A Procedure to Predict the Subcritical Turbulent Onset Criterion Applied to a Modified Hasegawa-Wakatani Model}
+\title{Non-Modal Predictions of Turbulent Properties Applied to the Hasegawa-Wakatani Model}
 
@@
 This paper outlines and illustrates a non-modal method to make quantitative predictions of turbulent properties including turbulent onset criteria. 
-Our approach is to define a non-modal growth rate that is analagous to an eigenmode growth rate, but which takes into account transient growth effects.
+Our approach is to define a non-modal growth rate that is analogous to an eigenmode growth rate, but which takes into account transient growth effects.
 
@@
 the $y$ and $z$ directions with zero-value boundaries in the $x$ direction. We use a domain size of $64 \rho_s \times 64 \rho_s \times 320 \rho_s$.
-For the 2D version of the model, we replace $-\xi \gradpar^2$ with a single parameter $\alpha$ (definied as $\alpha = 4 \pi k_z^2 \rho_s^2 \omega_{ce} /\nu_{ei})$, 
+For the 2D version of the model, we replace $-\xi \gradpar^2$ with a single parameter $\alpha$ (defined as $\alpha = 4 \pi k_z^2 \rho_s^2 \omega_{ce} /\nu_{ei})$, 
 which is commonly known as the adiabaticity parameter~\cite{camargo1995,camargo1998}. Additionally we use a domain size of $50 \rho_s \times 50 \rho_s$.
@@
 \eeqar
-with $\mathbf{A}_k$ being the linear coupling matrix and $\mathbf{N}_k$ the nonlinear matrix, which we don't write explicitly. 
-Now in assessing the degree of normality of the HW model, we are not interested in the normality of $\mathbf{A}_k$ itself. 
-Rather, we are interested in the energetically-weighted version of this matrix~\cite{camargo1998,schmid2007,camporeale2010}. Since the energy in a given Fourier mode is
+with $\mathbf{A}_k$ being the linear coupling matrix and $\mathbf{N}_k$ the nonlinear operator, which we don't write explicitly.
+Now in assessing the degree of normality of the HW model, we must make a choice of norms because one can always choose a norm that makes a particular matrix normal.
+However, it is much more revealing to choose
+a physically motivated norm, especially the energy-norm, than one that simply makes a given matrix normal~\cite{camargo1998,schmid2007,camporeale2010}. Since the energy in a given Fourier mode is
 
@@
 \eeqar
-Since the square $L_2$-norm of the new state vector $\frac{1}{\sqrt{2}} \left( \begin{array}{cc} n_k \\ k_\perp \phi_k \end{array} \right)$ gives the energy contained in the mode $k$,
-$\mathbf{B}_k$ is the energetically-weighted matrix of interest. In the adiabatic limit (as long as $k_\perp$ remains finite):
+The square $L_2$-norm of the state vector $\frac{1}{\sqrt{2}} \left( n_k , k_\perp \phi_k \right)^T$ gives the energy contained in the mode $k$.
+Therefore, we choose to cast the problem in terms of Eq.~\ref{fourier_en_eqn} rather than Eq.~\ref{fourier_eqn} and focus on the linear matrix $\mathbf{B}_k$ rather than $\mathbf{A}_k$.
+Thus, we can use
+the well-known $L_2$-norm rather than a problem-dependent energy norm but still recover the same results as if we used an energy-norm with the formulation of Eq.~\ref{fourier_eqn}.
+
+In the adiabatic limit (as long as $k_\perp$ remains finite):
 
@@
 \eeq
-where $u_k = \left( n_k, k_\perp \phi_k \right)^T$.
+where $u_k = \frac{1}{\sqrt{2}} \left( n_k, k_\perp \phi_k \right)^T$.
 The solution to this is $u_k(t) = e^{\mathbf{B}_k t} u_k(0)$ where the exponential of the matrix is defined in terms of
-its Taylor expansion. The solution is dependent on the initial condition $u_k(0)$. The energy as a function of time is
+its Taylor expansion. The solution is dependent on the initial condition $u_k(0)$.
+The energy, which is synonymous with the energy amplification factor since we set $||u_k(0)||^2 =1$ evolves as
 
@@
 \label{Et}
-E_k(t) = \frac{||u_k(t)||^2}{||u_k(0)||^2} = \frac{||e^{\mathbf{B}_k t} u_k(0)||^2}{||u_k(0)||^2}
+E_k(t) = ||u_k(t)||^2 = ||e^{\mathbf{B}_k t} u_k(0)||^2
 \eeq
@@
 \centerline{\includegraphics[width=0.45\textwidth]{max_en_growth}}
-\caption{$G_{k,\alpha,\rm{max}}(t) = ||e^{\mathbf{B}_{k,\alpha} t}||^2$ for $k_x=0, k_y=1, \kappa=1$ for different values of the 2D adiabaticity parameter 
+\caption{$E_{k,\alpha,\rm{max}}(t) = ||e^{\mathbf{B}_{k,\alpha} t}||^2$ for $k_x=0, k_y=1, \kappa=1$ for different values of the 2D adiabaticity parameter 
 $\alpha$ compared to the energy growth $e^{2 \gamma_{s,\alpha,k} t}$ of the least stable eigenmode of  $\mathbf{B}_{k,\alpha}$. Also indicated for each matrix is its scalar normality measure $1-H$.
@@
 where, again, $k$ stands for $(k_x,k_y,k_z)$.
-Then, we multiply Eq.~\ref{subs_fourier_n} by $n^*_{k'} \ {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$ and Eq.~\ref{subs_fourier_phi} by 
+Then, we multiply Eq.~\ref{n_eq} by $n^*_{k'} \ {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$ and Eq.~\ref{phi_eq} by 
 $-\phi^*_{k'} \ {\rm sin} \left( k'_x x \right) e^{-i k'_y y - i k'_z z}$, take the volume integral and add the two equations together. The result is
@@
 The term $\sum_{k'} T(k,k')$ represents the nonlinear contribution (three-wave transfers between different $k$'s), 
-which we do not write explicitly. This energy evolution equation may be written symbolically as
+which we do not write explicitly. This energy evolution equation may be written formally as
 
-\beq
+\beqar
 \label{dEdt_def}
-\diff{E_k}{t} = \diff{E_{l,k}}{t} + \diff{E_{nl,k}}{t}
-\eeq
-where $\diff{E_{l,k}}{t}$ represents the the linear terms on the RHS of Eq.~\ref{Explicit_En_eqn}, which are quadratic in the fluctuating quantities $n_k$ and $\phi_k$.
-$\diff{E_{l,k}}{t}$ describes the injection of energy into the fluctuations from the free energy in the equilibrium gradients plus the dissipation from collisions and hyperdiffusion/viscosity.
+\diff{E_k}{t} = \diff{E_{l,k}}{t} + \diff{E_{nl,k}}{t}, \quad {\rm where}\\
+\label{dEl_dt_def}
+\diff{E_{l,k}}{t} = -i k_y \kappa \phi_k n^*_k - \xi k_z^2 |n_k - \phi_k|^2 \nonumber \\
+- D k^4_\perp \left( |n_k|^2 + k^2_\perp |\phi_k|^2 \right) \\
+\label{dEnl_dt_def}
+ \diff{E_{nl,k}}{t} = \sum_{k'} T(k,k')
+\eeqar
+$\diff{E_{l,k}}{t}$ is composed of the linear terms on the RHS of Eq.~\ref{Explicit_En_eqn}, which are quadratic in the fluctuating quantities $n_k$ and $\phi_k$.
+It describes the injection of energy into the fluctuations from the free energy in the equilibrium gradients plus the dissipation from collisions and hyperdiffusion/viscosity.
 $\diff{E_{nl,k}}{t}$ represents $\sum_{k'} T(k,k')$ in  Eq.~\ref{Explicit_En_eqn} and comes from the nonlinear terms in Eqs.~\ref{n_eq} and~\ref{phi_eq}. 
@@
 \label{steady_state}
-\gamma_k = \gamma_{l,k} + \gamma_{nl,k} = 0, \\
-\gamma_k \equiv  \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_k/dt}{2 E_k} dt \nonumber \\
-= \displaystyle\lim_{T \to \infty} \frac{{\rm{Log}}\left[ E_k(T)/E_k(0) \right]}{2 T} = 0.
+\gamma_{l,k} + \gamma_{nl,k} = \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_k/dt}{2 E_k} dt = 0, \\
+\label{gamma_l_def}
+{\rm where}, \ \gamma_{l,k} \equiv \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{l,k}/dt}{2 E_k} dt \\
+\label{gamma_nl_def}
+ \gamma_{nl,k} \equiv \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{nl,k}/dt}{2 E_k} dt 
 \eeqar
-$\gamma_k$ vanishes, as indicated, for steady or quasi-steady state turbulence, meaning $\gamma_{l,k} = - \gamma_{nl,k}$.
+Thus, assuming steady or quasi-steady state turbulence $\gamma_{l,k} = - \gamma_{nl,k}$. Note that these, and all other growth rates, are defined to be time-independent (time-averaged) quantities.
+At the scales of energy injection, $\gamma_{l,k}$ is positive and can be considered the turbulence analog of the linear eigenmode growth rate $\gamma_{s,k}$~\cite{friedman2012b,terry2006b}. 
+In fact, $\gamma_{l,k} = \gamma_{s,k}$ if $n_k/\phi_k$ calculated from the nonlinear simulation is constant in time and equal to $n_k/\phi_k$ calculated from the fastest growing eigenmode at $k$. 
+In general, though, these ratios are not the same because the turbulence at $k$ is composed of some linear combination of the \emph{two} eigenmodes.
 
-It is the linear rate of energy injection into the fluctuations
-$\gamma_{l,k}$ that we wish to use as the turbulent growth rate
-$\gamma_{t,k}$ because $\gamma_{l,k}$ determines how fast and where in wavenumber space energy is injected into the turbulence.
-$\gamma_{l,k}$, in other words, is the turbulence equivalent to the linear eigenmode growth rate $\gamma_{s,k}$~\cite{friedman2012b,terry2006b}. 
-That is, in a linear simulation run for a sufficiently long time where transients die away, the calculation of $\gamma_{l,k}$ from the simulated $n_k$ and $\phi_k$ gives exactly $\gamma_{s,k}$, which is generally
-calculated from the eigenvalue problem rather than a direct simulation.
-However, we always calculate $\gamma_{t,k}$ from steady state turbulent $n_k$ and $\phi_k$ fluctuations.
+\begin{table}
+\begin{tabular}{|c|p {0.102\textwidth}|p {0.26\textwidth}|}
+\hline
+Symbol & Description & \qquad \qquad \quad Formula \\ \hline
+$\gamma_s$ & Eigenmode growth rate & \qquad max(Re\{ eig( $\mathbf{B}_k$ )\})\\ \hline
+$\gamma_l = \gamma_t$ & Turbulent growth rate & \quad $\displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{l,k}/dt}{2 E_k} dt $\\ \hline
+$\gamma_{nl}$ & Nonlinear growth rate & \quad $\displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{nl,k}/dt}{2 E_k} dt = - \gamma_{l}$\\ \hline
+$\gamma_{{\rm nm}}$ & Non-modal growth rate & \ $\frac{1}{4 \tau_{nl,k}} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \tau_{nl,k}} e^{\mathbf{B}^{\dagger} \tau_{nl,k}} \} \right] $ \\ \hline
+$\gamma_{\omega}$ & $\omega$ non-modal growth rate & \ $\frac{\omega_{s,k}}{4} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \omega_{s,k}^{-1}} e^{\mathbf{B}^{\dagger} \omega_{s,k}^{-1}} \} \right]$\\ \hline
+$\gamma_{\gamma}$ & $\gamma$ non-modal growth rate & \ $\frac{\gamma_{s,k}}{4} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \gamma_{s,k}^{-1}} e^{\mathbf{B}^{\dagger} \gamma_{s,k}^{-1}} \} \right]$\\ \hline
+\end{tabular}
+\caption{Growth rate symbol descriptions and formulas}
+\label{gamma_table}
+\end{table}
 
-The turbulent growth rate $\gamma_{t,k}$ is a useful and informative quantity. 
+In order to clarify that $\gamma_{l,k}$ is calculated from the quasi-steady-state portion of the nonlinear simulation, we define $\gamma_{t,k} \equiv \gamma_{l,k}$, and we call $\gamma_{t,k}$ the turbulent growth rate.
+$\gamma_{t,k}$ is a useful and informative quantity. 
 Note that in Eq.~\ref{Explicit_En_eqn} all of the linear terms on the right hand side of the equation are negative, except for the first term, which can have either sign. 
@@
 dispersion relation -- the damped eigenmode~\cite{makwana2011}. 
-Additionally, in the moderate normality region where $\gamma_{s}$ does match $\gamma_{t}$ well, it is misleading to think of the system as being composed of only the
-unstable eigenmode.
+Additionally, in the moderate non-normality region where $\gamma_{s}$ does match $\gamma_{t}$ well, it is misleading to think of the system as being composed of only the unstable eigenmode. 
 The damped mode simply has both damping and transient growth effects depending on its relative magnitude at a particular time.
@@
 In fact, the only quantity that affects $\gamma_{t,k}$ is the average complex ratio $n_k/\phi_k$.
-Thus, it is feasible that $\gamma_{t,k}$ may be approximated without large numerical simulations.
+Thus, it is feasible that $\gamma_{t,k}$ may be approximated without numerical simulation.
 
-Our method of approximation involves modelling the nonlinearities as instantaneous randomizing forces that act periodically.
+Our method of approximation involves modeling the nonlinearities as instantaneous randomizing forces that act periodically.
 In this regard, note that the advective nonlinearity terms in Eqs.~\ref{n_eq} and~\ref{phi_eq} have a certain time scale associated with them: $\tau_{nl}^{-1} \sim v_E k_\perp$.
@@
 Next, we do the same comparison for the 3D HW model as we did for the 2D HW model.
-However, there is one additional subtelty when considering the 3D model.
+However, there is one additional subtlety when considering the 3D model.
 In the 3D model, for $k_z=0$, the least stable linear
@@
 
-We show that the modifed 3D HW system is subcritical in Fig.~\ref{crit_amp_dependence}.
+We show that the modified 3D HW system is subcritical in Fig.~\ref{crit_amp_dependence}.
 In Fig.~\ref{crit_amp_dependence} a), we plot the energy versus time for fluctuations that are initialized with the same functional form, but with
@@
 
-In Fig.~\ref{crit_amp_dependence} b), we show the amplitude threashold curve -- like that of Fig.~\ref{subcritical_diagram}.
+In Fig.~\ref{crit_amp_dependence} b), we show the amplitude threshold curve -- like that of Fig.~\ref{subcritical_diagram}.
 Each point represents the minimum amplitude that is needed for the simulation to eventually become
@@
 \caption{{\bf a)} Total energy evolution of fluctuations of the modified 3D HW model with $\xi=10$ and $\kappa=1$ where the initial perturbations are set to different amplitudes. 
-All initial perturbations are initialized randomly in the $y$ and $z$ directions and with a Guassian in the $x$ direction. 
-There is a critical initial amplitude threashold above which turbulence develops and maintains itself and below which turbuelence never develops and the fluctuations decay away. 
-{\bf b)} The amplitude threashold as a function of $\kappa$ for $\xi=10$.
+All initial perturbations are initialized randomly in the $y$ and $z$ directions and with a Gaussian in the $x$ direction. 
+There is a critical initial amplitude threshold above which turbulence develops and maintains itself and below which turbulence never develops and the fluctuations decay away. 
+{\bf b)} The amplitude threshold as a function of $\kappa$ for $\xi=10$.
 The points are obtained from nonlinear simulations like those in part a) with each point indicating the minimum amplitude needed for the simulation
@@
 
-Our method for analytically esimating this growth rate involves using a type of closure, which means modelling the effect that the nonlinearities have
+Our method for analytically estimating this growth rate involves using a type of closure, which means modeling the effect that the nonlinearities have
 on the time-averaged complex ratio of the Fourier-decomposed state variables. 

### 2014-12-24 HW_non-modal.tex
Dec 24

@@
 
-\title{Non-Modal Predictions of Turbulent Properties Applied to the Hasegawa-Wakatani Model}
+\title{A Non-Modal Analytical Procedure to Predict Turbulent Properties Applied to the Hasegawa-Wakatani Model}
 
@@
 time to be the inverse of the least stable eigenmode frequency or growth rate, which allows for prediction without nonlinear numerical simulation. 
-We test this procedure on the 2D and 3D Hasegawa-Wakatani model
-[A. Hasegawa and M. Wakatani, Phys. Rev. Lett. 50, 682 (1983)] as well as on a modified subcritical version of the 3D Hasegawa-Wakatani model and find that
-the non-modal growth rate is a good predictor of energy injection rates in some regimes, though not yet in the subcritical regime.
+We test this procedure on the 2D and 3D Hasegawa-Wakatani model [A. Hasegawa and M. Wakatani, Phys. Rev. Lett. 50, 682 (1983)] and find that
+the non-modal growth rate is a good predictor of energy injection rates, especially in the strongly non-normal, fully-developed turbulence regime.
 \end{abstract}
@@
 The calculation of linear instabilities through normal mode analysis, which solves for the eigenvalues and eigenvectors of a linearized dynamical system,
-has successfully predicted the turbulent onset criterion for many plasma and fluid systems.
+has successfully been used to predict turbulent properties such as turbulent onset and levels of transport for many plasma and fluid systems.
 Nevertheless, it is well-known in the hydrodynamics community that normal mode analysis has failed to predict turbulent onset in many experiments -- those in which the
@@
 In other words, certain fluctuation structures can access free energy from background gradients even though normal mode fluctuations cannot.
-When combined with nonlinear effects, this allows for sustained subcritical turbulence by a bootstrapping process.
+
+Moreover, even if normal mode fluctuations are unstable, they may not be as unstable as certain non-normal ones, which can produce turbulence with properties that are not predictable
+from normal mode analysis.
+When combined with nonlinear effects, non-modal fluctuations can sustain themsevles and have a permanent signature on turbulence.
 Such behavior is obscured by traditional normal mode analysis, which only effectively describes the long time asymptotic behavior of fluctuations under  
@@
 Non-modal analysis generally employs tools such as pseudospectra and maximum growth curves, which have been used to successfully explain
-why turbulence is subcritical in various instances~\cite{trefethen1993,camargo1998}. 
-However, they have not been used to predict onset criteria in subcritical systems (e.g. the transition Reynold's number in pipe flows).
-This paper outlines and illustrates a non-modal method to make quantitative predictions of turbulent properties including turbulent onset criteria. 
+why turbulence is subcritical or driven by nonlinear instability in various instances~\cite{trefethen1993,camargo1998}. 
+However, they have not been used to make quantitative predictions of turbulent transport levels or onset criteria in subcritical systems (e.g. the transition Reynold's number in pipe flows).
+This paper outlines and illustrates a non-modal method to make quantitative predictions of turbulent properties. 
 Our approach is to define a non-modal growth rate that is analogous to an eigenmode growth rate, but which takes into account transient growth effects.
@@
 to give energy injection rates, turbulent saturation levels, and the turbulent onset criteria.
-That is, we argue that a finite positive non-modal growth rate
-indicates that turbulence can exist given a sufficiently large perturbative kick.
 
@@
 that we use throughout the paper and describe its linear non-normal properties,  while in Section~\ref{sec_non_norm_turb}, we explore the effect of linear non-normality in the model on turbulence. 
-In Section~\ref{sec_nm_procedure}, we cover the details of our procedure to calculate the non-modal growth rate and compare show it to growth rates derived directly from nonlinear numerical simulations.
-Finally, in Section~\ref{sec_subcrit_prediction}, we modify the 3D Hasegawa-Wakatani model to make it subcritical and show to what degree our non-modal technique predicts the critical gradient for the onset of turbulence.
+In Section~\ref{sec_nm_procedure}, we cover the details of our procedure to calculate the non-modal growth rate and compare it to growth rates derived directly from nonlinear numerical simulations. 
+Finally, in the Appendix, we explore the possibility of predicting the critical gradient for turbulent onset with a modified subcritical 3D Hasegawa-Wakatani model.
 
@@
 and $\kappa = \rho_s/L_n$, where $L_n$ is the density scale length. The terms $D \gradperp^4 n$ and $D \gradperp^6 \phi$ are artificial hyper-diffusion and hyper-viscosity used to dissipate
-energy on small scales. We set $D \to 10^{-5}$ except in Sec.~\ref{sec_subcrit_prediction} where we set it to $0.5$.
+energy on small scales. We set $D \to 10^{-5}$.
 We associate the direction $\mathbf{b}$ as the $z$ direction. The direction perpendicular $\mathbf{b}$ that
@@
 
-\section{Prediction of Control Parameter for Subcritical Turbulent Onset}
-\label{sec_subcrit_prediction}
+\section{Conclusion}
+
+When the nonlinear terms in the model equations are energetically conservative, as they are in the HW model, 
+the linear terms linear terms control the energy injection into the turbulent fluctuations.
+This property makes the turbulent growth rate, which is defined in terms of only the linear terms in the energetic model equations, a simple yet vital piece of information
+in understanding the nature of the system. When the system is non-normal, estimating the growth rate without nonlinear simulations almost surely requires non-modal techniques.
+
+Our method for analytically estimating this growth rate involves using a type of closure, which means modeling the effect that the nonlinearities have
+on the time-averaged complex ratio of the Fourier-decomposed state variables. 
+We present one specific phenomenological closure, in which the nonlinearities instantaneously randomize the state variables once every eddy turnover time, between
+which, the linear terms deterministically evolve them. We then estimate the eddy turnover time as the inverse linear eigenmode frequency or growth rate.
+
+While our nonlinear closure model is successful in some instances, such as the highly non-normal limit in the 2D HW model and in the 3D HW model when the gradient scale length factor $\kappa$ is large,
+it is relatively unsuccessful in others. This can be seen in the appendix, where we test the procedure to make an accurate prediction of the minimum value of $\kappa$ for which a subcritical version
+of the HW model will sustain turbulence. In that case, our method fails because our closure, which assumes a state of fully-developed turbulence with a finite decorrelation time and a critically
+balanced turbulent mechanism, breaks down.
+Perhaps in the future, different closure techniques will allow for better approximation of the turbulent growth rate in different regimes.
+
+\appendix
+\section{Subcritical Onset Prediction}
 
@@
 
-\section{Conclusion}
-
-When the nonlinear terms in the model equations are energetically conservative, as they are in the HW model, the linear terms must on average inject energy into at least one wavenumber to sustain turbulent fluctuations.
-This is manifest in the positivity of the maximum turbulent growth rate, which is defined in terms of only the linear terms in the energetic model equations.
-Thus, an estimation of the turbulent growth rate provides vital
-information into the nature of the system. Furthermore, such an estimate almost surely requires non-modal techniques when the system is non-normal.
-
-Our method for analytically estimating this growth rate involves using a type of closure, which means modeling the effect that the nonlinearities have
-on the time-averaged complex ratio of the Fourier-decomposed state variables. 
-We present one specific phenomenological closure, in which the nonlinearities instantaneously randomize the state variables once every eddy turnover time, between
-which, the linear terms deterministically evolve them. We then estimate the eddy turnover time as the inverse linear eigenmode frequency or growth rate.
-
-While our nonlinear closure model is successful in some instances, such as the highly non-normal limit in the 2D HW model and in the 3D HW model when the gradient scale length factor $\kappa$ is large, it is relatively unsuccessful
-in others, especially when the system is not highly non-normal. This prevents an accurate prediction of the transitional value $\kappa_g$ that marks where the system first becomes turbulent.
-Perhaps in the future, more advanced closure techniques will allow for better approximation of the turbulent growth rate.
-
 \begin{acknowledgments}
+This work was performed under the auspices of the U.S. Department of Energy by Lawrence Livermore National Laboratory under Contract DE-AC52-07NA27344.
 This work was supported by the National Science Foundation (Grant PHY-1202007)
@@
 %\bibliographystyle{phaip}
-\bibliography{refs}
+%\bibliography{refs}
+
+\begin{thebibliography}{38}
+\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi
+\expandafter\ifx\csname bibnamefont\endcsname\relax
+  \def\bibnamefont#1{#1}\fi
+\expandafter\ifx\csname bibfnamefont\endcsname\relax
+  \def\bibfnamefont#1{#1}\fi
+\expandafter\ifx\csname citenamefont\endcsname\relax
+  \def\citenamefont#1{#1}\fi
+\expandafter\ifx\csname url\endcsname\relax
+  \def\url#1{\texttt{#1}}\fi
+\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
+\providecommand{\bibinfo}[2]{#2}
+\providecommand{\eprint}[2][]{\url{#2}}
+
+\bibitem[{\citenamefont{Drazin and Reid}(1981)}]{drazin1981}
+\bibinfo{author}{\bibfnamefont{P.~G.} \bibnamefont{Drazin}} \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{W.~H.} \bibnamefont{Reid}},
+  \emph{\bibinfo{title}{Hydrodynamic Stability}} (\bibinfo{publisher}{Cambridge
+  University Press}, \bibinfo{year}{1981}).
+
+\bibitem[{\citenamefont{Waltz}(1985)}]{waltz1985}
+\bibinfo{author}{\bibfnamefont{R.~E.} \bibnamefont{Waltz}},
+  \bibinfo{journal}{Phys. Rev. Lett.} \textbf{\bibinfo{volume}{55}},
+  \bibinfo{pages}{1098} (\bibinfo{year}{1985}).
+
+\bibitem[{\citenamefont{Scott}(1990)}]{scott1990}
+\bibinfo{author}{\bibfnamefont{B.~D.} \bibnamefont{Scott}},
+  \bibinfo{journal}{Phys. Rev. Lett.} \textbf{\bibinfo{volume}{65}},
+  \bibinfo{pages}{3289} (\bibinfo{year}{1990}).
+
+\bibitem[{\citenamefont{Nordman et~al.}(1993)\citenamefont{Nordman, Pavlenko,
+  and Weiland}}]{nordman1993}
+\bibinfo{author}{\bibfnamefont{H.}~\bibnamefont{Nordman}},
+  \bibinfo{author}{\bibfnamefont{V.~P.} \bibnamefont{Pavlenko}},
+  \bibnamefont{and} \bibinfo{author}{\bibfnamefont{J.}~\bibnamefont{Weiland}},
+  \bibinfo{journal}{Phys. Fluids B} \textbf{\bibinfo{volume}{5}},
+  \bibinfo{pages}{402} (\bibinfo{year}{1993}).
+
+\bibitem[{\citenamefont{Biskamp and Zeiler}(1995)}]{biskamp1995}
+\bibinfo{author}{\bibfnamefont{D.}~\bibnamefont{Biskamp}} \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Zeiler}},
+  \bibinfo{journal}{Phys. Rev. Lett.} \textbf{\bibinfo{volume}{74}},
+  \bibinfo{pages}{706} (\bibinfo{year}{1995}).
+
+\bibitem[{\citenamefont{Drake et~al.}(1995)\citenamefont{Drake, Zeiler, and
+  Biskamp}}]{drake1995}
+\bibinfo{author}{\bibfnamefont{J.~F.} \bibnamefont{Drake}},
+  \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Zeiler}}, \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{D.}~\bibnamefont{Biskamp}},
+  \bibinfo{journal}{Phys. Rev. Lett.} \textbf{\bibinfo{volume}{75}},
+  \bibinfo{pages}{4222} (\bibinfo{year}{1995}).
+
+\bibitem[{\citenamefont{Itoh et~al.}(1996)\citenamefont{Itoh, Itoh, Yagi, and
+  Fukuyama}}]{itoh1996}
+\bibinfo{author}{\bibfnamefont{K.}~\bibnamefont{Itoh}},
+  \bibinfo{author}{\bibfnamefont{S.~I.} \bibnamefont{Itoh}},
+  \bibinfo{author}{\bibfnamefont{M.}~\bibnamefont{Yagi}}, \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Fukuyama}},
+  \bibinfo{journal}{Journal of the Physical Society of Japan}
+  \textbf{\bibinfo{volume}{65}}, \bibinfo{pages}{2749} (\bibinfo{year}{1996}).
+
+\bibitem[{\citenamefont{Camargo et~al.}(1998)\citenamefont{Camargo, Tippett,
+  and Caldas}}]{camargo1998}
+\bibinfo{author}{\bibfnamefont{S.~J.} \bibnamefont{Camargo}},
+  \bibinfo{author}{\bibfnamefont{M.~K.} \bibnamefont{Tippett}},
+  \bibnamefont{and} \bibinfo{author}{\bibfnamefont{I.~L.}
+  \bibnamefont{Caldas}}, \bibinfo{journal}{Phys. Rev. E}
+  \textbf{\bibinfo{volume}{58}}, \bibinfo{pages}{3693} (\bibinfo{year}{1998}).
+
+\bibitem[{\citenamefont{Krommes}(1999)}]{krommes1999}
+\bibinfo{author}{\bibfnamefont{J.~A.} \bibnamefont{Krommes}},
+  \bibinfo{journal}{Plasma Phys. Control. Fusion}
+  \textbf{\bibinfo{volume}{41}}, \bibinfo{pages}{A641} (\bibinfo{year}{1999}).
+
+\bibitem[{\citenamefont{Camporeale et~al.}(2009)\citenamefont{Camporeale,
+  Burgess, and Passot}}]{camporeale2009}
+\bibinfo{author}{\bibfnamefont{E.}~\bibnamefont{Camporeale}},
+  \bibinfo{author}{\bibfnamefont{D.}~\bibnamefont{Burgess}}, \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{T.}~\bibnamefont{Passot}},
+  \bibinfo{journal}{Phys. Plasmas} \textbf{\bibinfo{volume}{16}},
+  \bibinfo{pages}{030703} (\bibinfo{year}{2009}).
+
+\bibitem[{\citenamefont{Schekochihin et~al.}(2012)\citenamefont{Schekochihin,
+  Highcock, and Cowley}}]{schekochihin2012}
+\bibinfo{author}{\bibfnamefont{A.~A.} \bibnamefont{Schekochihin}},
+  \bibinfo{author}{\bibfnamefont{E.~G.} \bibnamefont{Highcock}},
+  \bibnamefont{and} \bibinfo{author}{\bibfnamefont{S.~C.}
+  \bibnamefont{Cowley}}, \bibinfo{journal}{Plasma Phys. Control. Fusion}
+  \textbf{\bibinfo{volume}{54}}, \bibinfo{pages}{055011}
+  (\bibinfo{year}{2012}).
+
+\bibitem[{\citenamefont{Highcock et~al.}(2012)\citenamefont{Highcock,
+  Schekochihin, Cowley, Barnes, Parra, Roach, and Dorland}}]{highcock2012}
+\bibinfo{author}{\bibfnamefont{E.~G.} \bibnamefont{Highcock}},
+  \bibinfo{author}{\bibfnamefont{A.~A.} \bibnamefont{Schekochihin}},
+  \bibinfo{author}{\bibfnamefont{S.~C.} \bibnamefont{Cowley}},
+  \bibinfo{author}{\bibfnamefont{M.}~\bibnamefont{Barnes}},
+  \bibinfo{author}{\bibfnamefont{F.~I.} \bibnamefont{Parra}},
+  \bibinfo{author}{\bibfnamefont{C.~M.} \bibnamefont{Roach}}, \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{W.}~\bibnamefont{Dorland}},
+  \bibinfo{journal}{Phys. Rev. Lett.} \textbf{\bibinfo{volume}{109}},
+  \bibinfo{pages}{265001} (\bibinfo{year}{2012}).
+
+\bibitem[{\citenamefont{Gustavsson}(1991)}]{gustavsson1991}
+\bibinfo{author}{\bibfnamefont{L.~H.} \bibnamefont{Gustavsson}},
+  \bibinfo{journal}{J. Fluid Mech} \textbf{\bibinfo{volume}{224}},
+  \bibinfo{pages}{221} (\bibinfo{year}{1991}).
+
+\bibitem[{\citenamefont{Butler and Farrell}(1992)}]{butler1992}
+\bibinfo{author}{\bibfnamefont{K.~M.} \bibnamefont{Butler}} \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{B.~F.} \bibnamefont{Farrell}},
+  \bibinfo{journal}{Phys. Fluids A} \textbf{\bibinfo{volume}{4}},
+  \bibinfo{pages}{1637} (\bibinfo{year}{1992}).
+
+\bibitem[{\citenamefont{Trefethen et~al.}(1993)\citenamefont{Trefethen,
+  Trefethen, Reddy, and Driscoll}}]{trefethen1993}
+\bibinfo{author}{\bibfnamefont{L.~N.} \bibnamefont{Trefethen}},
+  \bibinfo{author}{\bibfnamefont{A.~E.} \bibnamefont{Trefethen}},
+  \bibinfo{author}{\bibfnamefont{S.~C.} \bibnamefont{Reddy}}, \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{T.~A.} \bibnamefont{Driscoll}},
+  \bibinfo{journal}{Science} \textbf{\bibinfo{volume}{261}},
+  \bibinfo{pages}{578} (\bibinfo{year}{1993}).
+
+\bibitem[{\citenamefont{Reddy et~al.}(1993)\citenamefont{Reddy, Schmid, and
+  Henningson}}]{reddy1993}
+\bibinfo{author}{\bibfnamefont{S.~C.} \bibnamefont{Reddy}},
+  \bibinfo{author}{\bibfnamefont{P.~J.} \bibnamefont{Schmid}},
+  \bibnamefont{and} \bibinfo{author}{\bibfnamefont{D.~S.}
+  \bibnamefont{Henningson}}, \bibinfo{journal}{SIAM J. Appl. Math.}
+  \textbf{\bibinfo{volume}{53}}, \bibinfo{pages}{15} (\bibinfo{year}{1993}).
+
+\bibitem[{\citenamefont{Henningson and Reddy}(1994)}]{henningson1994}
+\bibinfo{author}{\bibfnamefont{D.~S.} \bibnamefont{Henningson}}
+  \bibnamefont{and} \bibinfo{author}{\bibfnamefont{S.~C.} \bibnamefont{Reddy}},
+  \bibinfo{journal}{Phys. Fluids} \textbf{\bibinfo{volume}{6}},
+  \bibinfo{pages}{1396} (\bibinfo{year}{1994}).
+
+\bibitem[{\citenamefont{Schmid}(2007)}]{schmid2007}
+\bibinfo{author}{\bibfnamefont{P.~J.} \bibnamefont{Schmid}},
+  \bibinfo{journal}{Annu. Rev. Fluid Mech.} \textbf{\bibinfo{volume}{39}},
+  \bibinfo{pages}{129} (\bibinfo{year}{2007}).
+
+\bibitem[{\citenamefont{Hasegawa and Wakatani}(1983)}]{hasegawa1983}
+\bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Hasegawa}} \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{M.}~\bibnamefont{Wakatani}},
+  \bibinfo{journal}{Phys. Rev. Lett.} \textbf{\bibinfo{volume}{50}},
+  \bibinfo{pages}{682} (\bibinfo{year}{1983}).
+
+\bibitem[{\citenamefont{Dudson et~al.}(2009)\citenamefont{Dudson, Umansky, Xu,
+  Snyder, and Wilson}}]{dudson2009}
+\bibinfo{author}{\bibfnamefont{B.~D.} \bibnamefont{Dudson}},
+  \bibinfo{author}{\bibfnamefont{M.~V.} \bibnamefont{Umansky}},
+  \bibinfo{author}{\bibfnamefont{X.~Q.} \bibnamefont{Xu}},
+  \bibinfo{author}{\bibfnamefont{P.~B.} \bibnamefont{Snyder}},
+  \bibnamefont{and} \bibinfo{author}{\bibfnamefont{H.~R.}
+  \bibnamefont{Wilson}}, \bibinfo{journal}{Computer Physics Communications} pp.
+  \bibinfo{pages}{1467--1480} (\bibinfo{year}{2009}).
+
+\bibitem[{\citenamefont{Camargo et~al.}(1995)\citenamefont{Camargo, Biskamp,
+  and Scott}}]{camargo1995}
+\bibinfo{author}{\bibfnamefont{S.~J.} \bibnamefont{Camargo}},
+  \bibinfo{author}{\bibfnamefont{D.}~\bibnamefont{Biskamp}}, \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{B.~D.} \bibnamefont{Scott}},
+  \bibinfo{journal}{Phys. Plasmas} \textbf{\bibinfo{volume}{2}},
+  \bibinfo{pages}{48} (\bibinfo{year}{1995}).
+
+\bibitem[{\citenamefont{Friedman et~al.}(2012)\citenamefont{Friedman, Carter,
+  Umansky, Schaffner, and Dudson}}]{friedman2012b}
+\bibinfo{author}{\bibfnamefont{B.}~\bibnamefont{Friedman}},
+  \bibinfo{author}{\bibfnamefont{T.~A.} \bibnamefont{Carter}},
+  \bibinfo{author}{\bibfnamefont{M.~V.} \bibnamefont{Umansky}},
+  \bibinfo{author}{\bibfnamefont{D.}~\bibnamefont{Schaffner}},
+  \bibnamefont{and} \bibinfo{author}{\bibfnamefont{B.}~\bibnamefont{Dudson}},
+  \bibinfo{journal}{Phys. Plasmas} \textbf{\bibinfo{volume}{19}},
+  \bibinfo{pages}{102307} (\bibinfo{year}{2012}).
+
+\bibitem[{\citenamefont{Camporeale et~al.}(2010)\citenamefont{Camporeale,
+  Burgess, and Passot}}]{camporeale2010}
+\bibinfo{author}{\bibfnamefont{E.}~\bibnamefont{Camporeale}},
+  \bibinfo{author}{\bibfnamefont{D.}~\bibnamefont{Burgess}}, \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{T.}~\bibnamefont{Passot}},
+  \bibinfo{journal}{The Astrophysical Journal} \textbf{\bibinfo{volume}{715}},
+  \bibinfo{pages}{260} (\bibinfo{year}{2010}).
+
+\bibitem[{\citenamefont{Trefethen and Embree}(2005)}]{trefethen2005}
+\bibinfo{author}{\bibfnamefont{L.~N.} \bibnamefont{Trefethen}}
+  \bibnamefont{and} \bibinfo{author}{\bibfnamefont{M.}~\bibnamefont{Embree}},
+  \emph{\bibinfo{title}{Spectra and Pseudospectra: The Behavior of Nonnormal
+  Matrices and Operators}} (\bibinfo{publisher}{Princeton Univ. Press},
+  \bibinfo{year}{2005}).
+
+\bibitem[{\citenamefont{Henrici}(1962)}]{henrici1962}
+\bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Henrici}},
+  \bibinfo{journal}{Numer. Math} \textbf{\bibinfo{volume}{4}},
+  \bibinfo{pages}{24} (\bibinfo{year}{1962}).
+
+\bibitem[{\citenamefont{Friedman et~al.}(2013)\citenamefont{Friedman, Carter,
+  Umansky, Schaffner, and Joseph}}]{friedman2013}
+\bibinfo{author}{\bibfnamefont{B.}~\bibnamefont{Friedman}},
+  \bibinfo{author}{\bibfnamefont{T.~A.} \bibnamefont{Carter}},
+  \bibinfo{author}{\bibfnamefont{M.~V.} \bibnamefont{Umansky}},
+  \bibinfo{author}{\bibfnamefont{D.}~\bibnamefont{Schaffner}},
+  \bibnamefont{and} \bibinfo{author}{\bibfnamefont{I.}~\bibnamefont{Joseph}},
+  \bibinfo{journal}{Phys. Plasmas} \textbf{\bibinfo{volume}{20}},
+  \bibinfo{pages}{055704} (\bibinfo{year}{2013}).
+
+\bibitem[{\citenamefont{Terry et~al.}(2006)\citenamefont{Terry, Baver, and
+  Gupta}}]{terry2006b}
+\bibinfo{author}{\bibfnamefont{P.~W.} \bibnamefont{Terry}},
+  \bibinfo{author}{\bibfnamefont{D.~A.} \bibnamefont{Baver}}, \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{S.}~\bibnamefont{Gupta}},
+  \bibinfo{journal}{Phys. Plasmas} \textbf{\bibinfo{volume}{13}},
+  \bibinfo{pages}{022307} (\bibinfo{year}{2006}).
+
+\bibitem[{\citenamefont{Makwana et~al.}(2011)\citenamefont{Makwana, Terry, Kim,
+  and Hatch}}]{makwana2011}
+\bibinfo{author}{\bibfnamefont{K.~D.} \bibnamefont{Makwana}},
+  \bibinfo{author}{\bibfnamefont{P.~W.} \bibnamefont{Terry}},
+  \bibinfo{author}{\bibfnamefont{J.~H.} \bibnamefont{Kim}}, \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{D.~R.} \bibnamefont{Hatch}},
+  \bibinfo{journal}{Phys. Plasmas} \textbf{\bibinfo{volume}{18}},
+  \bibinfo{pages}{012302} (\bibinfo{year}{2011}).
+
+\bibitem[{\citenamefont{Baver et~al.}(2002)\citenamefont{Baver, Terry, Gatto,
+  and Fernandez}}]{baver2002}
+\bibinfo{author}{\bibfnamefont{D.~A.} \bibnamefont{Baver}},
+  \bibinfo{author}{\bibfnamefont{P.~W.} \bibnamefont{Terry}},
+  \bibinfo{author}{\bibfnamefont{R.}~\bibnamefont{Gatto}}, \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{E.}~\bibnamefont{Fernandez}},
+  \bibinfo{journal}{Phys. Plasmas} \textbf{\bibinfo{volume}{9}},
+  \bibinfo{pages}{3318} (\bibinfo{year}{2002}).
+
+\bibitem[{\citenamefont{Hatch et~al.}(2011)\citenamefont{Hatch, Terry, Jenko,
+  Merz, and Nevins}}]{hatch2011}
+\bibinfo{author}{\bibfnamefont{D.~R.} \bibnamefont{Hatch}},
+  \bibinfo{author}{\bibfnamefont{P.~W.} \bibnamefont{Terry}},
+  \bibinfo{author}{\bibfnamefont{F.}~\bibnamefont{Jenko}},
+  \bibinfo{author}{\bibfnamefont{F.}~\bibnamefont{Merz}}, \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{W.~M.} \bibnamefont{Nevins}},
+  \bibinfo{journal}{Phys. Rev. Lett.} \textbf{\bibinfo{volume}{106}},
+  \bibinfo{pages}{115003} (\bibinfo{year}{2011}).
+
+\bibitem[{\citenamefont{Scott}(2002)}]{scott2002}
+\bibinfo{author}{\bibfnamefont{B.~D.} \bibnamefont{Scott}},
+  \bibinfo{journal}{New J. Physics} \textbf{\bibinfo{volume}{4}},
+  \bibinfo{pages}{52.1} (\bibinfo{year}{2002}).
+
+\bibitem[{\citenamefont{Scott}(2005)}]{scott2005}
+\bibinfo{author}{\bibfnamefont{B.~D.} \bibnamefont{Scott}},
+  \bibinfo{journal}{Phys. Plasmas} \textbf{\bibinfo{volume}{12}},
+  \bibinfo{pages}{062314} (\bibinfo{year}{2005}).
+
+\bibitem[{\citenamefont{Umansky et~al.}(2009)\citenamefont{Umansky, Xu, Dudson,
+  Lodestro, and Myra}}]{umansky2009}
+\bibinfo{author}{\bibfnamefont{M.~V.} \bibnamefont{Umansky}},
+  \bibinfo{author}{\bibfnamefont{X.~Q.} \bibnamefont{Xu}},
+  \bibinfo{author}{\bibfnamefont{B.}~\bibnamefont{Dudson}},
+  \bibinfo{author}{\bibfnamefont{L.}~\bibnamefont{Lodestro}}, \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{J.}~\bibnamefont{Myra}},
+  \bibinfo{journal}{Phys. Plasmas} \textbf{\bibinfo{volume}{180}},
+  \bibinfo{pages}{887} (\bibinfo{year}{2009}).
+
+\bibitem[{\citenamefont{Friedman and Carter}(2014)}]{friedman2014}
+\bibinfo{author}{\bibfnamefont{B.}~\bibnamefont{Friedman}} \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{T.~A.} \bibnamefont{Carter}},
+  \bibinfo{journal}{Phys. Rev. Lett.} \textbf{\bibinfo{volume}{113}},
+  \bibinfo{pages}{025003} (\bibinfo{year}{2014}).
+
+\bibitem[{\citenamefont{Grossmann}(2000)}]{grossmann2000}
+\bibinfo{author}{\bibfnamefont{S.}~\bibnamefont{Grossmann}},
+  \bibinfo{journal}{Reviews of Modern Phys.} \textbf{\bibinfo{volume}{72}},
+  \bibinfo{pages}{603} (\bibinfo{year}{2000}).
+
+\bibitem[{\citenamefont{Waleffe}(1995)}]{waleffe1995b}
+\bibinfo{author}{\bibfnamefont{F.}~\bibnamefont{Waleffe}},
+  \bibinfo{journal}{Phys. Fluids} \textbf{\bibinfo{volume}{7}},
+  \bibinfo{pages}{3060} (\bibinfo{year}{1995}).
+
+\bibitem[{\citenamefont{Hof et~al.}(2003)\citenamefont{Hof, Juel, and
+  Mullin}}]{hof2003}
+\bibinfo{author}{\bibfnamefont{B.}~\bibnamefont{Hof}},
+  \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Juel}}, \bibnamefont{and}
+  \bibinfo{author}{\bibfnamefont{T.}~\bibnamefont{Mullin}},
+  \bibinfo{journal}{Phys. Rev. Lett.} \textbf{\bibinfo{volume}{91}},
+  \bibinfo{pages}{244502} (\bibinfo{year}{2003}).
+
+\bibitem[{\citenamefont{Cossu}(2005)}]{cossu2005}
+\bibinfo{author}{\bibfnamefont{C.}~\bibnamefont{Cossu}},
+  \bibinfo{journal}{Comptes Rendus Mecanique} \textbf{\bibinfo{volume}{333}},
+  \bibinfo{pages}{331} (\bibinfo{year}{2005}).
+
+\end{thebibliography}
 

### 2014-12-26 HW_non-modal.tex
Dec 26

@@
 
-\title{A Non-Modal Analytical Procedure to Predict Turbulent Properties Applied to the Hasegawa-Wakatani Model}
+\title{A Non-Modal Analytical Method to Predict Turbulent Properties Applied to the Hasegawa-Wakatani Model}
 
@@
 In subcritical systems, turbulence develops even though all linear eigenvectors are stable such that infinitesimal perturbations on the equilibrium state cannot grow exponentially. 
-Nevertheless, finite amplitude perturbations still excite turbulence through nonlinear instabilities.
-Subcritical turbulence is common in plasmas as well, and several instances have been studied in the literature~\cite{waltz1985,scott1990,nordman1993,biskamp1995,drake1995,itoh1996,camargo1998,krommes1999,camporeale2009,schekochihin2012,highcock2012}.
+Nevertheless, finite amplitude perturbations excite turbulence through nonlinear instabilities.
+Subcritical turbulence is common in plasmas as well, and several instances have been studied~\cite{waltz1985,scott1990,nordman1993,biskamp1995,drake1995,itoh1996,camargo1998,krommes1999,camporeale2009,schekochihin2012,highcock2012}.
 
@@
 from normal mode analysis.
-When combined with nonlinear effects, non-modal fluctuations can sustain themsevles and have a permanent signature on turbulence.
+When combined with nonlinear effects, non-modal fluctuations can sustain themselves and have a permanent signature on turbulence.
 Such behavior is obscured by traditional normal mode analysis, which only effectively describes the long time asymptotic behavior of fluctuations under  
@@
 nonlinearities randomize the complex ratios on a characteristic nonlinear time scale.
-We estimate the nonlinear time scale by a characteristic linear time scale, taking either the inverse linear eigenmode frequency or growth rate as the linear time scale.
+We estimate the nonlinear time scale by a characteristic linear time scale, taking either the inverse linear eigenmode frequency or growth rate.
 Between the nonlinear randomization, the system evolves linearly, and we use this evolution to calculate the non-modal growth rate. 
@@
 
-The paper is organized as follows: in Section~\ref{sec_hw_model}, we specify our illustrative model, the 2D and 3D Hasegawa-Wakatani model,
-that we use throughout the paper and describe its linear non-normal properties,  while in Section~\ref{sec_non_norm_turb}, we explore the effect of linear non-normality in the model on turbulence. 
+The paper is organized as follows: in Section~\ref{sec_hw_model}, we describe the linear non-normal properties of our illustrative model, the 2D and 3D Hasegawa-Wakatani model,
+while in Section~\ref{sec_non_norm_turb}, we explore the effect of linear non-normality in the model on turbulence. 
 In Section~\ref{sec_nm_procedure}, we cover the details of our procedure to calculate the non-modal growth rate and compare it to growth rates derived directly from nonlinear numerical simulations. 
-Finally, in the Appendix, we explore the possibility of predicting the critical gradient for turbulent onset with a modified subcritical 3D Hasegawa-Wakatani model.
+We leave to the Appendix the unsuccessful attempt of predicting the critical gradient for turbulent onset with a modified subcritical 3D Hasegawa-Wakatani model.
 
@@
 
-In order to illustrate and test our non-modal technique, we use as an example the well-known and well-analyzed Hasegawa-Wakatani (HW) model~\cite{hasegawa1983}.
+In order to illustrate and test our non-modal technique, we use as an example the well-analyzed Hasegawa-Wakatani (HW) model~\cite{hasegawa1983}.
 In this section, we analyze the linear normality of the model.
@@
 \eeqar
-The normalizations are $\phi \to e \phi/T_e, n \to n/n_0, \grad \to \rho_s \grad , x,y,z \to x,y,z/\rho_s, t \to \omega_{ci} t $, where $\rho_s$ is the ion sound gyroradius at the electron temperature and $\omega_{ci}$ is
+The normalizations are $\phi \to e \phi/T_e, n \to n/n_0, \grad \to \rho_s \grad , (x,y,z) \to (x,y,z)/\rho_s, t \to \omega_{ci} t $, where $\rho_s$ is the ion sound gyroradius at the electron temperature and $\omega_{ci}$ is
 the ion cyclotron frequency.
@@
 The degree of non-normality of the linear operator in the HW model is largely a function of the adiabaticity parameter~\cite{camargo1998}, or in the 3D case, of $\xi \gradpar^2$. 
-A normal matrix is on that commutes with its adjoint: $\mathbf{D} \mathbf{D}^\dagger = \mathbf{D}^\dagger \mathbf{D}$.
+A normal matrix is one that commutes with its adjoint: $\mathbf{D} \mathbf{D}^\dagger = \mathbf{D}^\dagger \mathbf{D}$.
 The HW model has a normal linear operator in the adiabatic limit: $\alpha, \xi \to \infty$. 
@@
 with $\mathbf{A}_k$ being the linear coupling matrix and $\mathbf{N}_k$ the nonlinear operator, which we don't write explicitly.
-Now in assessing the degree of normality of the HW model, we must make a choice of norms because one can always choose a norm that makes a particular matrix normal.
+Now in assessing the degree of non-normality of the HW model, we must make a choice of norms because one can always choose a norm that makes a particular matrix normal.
 However, it is much more revealing to choose
@@
 \eeqar
-The square $L_2$-norm of the state vector $\frac{1}{\sqrt{2}} \left( n_k , k_\perp \phi_k \right)^T$ gives the energy contained in the mode $k$.
+The square $L_2$-norm of the state vector $\frac{1}{\sqrt{2}} \left( n_k , k_\perp \phi_k \right)^T$ is the energy in the mode $k$.
 Therefore, we choose to cast the problem in terms of Eq.~\ref{fourier_en_eqn} rather than Eq.~\ref{fourier_eqn} and focus on the linear matrix $\mathbf{B}_k$ rather than $\mathbf{A}_k$.
 Thus, we can use
-the well-known $L_2$-norm rather than a problem-dependent energy norm but still recover the same results as if we used an energy-norm with the formulation of Eq.~\ref{fourier_eqn}.
+the $L_2$-norm rather than a problem-dependent energy norm and still recover the same results as if we used an energy-norm with the formulation of Eq.~\ref{fourier_eqn}.
 
@@
 Thus, $\mathbf{B}_k$ is non-normal.
-In general, as the adiabaticity parameter is varied, the normality of $\mathbf{B}_k$ changes.
+In general, as the adiabaticity parameter is varied, the non-normality of $\mathbf{B}_k$ changes.
 There are several ways to assess the non-normality of a matrix~\cite{trefethen2005}.
@@
 where the operator $|| \cdot ||$ is the $L_2$-norm.
-This scalar normality measure can take on values: $0 \le H < 1$, where $H=0$ for a normal matrix.
+This scalar non-normality measure can take on values: $0 \le H < 1$, where $H=0$ for a normal matrix.
 Since we are generally interested in highly non-normal
@@
 its Taylor expansion. The solution is dependent on the initial condition $u_k(0)$.
-The energy, which is synonymous with the energy amplification factor since we set $||u_k(0)||^2 =1$ evolves as
+The energy, which is synonymous with the energy amplification factor since we set $||u_k(0)||^2 =1$, evolves as
 
@@
 The equality holds when $\mathbf{B}_k$ is normal. 
-Thus, non-normality in practice allows a system to undergo more energy growth than is dictated by the most unstable linear eigenmode. We show this in Fig.~\ref{max_en_growth} where we plot
+Thus, non-normality in practice allows a system's energy to grow faster than is dictated by the most unstable linear eigenmode. We show this in Fig.~\ref{max_en_growth} where we plot
 $||e^{\mathbf{B}_k t}||^2$ for different values of $\alpha$ along with $e^{2 \gamma_{s,k} t}$, showing that transient non-modal growth is strong for small values of the scalar normality measure $1-H$, which scales monotonically
@@
 \caption{$E_{k,\alpha,\rm{max}}(t) = ||e^{\mathbf{B}_{k,\alpha} t}||^2$ for $k_x=0, k_y=1, \kappa=1$ for different values of the 2D adiabaticity parameter 
-$\alpha$ compared to the energy growth $e^{2 \gamma_{s,\alpha,k} t}$ of the least stable eigenmode of  $\mathbf{B}_{k,\alpha}$. Also indicated for each matrix is its scalar normality measure $1-H$.
+$\alpha$ compared to the energy growth $e^{2 \gamma_{s,\alpha,k} t}$ of the least stable eigenmode of  $\mathbf{B}_{k,\alpha}$.
+Also indicated for each matrix is its scalar non-normality measure $1-H$.
 The time axis is normalized to $\gamma_{s,\alpha}$, which is different for each of the curves. This normalization makes the $e^{2 \gamma_{s,\alpha,k} t}$ line the same for all $\alpha$. 
@@
 
-Non-normality manifests itself in the eigenvectors -- non-normal matrices or operators have nonorthogonal linear eigenvectors.
+Non-normality manifests itself in the eigenvectors; non-normal matrices or operators have nonorthogonal linear eigenvectors.
 For the HW model, each Fourier matrix $\mathbf{B}_k$ has two eigenvectors associated
@@
 When the eigenvectors are nearly anti-parallel and initialized with finite amplitude,
-their superposition, which constitutes the total system at $k$, may linearly evolve so that the norm of the superposition grows faster than either individual eigenvector. A paradigmatic illustration
+their linear superposition constituting the component of the system at $k$ may linearly evolve so that the norm of the superposition grows faster than either individual eigenvector.
+A paradigmatic illustration
 of this may be seen in a review by Schmid~\cite{schmid2007}.
@@
 $\diff{E_{l,k}}{t}$ is composed of the linear terms on the RHS of Eq.~\ref{Explicit_En_eqn}, which are quadratic in the fluctuating quantities $n_k$ and $\phi_k$.
-It describes the injection of energy into the fluctuations from the free energy in the equilibrium gradients plus the dissipation from collisions and hyperdiffusion/viscosity.
-$\diff{E_{nl,k}}{t}$ represents $\sum_{k'} T(k,k')$ in  Eq.~\ref{Explicit_En_eqn} and comes from the nonlinear terms in Eqs.~\ref{n_eq} and~\ref{phi_eq}. 
+It describes the injection of energy into the fluctuations from the free energy in the equilibrium gradients minus the dissipation from collisions and hyperdiffusion/viscosity.
 The terms in $T(k,k')$ are triadic in the fluctuating quantities and account for the energy exchange between fluctuations with different $k$'s. 
@@
 \label{gamma_l_def}
-{\rm where}, \ \gamma_{l,k} \equiv \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{l,k}/dt}{2 E_k} dt \\
+{\rm where}, \ \gamma_{l,k} \equiv \displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{l,k}/dt}{2 E_k} dt, \\
 \label{gamma_nl_def}
@@
 Thus, assuming steady or quasi-steady state turbulence $\gamma_{l,k} = - \gamma_{nl,k}$. Note that these, and all other growth rates, are defined to be time-independent (time-averaged) quantities.
+To keep track of all of the growth rates we define here and in the next section, refer to Table~\ref{gamma_table}.
+
 At the scales of energy injection, $\gamma_{l,k}$ is positive and can be considered the turbulence analog of the linear eigenmode growth rate $\gamma_{s,k}$~\cite{friedman2012b,terry2006b}. 
 In fact, $\gamma_{l,k} = \gamma_{s,k}$ if $n_k/\phi_k$ calculated from the nonlinear simulation is constant in time and equal to $n_k/\phi_k$ calculated from the fastest growing eigenmode at $k$. 
-In general, though, these ratios are not the same because the turbulence at $k$ is composed of some linear combination of the \emph{two} eigenmodes.
+In general, though, these ratios are not the same because the turbulence at $k$ is composed of some linear combination of the \emph{two} eigenmodes, which have different $n_k/\phi_k$ ratios.
 
@@
 Symbol & Description & \qquad \qquad \quad Formula \\ \hline
-$\gamma_s$ & Eigenmode growth rate & \qquad max(Re\{ eig( $\mathbf{B}_k$ )\})\\ \hline
-$\gamma_l = \gamma_t$ & Turbulent growth rate & \quad $\displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{l,k}/dt}{2 E_k} dt $\\ \hline
-$\gamma_{nl}$ & Nonlinear growth rate & \quad $\displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{nl,k}/dt}{2 E_k} dt = - \gamma_{l}$\\ \hline
+$\gamma_s$ & Eigenmode growth rate & \qquad max($Re\{ eig( \mathbf{B}_k)\}$)\\ \hline
+$\gamma_l \equiv \gamma_t$ & Turbulent growth rate & \quad $\displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{l,k}/dt}{2 E_k} dt $\\ \hline
+$\gamma_{nl}$ & Nonlinear growth rate & \ \ \ $\displaystyle\lim_{T \to \infty} \frac{1}{T} \int_0^T \frac{dE_{nl,k}/dt}{2 E_k} dt = - \gamma_{l}$\\ \hline
 $\gamma_{{\rm nm}}$ & Non-modal growth rate & \ $\frac{1}{4 \tau_{nl,k}} \ {\rm{Log}} \left[ {\rm{tr}} \{ e^{\mathbf{B}_k \tau_{nl,k}} e^{\mathbf{B}^{\dagger} \tau_{nl,k}} \} \right] $ \\ \hline
@@
 \end{tabular}
-\caption{Growth rate symbol descriptions and formulas}
+\caption{Growth rate symbols, descriptions, and formulas}
 \label{gamma_table}
@@
 
-In order to clarify that $\gamma_{l,k}$ is calculated from the quasi-steady-state portion of the nonlinear simulation, we define $\gamma_{t,k} \equiv \gamma_{l,k}$, and we call $\gamma_{t,k}$ the turbulent growth rate.
-$\gamma_{t,k}$ is a useful and informative quantity. 
-Note that in Eq.~\ref{Explicit_En_eqn} all of the linear terms on the right hand side of the equation are negative, except for the first term, which can have either sign. 
+In order to remind the reader that $\gamma_{l,k}$ is calculated from the quasi-steady-state turbulence portion of the nonlinear simulation, we define $\gamma_{t,k} \equiv \gamma_{l,k}$, 
+and we call $\gamma_{t,k}$ the turbulent growth rate. Moreover, $\gamma_{t,k}$ is a useful and informative quantity and one that we use heavily in this paper.
+To understand why $\gamma_{t,k}$ is informative, tote that in Eq.~\ref{Explicit_En_eqn} all of the linear terms on the right hand side of the equation are negative, except for the first term, which can have either sign.
 This term is proportional to the cross-field particle flux $\Gamma_x = \left< v_{E,x} n \right> \sim -i k_y \phi_k n_k$, meaning that a positive $\gamma_{t,k}$ indicates an outward particle flux
@@
 $\gamma_{s,k}$ and $\gamma_{t,k}$ for two different values of $\alpha$. In Fig.~\ref{alpha1e-5_gamma_s_t}, we show these growth rate spectra for $\alpha = 10^{-5}$, for which the system is
-highly non-normal, and in Fig.~\ref{alpha1_gamma_s_t} for $\alpha = 1$, for which the system is closer to being normal (see the inset of Fig.~\ref{gamma_max_vs_alpha}).
+highly non-normal, and in Fig.~\ref{alpha1_gamma_s_t} for $\alpha = 1$, for which the system is weakly non-normal (see the inset of Fig.~\ref{gamma_max_vs_alpha}).
 All calculations in this section use $\kappa=1$.
 In the highly non-normal case, the turbulent growth rate is very similar to the eigenmode growth rate for most values of $k_x,k_y$, but about 6 times higher where the spectrum peaks.
-For the more normal
+For the weakly non-normal
 case, in contrast, $\gamma_{t,k}$ is generally lower than $\gamma_{s,k}$, with the peak at nearly a factor of 2 less, with the peaks being at slightly different values of $k_y$.
@@
 As expected, when $\alpha$ is small ($\lesssim 10^{-3}$), $\gamma_{t} > \gamma_{s}$, with
-$\gamma_{t} \gg \gamma_{s}$ as $\alpha \to 0$. Additionally, when $\alpha$ becomes larger and the system becomes more normal ($\gtrsim 10^{-1}$), $\gamma_{t} < \gamma_{s}$.
-For both the normal and highly non-normal limits of the 2D HW model, $\gamma_{s}$ does a poor job of predicting $\gamma_{t}$ because $\gamma_{s}$ neglects the stable branch of the
+$\gamma_{t} \gg \gamma_{s}$ as $\alpha \to 0$. Additionally, when $\alpha$ becomes larger and the system becomes weakly non-normal ($\gtrsim 10^{-1}$), $\gamma_{t} < \gamma_{s}$.
+For both the weakly and highly non-normal limits of the 2D HW model, $\gamma_{s}$ does a poor job of predicting $\gamma_{t}$ because $\gamma_{s}$ neglects the stable branch of the
 dispersion relation -- the damped eigenmode~\cite{makwana2011}. 
@@
 The damped mode simply has both damping and transient growth effects depending on its relative magnitude at a particular time.
-These effects nearly cancel out on average when $\gamma_{t} \approx \gamma_{s}$.
+These effects nearly cancel out on average making $\gamma_{t} \approx \gamma_{s}$.
 
@@
 The importance of the stable eigenmode branch~\cite{baver2002} (or branches) has started to gain acceptance in the plasma community, although it has generally been used in normal
-or nearly normal systems to explain turbulent saturation~\cite{terry2006b,hatch2011,makwana2011} ($\gamma_{t} < \gamma_{s}$). 
+or weakly non-normal systems to explain turbulent saturation~\cite{terry2006b,hatch2011,makwana2011} ($\gamma_{t} < \gamma_{s}$). 
 The case of highly non-normal turbulence and the associated increased turbulent excitation hasn't been understood as well. 
@@
 
-As Biaskamp and Zeiler~\cite{biskamp1995} discovered, the 3D HW model nonlinear simulation has a fascinating property -- 
+As Biskamp and Zeiler~\cite{biskamp1995} discovered, the 3D HW model nonlinear simulation has a fascinating property -- 
 the turbulent energy injection is positive at $k_z = 0$ despite the fact that there can be no unstable linear eigenmodes at $k_z=0$.
@@
 Although the instability is a nonlinear instability, the first part of the mechanism -- the transport of background density by the convective cells -- is a linear one. This process occurs at $k_z=0$
-for which the adiabaticity parameter iz zero ($\alpha = k_z^2 \xi = 0$) meaning the linear matrix operator is highly non-normal and subject to transient growth.
+for which the adiabaticity parameter is zero ($\alpha = k_z^2 \xi = 0$) meaning the linear matrix operator is highly non-normal and subject to transient growth.
 So non-normal effects play a role in the 3D HW system for all $\xi$ due to the presence of $k_z=0$ structures.
@@
 \centerline{\includegraphics[width=0.4\textwidth]{gamma_max_nn}}
-\caption{The eigenmode $\gamma_{s}$ and turbulent $\gamma_{t}$ growth rates as a function of $\alpha$, which controls the normality of the linear operator.
+\caption{The eigenmode $\gamma_{s}$ and turbulent $\gamma_{t}$ growth rates as a function of $\alpha$, which controls the non-normality of the linear operator.
 All growth rates are peaks in $k_x-k_y$ space.
-The insert shows the scalar normality measure $1-H$ as a function of $\alpha$ for the matrices $\mathbf{B}_k$ that give the $\gamma_{s}$ curve.}
+The inset shows the scalar non-normality measure $1-H$ as a function of $\alpha$ for the matrices $\mathbf{B}_k$ that give the $\gamma_{s}$ curve.}
 \label{gamma_max_vs_alpha}
@@
 
-\section{Non-Modal Linear Procedure To Calculate Turbulent Growth Rate}
+\section{Analytic Non-Modal Approximation of the Turbulent Growth Rate}
 \label{sec_nm_procedure}
 
-In this section, we outline a procedure that uses quick and simple methods to approximate the turbulent growth rate spectrum $\gamma_{t,k}$. To do this, first recall that $\gamma_{t,k}$ is calculated
+In this section, we derive an analytic formula to approximate the turbulent growth rate spectrum $\gamma_{t,k}$. To do this, first recall that $\gamma_{t,k}$ is calculated
 from only linear terms in the original model equations (Eqs.~\ref{n_eq} and~\ref{phi_eq}).
@@
 This model equivalently describes a system in which 1) the turbulence begins as a random state, 
-2) evolves linearly for a time $\tau_{nl}$, and 3) randomizes again -- due to nonlinear energy transfer -- at which point the step 2 is repeated and so on.
+2) evolves linearly for a time $\tau_{nl}$, and 3) randomizes again -- due to nonlinear energy transfer -- at which point step 2 and 3 are repeated over and over.
 
@@
 Since $\gamma_{t,k}$ is a time-averaged quantity and the nonlinear model randomizes the system periodically, we can use an ensemble of initial conditions ($u(0)$'s that are random normalized vectors with uncorrelated components),
-and take two averages (over $t=0$ to $t = \tau_{nl}$ and over the ensemble) in place of the single time average that is generally used to calculate $\gamma_{t,k}$.
+and take two averages (one over $t=0$ to $t = \tau_{nl}$ and the other over the ensemble) in place of the single time average that is generally used to calculate $\gamma_{t,k}$.
 The two averages commute, but for simplicity,
@@
 \centerline{\includegraphics[width=0.4\textwidth]{gamma_max_with_nm_nn}}
-\caption{Comparison of the non-modal growth rates $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$ to the eigenmode $\gamma_{s}$ and turbulent $\gamma_{t}$ growth rates as a function of $\alpha$ for the 2D HW model.
+\caption{Comparison of the non-modal growth rates $\gamma_{\omega}$ and $\gamma_{\gamma}$ to the eigenmode $\gamma_{s}$ and turbulent $\gamma_{t}$ growth rates as a function of $\alpha$ for the 2D HW model.
 All growth rates are peaks in $k_x-k_y$ space.}
@@
 of the dominant eigenvalue of the linear matrix operator $\mathbf{B}_k$.
-The reason for using the least stable eigenmode growth rate rather than the most stable growth rate is that the least stable one should contribute more to the nonlinearity at long time after the randomization step.
+The reason for using the least stable eigenmode growth rate rather than the most stable growth rate is that the least stable one should contribute more to the nonlinearity long after the randomization step.
 As for the eigenmode frequency, recall that the imaginary part of the linear eigenvalues is the same for both eigenvalues of $\mathbf{B}_k$, so there is no question of which eigenfrequency to use. 
@@
 \caption{Comparison of the non-modal growth rates $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$ to the growth rates of the fastest growing eigenmode $\gamma_{s,k}$ 
-and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model with $\xi = 10^3$. The non-modal growth rates correctly predict the growth rate dominance of the $k_z=0$ mode
-despite the $k_z=0$ eigenmode stability.}
+and the of the turbulence $\gamma_{t,k}$ as a function of $k_z$ for the 3D HW model with $\xi = 10^3$.}
 \label{gamma_max_vs_kz_with_nm}
@@
 
-We test the model against the full direct numerical simulations (DNS) by comparing $\gamma_{\omega,k} \rm{and} \ \gamma_{\gamma,k}$ to $\gamma_{t,k}$. 
+We test the model against the full direct numerical simulations (DNS) by comparing $\gamma_{\omega,k}$ and $\gamma_{\gamma,k}$ to $\gamma_{t,k}$. 
 As seen in Fig.~\ref{gamma_max_with_nm}, the peak non-modal growth rate predictions for 2D HW turbulence generally agree within a factor of 2 with the turbulent growth rate. For $\alpha \le 10^{-4}$,
-in which the system is fairly non-normal, $\gamma_{\omega,k} \rm{and} \ \gamma_{\gamma,k}$ predict $\gamma_{t,k}$ quite accurately, especially compared to the eigenmode prediction. 
-This limit, of course, is important in subcritical systems, making the non-modal growth rates quite useful.
-As the system becomes more normal, $\gamma_{t,k}$ diverges from $\gamma_{\omega,k} \rm{and} \ \gamma_{\gamma,k}$, with $\gamma_{\gamma,k}$ doing a better job than $\gamma_{\omega,k}$.
+in which the system is fairly non-normal, $\gamma_{\omega,k} \ \rm{and} \ \gamma_{\gamma,k}$ predict $\gamma_{t,k}$ quite accurately, especially compared to the eigenmode prediction. 
+This limit is important in some systems, especially subcritical ones, making the non-modal growth rates quite useful.
+As the system becomes less non-normal, $\gamma_{t,k}$ diverges from $\gamma_{\omega,k} \ \rm{and} \ \gamma_{\gamma,k}$, with $\gamma_{\gamma,k}$ doing a better job than $\gamma_{\omega,k}$.
 
@@
 
-Despite these complications at $k_z=0$,  $\gamma_{\omega,k} \rm{and} \ \gamma_{\gamma,k}$ approximate $\gamma_{t,k}$ remarkably well, even when the linear matrix is only moderately ($1-H \approx 10^{-1}$) non-normal
+Despite these complications at $k_z=0$,  $\gamma_{\omega,k} \ \rm{and} \ \gamma_{\gamma,k}$ approximate $\gamma_{t,k}$ remarkably well, even when the linear matrix is only weakly ($1-H \approx 10^{-1}$) non-normal
 (see Figs.~\ref{gamma_max_vs_kz_with_nm} and~\ref{gamma_max_vs_xi}).
@@
 When the nonlinear terms in the model equations are energetically conservative, as they are in the HW model, 
-the linear terms linear terms control the energy injection into the turbulent fluctuations.
+the linear terms control the energy injection into the turbulent fluctuations.
 This property makes the turbulent growth rate, which is defined in terms of only the linear terms in the energetic model equations, a simple yet vital piece of information
-in understanding the nature of the system. When the system is non-normal, estimating the growth rate without nonlinear simulations almost surely requires non-modal techniques.
+in understanding the nature of the system. When the system is significantly non-normal, estimating the growth rate without nonlinear simulations almost surely requires non-modal techniques.
 
@@
 case, there is some transitional value of $R$, labeled $R_g$, below
-which turbulence cannot occur. When a perturbation with amplitude
-($A$) above the line occurs in the system due to external input or noise, 
-turbulence will be excited. Perturbations with amplitude below the line will decay away. The higher the value of $R$, the smaller the perturbation needs to be to excite turbulence.
+which turbulence cannot occur. When a particular perturbation with amplitude ($A$) above the line occurs in the system due to external input or noise, turbulence will be excited. 
+Perturbations with amplitude below the line will decay away. The higher the value of $R$, the smaller the perturbation needs to be to excite turbulence.
 Experimental evidence and theoretical arguments show this curve should have a $(R-R_g)^{-\lambda}$ dependence (for asymptotically large $R$), with $\lambda \sim 1$~\cite{waleffe1995b,grossmann2000,hof2003}.
-A paradigmatic example of this kind of system is Hagen-Poisuille pipe flow -- flow in a cylindrical pipe with a parabolic flow profile -- for which $R_g \approx 2000$.
+A paradigmatic example of this kind of system is Hagen-Poiseuille pipe flow -- flow in a cylindrical pipe with a parabolic flow profile -- for which $R_g \approx 2000$.
 
 The lower curve represents the second kind of subcritical system. Above a critical value of the control parameter $R_c$, the system is linearly unstable. But, for $R_g < R < R_c$, the system is
-unstable to finite amplitude perturbations, or ``nonlinearly'' unstable. A paradigmatic example of this is Poissuille channel flow -- the rectangular analog of pipe flow -- for which
+unstable to finite amplitude perturbations, or ``nonlinearly'' unstable. A paradigmatic example of this is Poiseuille channel flow -- the rectangular analog of pipe flow -- for which
 $R_c = 5772$ and $R_g \approx 1000$~\cite{grossmann2000}.
@@
 Clearly, our non-modal procedure with our particular choice of linear time scales does not provide even an order-of-magnitude accurate prediction of $\kappa_g$ for all values of $\xi$.
-Nevertheless, the non-modal predictions display some qualitative agreement with the DNS results in that the non-modal predictions trend fairly accurately with $\xi$. 
+The cause of this failure is the breakdown of our model of the nonlinearity. After all, modeling the nonlinearity is the only way in which our method diverges from a DNS.
+We speculate that the breakdown is due to our model's reliance on a picture of fully-developed turbulence in which turbulent structures decorrelate.
+As we approach the subcritical point, the turbulence becomes somewhat coherent.
+In this case, our critical balance argument seems to severely under-predict the time scale in which the turbulence decoheres.
+
+Nevertheless, our non-modal predictions for subcritical onset display some qualitative agreement with the DNS results in that the non-modal predictions trend fairly accurately with $\xi$. 
 And they do provide a prediction to the transition density gradient $\kappa_g$ in contrast to the eigenmode growth rates, which are always negative in this system and thus do not predict turbulence for any $\kappa$.
 So, there is some value in the non-modal predictions, especially in light of the fact that it took over $10^5$ computer hours to calculate the $\kappa_g$ line using DNS and only seconds using the non-modal procedure.
-Furthermore, it is likely that more sophisticated models for the nonlinearities will improve the predictions in the future.
+Furthermore, it is likely that more sophisticated models for the nonlinearities can improve the predictions.
 

### 2014-12-26 Referee_Responses/referee_response.tex
Dec 26

@@
 we believe that it is still valuable and provides some important insight. What we have done, then, is move it to an appendix. 
-We have also title, abstract, and introduction to de-emphasize that part of the paper. \\ \\
+We have also adjusted the title, abstract, and introduction to de-emphasize that part of the paper. \\ \\
 
@@
 So each member of the ensemble is initially one possible value of this complex ratio. But once the initial states have evolved under the linear operator for some time, not all of
-the possible values of the ratio are populated. In fact, after a long enough time, this ratio is nearly equal to that for the faster growing eigenmode for all members of the ensemble.
+the possible values of the ratio are populated. In fact, after a long enough time, this ratio is nearly equal to that of the faster growing eigenmode for all members of the ensemble.
 The time average, then, produces something meaningful -- that is, an average over many ratios between a random ratio and the eigenmode ratio.
